---
title: "Analyzing Probability Samples"
author: "Michael Dumelle, Tony Olsen, Tom Kincaid, and Marc Weber"
output: 
  html_document:
    theme: flatly
    number_sections: true
    highlighted: default 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Analyzing Probability Samples}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
```

After selecting a probability sample, population parameters (e.g. means and totals) can be estimated using spsurvey's analysis functions. We focus mainly on using the `NLA_PNW` data to introduce some of these analysis options. The `NLA_PNW` data contains response variables measured at several lakes in the Pacific Northwest Region of the United States. There are six variables in `NLA_PNW` used throughout this vignette:

* `SITE_ID`: a unique site identifier
* `WEIGHT`: the sampling design weights
* `NITR_COND`: nitrogen condition categories (`Good`, `Fair`, and `Poor`)
* `URBAN`: urban land categories (`Urban` and `Non-Urban`)
* `STATE`: state (with levels `California`, `Oregon`, and `Washington`)
* `BMMI`: benthic macroinvertebrate multi-metric index

Before proceeding, we load spsurvey:
```{r}
library(spsurvey)
```

# Categorical Variable Analysis

Categorical variables are analyzed in spsurvey using the `cat_analysis()` function. The `cat_analysis()` function estimates the proportion of observations and the total extent (number of units, length or area) that belongs to each level of the categorical variable. Several useful pieces of output are returned by `cat_analysis()`, including estimates, standard errors, margins of error, and confidence intervals. The `.P` suffix correspond to estimates of proportions for each category, while the `.U` estimates correspond to estimates of number of units, length or area for each category. 

## Unstratified Analysis

To estimate the proportion of total lakes in each nitrogen condition category and the total number of lakes in each nitrogen condition category, run
```{r}
ests <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT"
)
ests
```
The estimate of the proportion of lakes in `Good` condition is 51.35% with a 95% confidence interval of (36.8%, 65.9%), while the estimate of the total number of lakes in `Good` condition is 5484 lakes with a 95% confidence interval of (3089, 7882). In each case, the estimated standard error and margin of error is given.

Sometimes the goal is to estimate proportions and totals separately for subsets of the population -- these subsets are known as subpopulations. To estimate the proportion of total lakes in each nitrogen condition category and the total number of lakes in each nitrogen condition category separately for `Urban` and `Non-Urban` lakes, run
```{r}
subpop_ests <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT",
  subpop = "URBAN"
)
subpop_ests
```
If more than one type of subpopulation is of interest,  then `subpop` can be a vector of the subpopulation variables.

## Stratified Analysis

To estimate the proportion of total lakes in each nitrogen condition category and the total number of lakes in each nitrogen condition category for a design stratified by state, run
```{r}
strat_ests <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT",
  stratumID = "STATE"
)
strat_ests
```
Notice that the stratified design's standard errors are smaller than the unstratified design's standard errors.

To then compute these estimates separately for `Urban` and `Non-Urban` lakes, run
```{r}
strat_subpop_ests <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT",
  stratumID = "STATE",
  subpop = "URBAN"
)
strat_subpop_ests
```

# Continuous Variable Analysis

Continuous variables are analyzed in spsurvey using the `cont_analysis()` function. The `cont_analysis()` function estimates cumulative distribution functions, quantiles, and means of continuous variables. By default, all these quantities are estimated, though the type of estimates returned is controlled using the `statistics` argument. For the quantities requiring estimation, several useful pieces of output are returned by `cont_analysis()`, including estimates, standard errors, margins of error, and confidence intervals. 

## Unstratified Analysis

To estimate quantiles and means of `BMMI`, run
```{r}
ests <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT",
  statistics = c("pct", "mean")
)
ests
```

To estimate quantiles and means of `BMMI` separately `Urban` and `Non-Urban` lakes
```{r}
subpop_ests <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT",
  subpop = "URBAN",
  statistics = c("pct", "mean")
)
subpop_ests
```


## Stratified Analysis

To estimate quantiles and means of `BMMI` for a design stratified by state, run
```{r}
strat_ests <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT",
  stratumID = "STATE",
  statistics = c("pct", "mean")
)
strat_ests
```

To then compute these estimates separately for `Urban` and `Non-Urban` lakes, run
```{r}
strat_subpop_ests <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT",
  stratumID = "STATE",
  subpop = "URBAN",
  statistics = c("pct", "mean")
)
strat_subpop_ests
```

# Additonal Analysis Approaches

Alongside the `cat_analysis()` and `cont_analysis()` functions, spsurvey offers analysis functions for attributable risk, relative risk, change, and trend. Attributable and relative risk analyses capture attributable and relative risks, respectively, of environmental resources being in poor condition after exposure to a stressor. Attributable risk analysis is performed using the `attrisk_analysis()` function, while relative risk analysis is performed using the `relrisk_analysis()` function. Change and trend analysis capture the behavior of environmental resources between two probability samples, while trend analysis generalizes this approach to include multiple probability samples. Often, change and trend analyses are performed to study an environmental resources through time.  Change analysis is performed using the `change_analysis()` function, and trend analysis is performed using the `trend_analysis()` function. The `attrisk_analysis()`, `relrisk_analysis()`, `change_analysis()` and `trend_analysis()` functions all share very similar syntax to the `cat_analysis()` and `cont_analysis()` functions.

## Attributable and Relative Risk

The attributable risk is defined as $1 -  \frac{P(Response = Poor | Stressor = Good)}{P(Response = Poor)}$, where $P(\cdot)$ is a probability. The attributable risk measures the proportion of the response variable in poor condition that could be eliminated if the stressor was always in good condition. To estimate the attributable risk of benthic macroinvertebrates with a phosphorous condition stressor, run
```{r}
attrisk <- attrisk_analysis(
  NLA_PNW, 
  siteID = "SITE_ID",
  vars_response = "BMMI_COND",
  vars_stressor = "PHOS_COND",
  weight = "WEIGHT"
)
attrisk
```

The relative risk is defined as $\frac{P(Response = Poor | Stressor = Poor)}{P(Response = Poor | Stressor = Good)}$, which measures the risk of the response variable being in poor condition relative to the stressor. To estimate the relative risk of benthic macroinvertebrates being in poor condition relative to a phosphorous condition category stressor, run
```{r}
relrisk <- attrisk_analysis(
  NLA_PNW, 
  siteID = "SITE_ID",
  vars_response = "BMMI_COND",
  vars_stressor = "PHOS_COND",
  weight = "WEIGHT"
)
relrisk
```

The default levels of `vars_response` and `vars_stressor` are `"Poor"` (event occurs) and `"Good"` (event does not occur), though this can be changed using the `response_levels` and `stressor_levels` arguments, respectively. Similar to `cat_analysis()` and `cont_analysis()` from the previous sections, subpopulations and stratification are incorporated using `subpops` and `stratumID`, respectively. For more on attributable and relative risk in an environmental resource context, see Van Sickle and Paulsen (2008).

## Change and Trend Analysis

To demonstrate change analysis, we use the `NRSA_EPA7` data. The `NRSA_EPA7` data contains several variables, five of which we use next:

* `SITE_ID`: a unique site identifier
* `WEIGHT`: the sampling design weights
* `NITR_COND`: nitrogen condition category (`Good`, `Fair`, and `Poor`)
* `BMMI`: benthic macroinvertebrate multi-metric index
* `YEAR`: probability sample (survey) year

To estimate the change between time points for `BMMI` (a continuous variable) and `NITR_COND`, run
```{r}
change <- change_analysis(
  NRSA_EPA7,
  siteID = "SITE_ID",
  vars_cont = "BMMI",
  vars_cat = "NITR_COND",
  surveyID = "YEAR",
  weight = "WEIGHT"
)
change
```
The `surveyID` argument is the variable in the data that indicates the probability sample (`YEAR` in the previous example).

Trend analysis generalizes change analysis to multiple probability samples. Though we omit an example here, the arguments to `trend_analysis()` are very similar to the arguments for `change_analysis()`. One difference is that `trend_analysis()` contains arguments that specify which statistical model to apply to the estimates from each probability sample.

# Variance Estimation

The default variance parameter estimator in spsurvey is the local neighborhood variance estimator (Stevens and Olsen, 2003). The local neighborhood variance estimator incorporates spatial locations by weighting nearby observations more heavily when estimating the variance. Because it incorporates spatial locations, the local neighborhood variance estimate tends to be smaller than variance estimates from approaches ignoring spatial locations. Several additional variance estimation options are available in spsurvey's analysis functions through the `vartype` and `jointprob` arguments.

# References

Sickle, J. V., & Paulsen, S. G. (2008). Assessing the attributable risks, relative risks, and regional extents of aquatic stressors. *Journal of the North American Benthological Society*, 27(4), 920-931.

Stevens Jr, D. L., & Olsen, A. R. (2003). Variance estimation for spatially balanced samples of environmental resources. *Environmetrics*, 14(6), 593-610.




