---
title: "Selecting Spatially Balanced Samples from a Finite Population using spsurvey and the GRTS algorithm"
author: "Michael Dumelle, Tony Olsen, Tom Kincaid, and Marc Weber"
output: 
  html_document:
    theme: flatly
    number_sections: true
    highlighted: default 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
library(devtools)
load_all()
data("NE_lakes")
```

# Introduction

In this vignette, you will use `spsurvey` to select probability samples using the the GRTS algorithm (Stevens and Olsen, 2004) and independent random sampling. A greater focus will be placed on GRTS, as this algorithm yields samples having "spatial balance." More specifically, throughout this vignette, you will explore:

* [The `NE_Lakes` Data] 
* [GRTS Designs]
    * [An Unstratified, Equal Probability GRTS Design] 
    * [An Unstratified, Unequal Probability GRTS Design]
    * [A Stratified, Equal Probability GRTS Design]
    * [A Stratified, Proportional Probability GRTS Design]
* [Independent Random Sampling Designs]
* [Additional Design Features]
    * [Legacy Sites]
    * [Minimum Distance Between Sites]
    * [Replacement Sites]
* [Calculating the Spatial Balance of a Sample]

If you have not yet done so, please read the "Start Here" vignette, accessed [here](dummy link for now) or by running `vignette("start_here", package = spsurvey)`. In this vignette, you will review Coordinate Reference Systems (CRS), `sf` objects, installing and loading the `spsurvey` package, setting a reproducible seed, and using the "pipe" operator (`%>%`).

# The `NE_Lakes` Data

Before anything else, you must load `spsurvey` by running
```{r setup, eval = FALSE}
library(spsurvey)
```
If you want to reproduce the results from code in this vignette, you must also set a seed by running
```{r fake seed, eval = FALSE}
set.seed(5)
```

```{r real seed, echo = FALSE}
# actually setting seed on our end
set.seed(5)
```

Throughout this vignette, you will use select probability samples from the `NE_Lakes` dataset. The `NE_lakes` dataset is made up of 5328 lakes in the Southern New England Region of the United States of America. To load the data into R, run
```{r load data, eval = FALSE}
data("NE_lakes")
```
To view the first few rows, run
```{r view data}
head(NE_lakes)
```
You will immediately notice some geometry information has been returned:

* `geometry type`: This is a `POINT` geometry because each lake has a set of `x` and `y` coordinates.
* `dimension`: There are two dimensions, `X` (east-west) and `Y` (north-south).
* `bbox`: The bounding box of the spatial locations.
* `projected CRS`: The coordinates in a USA Contiguous Albers Equal Area Conic Projection.

The next few lines contain regarding variables in `NE_lakes`:

* `LAT_DD_N83`: The latitute coordinate for each lake centroid using the NAD83 CRS (EPSG: 5070)
* `LON_DD_N83`: The longitude coordinate for each lake centroid using the NAD83 CRS (EPSG: 5070)
* `AREA_HA`: The area (in hectares) of each lake.
* `ELEVATION`: The elevation (in meters) of each lake.
* `geometry`: The projected spatial coordinates for each lake centroid.

The `ELEVATION` column has continuous elevation measurements, but next will turn this variable into a categorical one for use throughout the rest of the vignette. There will be three groups:

* "low" -- lake elevation from 0 to 50 meters
* "medium" -- lake elevation from 50 to 200 meters
* "high" -- lake elevation greater than 200 meters

To convert `ELEVATION` into this categorical variable, run
```{r elev cat}
max_elevation <- max(NE_lakes$ELEVATION) # store the max elevation 
NE_lakes$ELEVATION <- cut(NE_lakes$ELEVATION,
                          breaks = c(0, 50, 200, max_elevation),
                          labels = c("low", "medium", "high"),
                          include.lowest = TRUE
                      )
```

To view the number of lakes in each category, run
```{r}
table(NE_lakes$ELEVATION)
```

Next, we save the bounding box of coordinates from the data for maintaining consistent aspect ratios across plots generated throught this vignette:

```{r}
boundbox <- st_bbox(NE_lakes)
xmin <- boundbox["xmin"]
xmax <- boundbox["xmax"]
ymin <- boundbox["ymin"]
ymax <- boundbox["ymax"]
```

# GRTS Designs

In this section, you will select spatially balanced samples -- but what exactly is a "spatially balanced sample?" And why are they important? Stevens and Olsen (2004) defined a spatially balanced sample as a sample mirroring the density of the population -- this is often refferred to as a sample that is a *miniature* of the population. This is different from spatial balance with respect to geography. To elucidate this distinction, consider the following plot of the `NE_lakes` locations:
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Locations of lakes in the `NE_lakes` dataset."}
# gray background
par(bg = "gray95")
# plot the population sites
plot(st_geometry(NE_lakes), xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
```
Notice that lakes are have the highest density in the northeastern locations and the sparsest density in the northwestern locations. A spatially balanced sample with respect to the population mirrors the density in the population, sampling more in the northeast and less in the northwest, while a spatially balanced sample with respect to geography omits this density information but takes into account the absolute distances among neighboring sites. These two types of spatial balance are useful in different contexts, and no type is superior over the other. But it is important to be aware of this distinction and its implications. Henceforth, when we refer to "spatial balance," we mean spatial balance with respect to the population.

Spatially balanced samples are so important because they generally yield estimators of means and totals *having narrower confidence intervals* than samples lacking spatial balance. This means our estimators are more *precise*, which can lead to reduces resource costs and a *more accurate representation* of the true, uknonwn values of the quantities we are studying.

## An Unstratified, Equal Probability GRTS Design

To select an equal proability GRTS sample, run:
```{r}
# equal probability design
eqprob <- grts(sframe = NE_lakes, seltype = "equal", n.samp = 60)
```

`eqprob` contains an `sf` object with the selected sites. This `sf` object is stored as `eqprob$sites.base`. To view the first few rows, run:
```{r}
head(eqprob$sites.base)
```

You will notice several variables are contained in `eqprob$sites.base`. The rownames of `eqprob$sites.base` match the row locations in `NE_lakes`. `stratum` indicates the strata of each row; for this non-stratified design, all entries in `stratum` are `"None"`. `wgt` and `ip` give the survey design weights and inclusion probabilities of each site, respectively. `caty` indicates the unequal probability cateogry; for this equal probability design, all entries in `caty` are `"None"`.

To overlay the spatially balanced sample on the original population, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Population sites (black) and sampled sites (red) obtained for an equal probability GRTS design."}
# gray background
par(bg = "gray95")
# plot the population sites
plot(st_geometry(NE_lakes), main = "", reset = FALSE, 
     xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
# add the sampled sites
plot(st_geometry(eqprob$sites.base), pch = 19, add = TRUE, cex = 1.5, col = "red")
# add a legend
legend(x = "bottomright", legend = c("Population Sites", "Sampled Sites"),
       pch = c(1, 19), col = c("black", "red"), cex = 1.25)
```

`grts` returns several design elements used by the function as a tool to both keep track of and to verify the specifications of your design. You can access this information by running `eqprob$dsgn`. A list is printed containing several pieces of information -- several of which may be `NULL`. These `NULL` values are expected and change based on the specifications of your design; this will be made more clear later on. For now, you will focus on the relevant non-`NULL` information.
```{r, echo = FALSE}
eqprob$dsgn$legacy_option <- NULL
for (s in names(eqprob$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob$dsgn[[s]])) == 0) eqprob$dsgn[[s]] <- NULL
}
eqprob$dsgn
```

The `stratum` variable (the `stratum_var` argument to `grts()`) in `eqprob$sites.base` is given the name `stratum`.  All entries in `stratum` are given the same value: `"None"`.  The selection type for the `"None"` stratum is equal probability -- as indicated by `"equal"`. There are 60 (`n.samp`) samples in the `"None"` stratum


## An Unstratified, Unequal Probability GRTS Design

Suppose we wish to sample the lakes in the highest elevation group with a higher probability than all other lakes. To select an unequal probability GRTS sample by elevation group, run:
```{r}
# expected sample size for each group with names
caty.n <- c(low = 15, medium = 15, high = 30)
# printing caty.n
caty.n
# selecting the unequal probability sample
uneqprob <- grts(sframe = NE_lakes, seltype = "unequal", 
                          n.samp = 60, caty_var = "ELEVATION", 
                          caty.n = caty.n)
```

The `seltype` argument is "unequal" for an unequal probaility design, the `caty_var` argument is the variable in `NE_lakes` for which uneuqal probaiblity sampling is desired, and the `caty.n` argument is a vector with names equal to the categories in `caty_var` and values indicating the expected samples sizes desired in each category. The sum of the values in `caty.n` must add up to the total desired sample size indicated by the `n.samp` argument. Though the expected sample sizes are given by `caty.n`, there is no guarantee your sample will contain within-group sample sizes equaling these values. 

The non-NULL `dsgn` output is
```{r, echo = FALSE}
uneqprob$dsgn$legacy_option <- NULL
for (s in names(uneqprob$dsgn)) {
    if (length(Filter(Negate(is.null), uneqprob$dsgn[[s]])) == 0) uneqprob$dsgn[[s]] <- NULL
}
uneqprob$dsgn
```


## A Stratified, Equal Probability GRTS Design

Unlike unequal probability designs, stratified designs ensure each within-group sample size equals the expected sample size. To select a spatially balanced, stratified design having equal probabilities of selection within each strata (state), run:
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Population sites (black) and sampled sites (red) obtained for the medium `ELEVATION` group using stratified GRTS design having equal selection probabilities within strata."}
# names of the strata
strata <- c("low", "medium", "high")
# sample sizes of the strata
strata.n <- c(low = 20, medium = 20, high = 20)
# selecting the stratified, equal probability sample
strat_eqprob <- grts(sframe = NE_lakes, stratum_var = "ELEVATION",
                        stratum = strata, seltype = "equal", 
                        n.samp = strata.n)
```

The arguments are similar to the `caty` arguments from [An Unequal Probability GRTS Design]; `stratum` is used instead of `caty`. The argument to `stratum_var` is the name of the `stratum` variable in `NE_lakes` (`ELEVATION`), the argument to `stratum` is a vector containing the name of each strata, and the argument to `n.samp` a named vector indicating sample sizes for each strata. The argument to `seltype` is `"equal"`, which indicates equal selection probabilities within strata.

The non-NULL `dsgn` output is:
```{r, echo = FALSE}
eqprob$dsgn$legacy_option <- NULL
for (s in names(strat_eqprob$dsgn)) {
    if (length(Filter(Negate(is.null), strat_eqprob$dsgn[[s]])) == 0) strat_eqprob$dsgn[[s]] <- NULL
}
strat_eqprob$dsgn
```

To visuzlize the sample for the medium `ELEVATION` category, run
```{r, fig.show = "hold", fig.align = "center"}
# gray background
par(bg = "gray95")
# subset the population to include ma lakes
low_pop <- subset(NE_lakes, ELEVATION == "low")
# subset the sample to include ma lakes
low_samp <- subset(strat_eqprob$sites.base, ELEVATION == "low")
# plot the population man-made lakes
plot(st_geometry(low_pop), main = "", reset = FALSE, 
     xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
# add the sampe man-made lakes
plot(st_geometry(low_samp), pch = 19, add = TRUE, cex = 1.5, col = "red")
# add a legend
legend(x = "bottomright", legend = c("Population Sites", "Sampled Sites"),
       pch = c(1, 19), col = c("black", "red"), cex = 1.25)
```

## A Stratified, Proportional Probability GRTS Design

The previous stratified design can be further generalized by proportional selection within strata. To select a stratified design where in each strata, lakes are sample proportionally to their area (in hectares; `AREA_HA`), run
```{r}
# names of the strata
strata <- c("low", "medium", "high")
# sample sizes of the strata
strata.n <- c(low = 20, medium = 20, high = 20)
# selecting the stratified, unequal probability sample
strat_uneqprob <- grts(sframe = NE_lakes, stratum_var = "ELEVATION",
                        stratum = strata, seltype = "proportional", 
                        n.samp = strata.n, aux_var = "AREA_HA")
```

The `seltype` argument indicates proportional sampling within strata, and the `aux_var` argument indicates the variable in `NE_lake` to proportionally sample by. Proportional sampling is appropriate when the proportional variable is positive -- for example, all lake areas are greater than zero.

The non-NULL `dsgn` output is:
```{r, echo = FALSE}
strat_uneqprob$dsgn$legacy_option <- NULL
for (s in names(strat_uneqprob$dsgn)) {
    if (length(Filter(Negate(is.null), strat_uneqprob$dsgn[[s]])) == 0) strat_uneqprob$dsgn[[s]] <- NULL
}
strat_uneqprob$dsgn
```

# Independent Random Sampling Designs

If you want to conduct an independent random sample (IRS) instead of a GRTS sample, simply replace the `grts()` function with the `irs()` function. All of the arguments are identical, so any of the aforementioned examples can be easily applied to the IRS scenario as well. For example, to select an equal probability IRS, run
```{r}
# equal probability design
eqprob_irs <- irs(sframe = NE_lakes, seltype = "equal", n.samp = 60)
```

To overlay the IRS on the original population, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Population sites (black) and sampled sites (red) obtained for an equal probability IRS design."}
# gray background
par(bg = "gray95")
# plot the population sites
plot(st_geometry(NE_lakes), main = "", reset = FALSE, 
     xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
# add the sampled sites
plot(st_geometry(eqprob_irs$sites.base), pch = 19, add = TRUE, cex = 1.5, col = "red")
# add a legend
legend(x = "bottomright", legend = c("Population Sites", "Sampled Sites"),
       pch = c(1, 19), col = c("black", "red"), cex = 1.25)
```

The non-NULL design output is
```{r, echo = FALSE}
eqprob_irs$dsgn$legacy_option <- NULL
for (s in names(eqprob_irs$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_irs$dsgn[[s]])) == 0) eqprob_irs$dsgn[[s]] <- NULL
}
eqprob_irs$dsgn
```

# Additional Design Features

Besides selection probability type and stratification, there are several other special cases that GRTS and IRS designs now accomodate: [Legacy Sites], [Minimum Distance Between Sites], and [Replacement Sites]. The following examples are shown using the `grts()` function, but the `irs()` function can replace `grts()` if so desired; the function arguments are identical.

## Legacy Sites

It is often the case that a user wants to wants to make sure a specific set of sites are guaranteed to be included in the sample; these are known as legacy sites.  It is often the case that legacy sites were selected as part of a previous probability sample; this is the scenario we turn our attention to.  Suppose a new variable, `legacy`, is added to `NE_lakes` and contains five randomly selected legacy sites.

```{r}
# creating a new legacy variable where each value is FALSE
NE_lakes$legacy <- NA
# randomly choosing 5 observations to treat as legacy sites
NE_lakes$legacy[sample(1:nrow(NE_lakes), size = 5)] <- TRUE
```

To accomodate legacy sites in an unstratified, equal probability sample, run:
```{r}
eqprob_legacy <- grts(NE_lakes, seltype = "equal", n.samp = 60,
                         legacy_option = TRUE, legacy_var = "legacy")
```

To plot the population sites next to the sampled sites, run:
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Population sites (black) and sampled non-legacy sites (aquamarine), and sampled legacy sites (orange) obtained for an equal probability GRTS design."}
# subsetting the legacy sites
sampled_legacy <- subset(eqprob_legacy$sites.base, legacy == TRUE)
# subsetting the sampled nonlegacy sites
sampled_nonlegacy <- subset(eqprob_legacy$sites.base, is.na(legacy))
# gray background
par(bg = "gray95")
# plotting the population sites
plot(st_geometry(NE_lakes), main = "", reset = FALSE, 
     xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
# adding the nonlegacy sampled sites
plot(st_geometry(sampled_nonlegacy), pch = 19, add = TRUE, cex = 1.5, col = "aquamarine")
# adding the legacy sites
plot(st_geometry(sampled_legacy), pch = 19, add = TRUE, cex = 1.5, col = "orange")
# adding a legend
legend(x = "bottomright", legend = c("Population Sites", "Sampled Sites (Non-Legacy)", "Sampled Sites (Legacy)"), pch = c(1, 19, 19), col = c("black", "aquamarine", "orange"), cex = 1)
```

With some data, you may encounter a warning message saying `Of the x grid cells from which sample points were selected, y (y/x%) of the cells contained more than one sample point. This warning message exists because sometimes the requirement that legacy sites must be included prevents the sample from being as spatially balanced as a sample from a standard GRTS algorithm. 

The non-null design output is
```{r}
for (s in names(eqprob_legacy$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_legacy$dsgn[[s]])) == 0) eqprob_legacy$dsgn[[s]] <- NULL
}
eqprob_legacy$dsgn
```


## Minimum Distance Between Sites

The standard GRTS algorithm may select sites that are closer together than desired. You can adjust the algorithm to require some minimum distance between sites.  Supppose we requires sites be 5,000. meters ($\approx$ 5 miles) apart. To include this requirement in an unstratified, equal probability design, run:
```{r}
# setting the minimum distance
mindis <- 5000
eqprob_mindis <- grts(NE_lakes, seltype = "equal", n.samp = 60,
                         mindis = mindis)
```

To plot the population sites next to the sampled sites, run:
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Population sites (black) and sampled sites (red) for an equal probability GRTS design having a minimum distance between sites of 500 meters."}
# gray background
par(bg = "gray95")
# plotting the population sites
plot(st_geometry(NE_lakes), main = "", reset = FALSE, 
     xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
# adding the sampled sites
plot(st_geometry(eqprob_mindis$sites.base), pch = 19, add = TRUE, cex = 1.5, col = "red")
# adding a legend
legend(x = "bottomright", legend = c("Population Sites", "Sampled Sites"),
       pch = c(1, 19), col = c("black", "red"), cex = 1.25)
```
It is possible that the GRTS algorithm will not be able to ensure the minimum distance is met for all sampled sites.  In this context, the algorithm will attempt to ensure the minimum distance is met for as many sites as possible and return a warning stating that the requirement was not met after several iterations of the algorithm.

## Replacement Sites

It is often the case that a selected site is unable to be sampled. This can happen for a variety of reasons: landowner deinial, lack of funding, or lack of resources, to name a few. If you expect there are sites you may not be able to sample, you can include potential replacement sites in your GRTS design. `spsurvey` gives you two methods to select these replacement sites: reverse heirarchical ordering and nearest neighbor.

First, we discuss reverse heirarchical ordering replacement. In this context, the number desired sample size, $n$, and the number of replacement sites, $n_r$, must be specified before the design is selected. A standard GRTS algorithm is used to select a sample of size $n + n_r$. Then, these sites are placed into "reverse heirarchical order" and the $n$ base sites and $n_r$ replacement sites are classified. The $n$ base sites should be sampled first. If any of these $n$ sites are unable to be sampled, replacement sites should be used. For more details, see Stevens and Olsen (2004). To select a GRTS sample of size 60 with 10 reverse heirarchical ordering replacement sites, run
```{r}
eqprob_rho <- grts(NE_lakes, seltype = "equal", n.samp = 60,
                         n.over = 10)
```

`eqprob_rho$sites.base` contains the $n$ base sites, and `eqprob_reverse$sites.over` contains the $n_r$ replacement sites. The non-null design output is
```{r}
eqprob_irs$dsgn$legacy_option <- NULL
for (s in names(eqprob_rho$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_rho$dsgn[[s]])) == 0) eqprob_rho$dsgn[[s]] <- NULL
}
eqprob_rho$dsgn
```

To plot the population sites next to the sampled sites and the replacement sites, run:
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Population sites (black), sampled sites (aquamarine), and replacement sites (orange) for an equal probability GRTS design using reverse heirarchical ordering for the replacement sites."}
# gray background
par(bg = "gray95")
# plotting the population sites
plot(st_geometry(NE_lakes), main = "", reset = FALSE, 
     xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
# adding the base sample sites
plot(st_geometry(eqprob_rho$sites.base), pch = 19, add = TRUE, cex = 1.5, col = "aquamarine")
# adding the replacement sites
plot(st_geometry(eqprob_rho$sites.over), pch = 19, add = TRUE, cex = 1.5, col = "orange")
# adding a legend
legend(x = "bottomright", legend = c("Population Sites", "Sampled Sites", "Replacement Sites"), pch = c(1, 19, 19), col = c("black", "aquamarine", "orange"))
```

The second method for selecting replacement sites is the nearest neighbor method. A base GRTS sample of size $n$ is first selected. If a site is then unable to be sampled, the nearest neighbor method provides $n_r$ (up to 10) replacement sites that are closest to the original, unsampled site. This implies that for each site in the GRTS sample, $n_r$ replacement sites are available. To select a GRTS sample of size 60 with a 10 nearest neighbors for each site, run
```{r}
eqprob_nn <- grts(NE_lakes, seltype = "equal", n.samp = 60,
                         n.near = 10)
```

This nearest neighbor method is more computationally intensive than the reverse heirarchical ordering method. The above code can take a almost a minute to run on your machine.  view the population sites, base sample sites, and each sampled site's nearest replacementsite by running
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Population sites (black), sampled sites (aquamarine), and replacement sites (orange) for an equal probability GRTS design using nearest neighbors for the replacement sites."}
# gray background
par(bg = "gray95")
# plotting the population sites
plot(st_geometry(NE_lakes), main = "", reset = FALSE, 
     xlim = c(xmin, xmax), ylim = c(ymin, ymax), cex = 0.5)
# adding the sampled sites
plot(st_geometry(eqprob_nn$sites.base), pch = 19, add = TRUE, cex = 2.25, col = "aquamarine")
# adding the first replacement site
plot(st_geometry(subset(eqprob_nn$sites.near, siteuse == "Near__1st")), pch = 19, add = TRUE, cex = 1.25, col = "orange")
# adding a legend
legend(x = "bottomright", legend = c("Population Sites", "Sampled Sites", "Replacement Sites"), pch = c(1, 19, 19), col = c("black", "aquamarine", "orange"))
```
You will notice that these replacement sites are very close to the originally sampled sites.


# Calculating the Spatial Balance of a Sample

To measure the spatial balance of a design, Stevens and Olsen (2004) proposed using a set of Voroni polygons. A single voroni polygon for a sampled site contains all population sites closer to this sampled site than any other sampled site. The set of Voroni polygons for all sampled sites will contain all population sites. If a sample is spatially balanced, we would expect that the proportion of total population sites in each Voroni polygon to be approximately $1/n$.

Let $p_i$ be the proportion of population sites in Voroni polygon $i$. The Pielou's evenness index can be used as a measure of spatial balance and can be defined as $1 + \frac{1}{n}\sum_{i = 1}^n p_i log(p_i)$. One attractive feature of Pielou's evenness index is that it is unitless; a value of 0 indicates perfect spatial balance, and a value of 1 indicates no spatial balance. To compute the Pielou's evenness index of the equal probabily GRTS and equal probability IRS designs, run
```{r}
# compute for grts
pielou_grts <- spbalance(sample = eqprob$sites.base, population = NE_lakes)
pielou_grts
# compute for irs
pielou_irs <- spbalance(sample = eqprob_irs$sites.base, population = NE_lakes)
pielou_irs
```
Unsurprisingly, the GRTS design has better spatial balance than the IRS design. By default, `spbalance()` computes only the Pielou's evenness index, but several other spatial balance metrics are available -- run `?spbalance` for more information.

# References

Stevens Jr, D. L. and Olsen, A. R. (2004). Spatially balanced sampling of natural resources.
*Journal of the American Statistical Association*, 99(465):262-278.