---
title: "Selecting Spatially Balanced Samples from a Linear Network"
author: "Michael Dumelle, Tony Olsen, Tom Kincaid, and Marc Weber"
output: 
  html_document:
    theme: flatly
    number_sections: true
    highlighted: default 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Selecting Spatially Balanced Samples from an Linear Network}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
library(devtools)
load_all()
data("Illinois_River")
```

# Introduction

In this vignette, you will use `spsurvey` to select probability samples using the the GRTS algorithm (Stevens and Olsen, 2004) and independent random sampling. A greater focus will be placed on GRTS, as this algorithm yields samples having "spatial balance." More specifically, throughout this vignette, you will explore:

* [The `Illinois_River` Data] 
* [GRTS Designs]
    * [An Unstratified, Equal Probability GRTS Design] 
    * [An Unstratified, Unequal Probability GRTS Design]
    * [A Stratified, Equal Probability GRTS Design]
* [Independent Random Sampling Designs]
* [Additional Design Features]
    * [Legacy Sites]
    * [Minimum Distance Between Sites]
    * [Replacement Sites]
* [Calculating the Spatial Balance of a Sample]

If you have not yet done so, please read the "Start Here" vignette, accessed [here](dummy link for now) or by running `vignette("start_here", package = "spsurvey")`. 

# The `Illinois_River` Data

Before anything else, you must load `spsurvey` by running
```{r setup, eval = FALSE}
library(spsurvey)
```
If you want to reproduce the results from code in this vignette, you must also set a seed by running
```{r}
set.seed(5)
```

Throughout this vignette, you will use select probability samples from the `Illinois_River` dataset. The `Illinois_River` dataset is composed of 9087 line segments from the Illinois River in Arkansas and Oklahoma. To load the data into R, run
```{r load data, eval = FALSE}
data("Illinois_River")
```
To view the first few rows, run
```{r view data}
head(Illinois_River)
```
You will immediately notice some geometry information has been returned:

* `geometry type`: This is a `MULTILINESTRING` geometry because each portion of the river is a linear network.
* `dimension`: There are two dimensions, `X` (east-west) and `Y` (north-south).
* `bbox`: The bounding box of the spatial locations.
* `projected CRS`: The coordinates in a USA Contiguous Albers Equal Area Conic Projection.

The next few lines contain regarding variables in `Illinois_River`:

* `STATE_NAME`: The name of the state (Arkansas or Oklahoma) containing the river segment.
* `SO_ID`: A Strahler Order identifier.
* `geometry`: The projected spatial coordinates for each lake centroid.

Later you will sumamrize and visualize the `Illinois_River` data. Before doing so, you must store the data as an `sframe` object:
```{r}
Illinois_River <- sframe(Illinois_River)
```

# GRTS Designs

In this section, you will select spatially balanced samples -- but what exactly is a "spatially balanced sample?" And why are they important? Stevens and Olsen (2004) defined a spatially balanced sample as a sample mirroring the density of the population -- this is often refferred to as a sample that is a *miniature* of the population. This is different from spatial balance with respect to geography. To elucidate this distinction, consider the following plot of the `Illinois_River` locations:
```{r, fig.align = "center", fig.cap = "The Illinois River Data."}
plot(Illinois_River)
```
Notice that river segments have the highest density in the southwestern locations and the sparsest density in the central locations. A spatially balanced sample with respect to the population mirrors the density in the population, sampling more in the southwest and less in the central region, while a spatially balanced sample with respect to geography omits this density information but takes into account the absolute distances among neighboring river segments. These two types of spatial balance are useful in different contexts, and no type is superior over the other. But it is important to be aware of this distinction and its implications. Henceforth, when we refer to "spatial balance," we mean spatial balance with respect to the population.

Spatially balanced samples are so important because they generally yield estimators of means and totals *having narrower confidence intervals* than samples lacking spatial balance. This means our estimators are more *precise*, which can lead to reduces resource costs and a *more accurate representation* of the true, uknonwn values of the quantities we are studying.

## An Unstratified, Equal Probability GRTS Design

To select an unstratified, equal proability GRTS sample, run
```{r}
# equal probability design
eqprob <- grts(sframe = Illinois_River, seltype = "equal", n_base = 60)
```

* `seltype` is `"equal"` for an equal probability design
* `n_base` is 60, the desired sample size

`eqprob` contains an `sf` object, `eqprob$sites_base`, with the selected sites. To view the first few rows, run:
```{r}
head(eqprob$sites_base)
```

There are several variables contained in `eqprob$sites_base`:

* The row names of `eqprob$sites_base` match the row locations in `NE_lakes` so that you can easily identify sampled sites from your population
* `ip` is the site's original inclusion probability
* `wgt` is the site's survey design weight
* `stratum` indicates each site's strata; for this unstratified design, all entries in `stratum` are `"None"`
* `caty` indicates each site's unequal probability grouping; for this equal probability design, all entries in `caty` are `"None"`.
* Several other variables have `NA` values -- this just means they are not relevant to your design and can be ignored for now.

To overlay the spatially balanced sample on the original population, run
```{r, fig.align = "center", fig.cap = "Sampled sites (sites_base) from an equal probability GRTS design."}
plot(eqprob, Illinois_River, pch = 19, key.pos = 1)
```
The `pch = 19` argument tells `spsurvey` to plot closed circles instead of open circles (the default), and the `key.pos = 1` argument tells `spsurvey` to put the legend at the bottom of the plot.


`grts` returns several design elements used by the function as a tool to both keep track of and to verify the specifications of your design. You can access this information by running `eqprob$dsgn`. A list is printed containing several pieces of information -- several of which may be `NULL`. These `NULL` values are expected and change based on the specifications of your design, similar to how the columns having only `NA` in `eqprob$sites_base` will eventually change; this will be made more clear later on. For now, you will focus on the relevant non-`NULL` information.
```{r, echo = FALSE}
eqprob$dsgn$legacy_option <- NULL
for (s in names(eqprob$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob$dsgn[[s]])) == 0) eqprob$dsgn[[s]] <- NULL
}
eqprob$dsgn
```

* The `stratum` variable (the `stratum_var` argument to `grts()`) in `eqprob$sites_base` is given the name `stratum`
* All entries in `stratum` are given the same value: `"None"`
* The selection type for the `"None"` stratum is equal probability -- as indicated by `"equal"`
* There are 60 (`n_base`) samples in the `"None"` stratum


## An Unstratified, Unequal Probability GRTS Design

Suppose we wish to sample river segments in Oklahoma with a higher probability than river segments in Arkansas. To better understand our data, you can first summarize lakes in each elevation category:
```{r}
summary(Illinois_River, ~ STATE_NAME)
```
There are 9087 total river segments, 4717 of which are in the Arkansas and 4370 of which are from Oklahoma.
Next, you can plot the river segments in each state:
```{r, fig.align = "center", fig.cap = "Population sites by `ELEVATION_CAT` categories."}
plot(Illinois_River, formula = ~ STATE_NAME, key.pos = 1)
```

To select an unequal probability GRTS sample by state, run
```{r}
# expected sample size for each group with names
caty_n <- c(Arkansas = 20, Oklahoma = 40)
# selecting the unequal probability sample
uneqprob <- grts(sframe = Illinois_River, seltype = "unequal", 
                          n_base = 60, caty_var = "STATE_NAME", 
                          caty_n = caty_n)
```

* `seltype` is "unequal" for an unequal probaility design
* `n_base` is the desired **total** sample size, 60
* `caty_var` matches the name of the variable in `NE_lakes` with the unequal probability groups
* `caty_n` is a vector with expected sample sizes for each group in `caty_var`

    * The names of `caty.n` must match the names of each group in `caty_var`
    * The sum of `caty.n` values must equal `n_base`

Though the expected sample sizes for this design are specified by `caty.n`, there is no guarantee your within-group sample sizes will match these values **exactly**.

The relevant non-NULL `dsgn` output is
```{r, echo = FALSE}
uneqprob$dsgn$legacy_option <- NULL
for (s in names(uneqprob$dsgn)) {
    if (length(Filter(Negate(is.null), uneqprob$dsgn[[s]])) == 0) uneqprob$dsgn[[s]] <- NULL
}
uneqprob$dsgn
```


## A Stratified, Equal Probability GRTS Design

Unlike unequal probability designs, stratified designs ensure each within-group sample size equals the expected sample size. To select a spatially balanced, stratified design having equal probabilities of selection within each state, run
```{r}
# sample sizes of the strata
strata_n <- c("Arkansas" = 30, "Oklahoma" = 30)
# selecting the stratified, equal probability sample
strat_eqprob <- grts(sframe = Illinois_River, stratum_var = "STATE_NAME",
                        stratum = strata, seltype = "equal", 
                        n_base = strata_n)
```


* `seltype` is "equal" for equal probability of selection within strata
* `stratum_var` matches the name of the variable in `Illinois_River` with the strata
* `n_base` is a names vector whose names identify strata and whose values inidicate the within-strata sample sizes

The relevant non-NULL `dsgn` output is
```{r, echo = FALSE}
strat_eqprob$dsgn$legacy_option <- NULL
for (s in names(strat_eqprob$dsgn)) {
    if (length(Filter(Negate(is.null), strat_eqprob$dsgn[[s]])) == 0) strat_eqprob$dsgn[[s]] <- NULL
}
strat_eqprob$dsgn
```

To visualize the Oklahoma river segments separately for each state, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Sampled sites (sites_base) obtained for Oklahoma river segments using a stratified GRTS design having equal selection probabilities within each strata."}
plot(strat_eqprob, Illinois_River, sites ~ STATE_NAME, pch = 19, key.pos = 1)
```

# Independent Random Sampling Designs

If you want to conduct an independent random sample (IRS) instead of a GRTS sample, simply replace the `grts()` function with the `irs()` function. All of the arguments are identical, so any of the aforementioned examples can be easily applied to the IRS scenario as well. For example, to select an equal probability IRS, run
```{r}
# equal probability design
eqprob_irs <- irs(sframe = Illinois_River, seltype = "equal", n_base = 60)
```

The relevant non-NULL design output is
```{r, echo = FALSE}
eqprob_irs$dsgn$legacy_option <- NULL
for (s in names(eqprob_irs$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_irs$dsgn[[s]])) == 0) eqprob_irs$dsgn[[s]] <- NULL
}
eqprob_irs$dsgn
```

To overlay the sample on the original population, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Sampled sites (sites_base) obtained for an equal probability IRS design."}
plot(eqprob_irs, Illinois_River, pch = 19, key.pos = 1)
```

Notice how several of the sampled locations are closer together than they were for the unstratified, equal probability GRTS design.

# Additional Design Features

Besides selection probability type and stratification, there are several other design features `grts()` and `irs()` functions can accomodate (though we focus on applying the `grts()` function here): [Legacy Sites], [Minimum Distance Between Sites], and [Replacement Sites]. 

## Legacy Sites

Researchers will often want to ensure that a specific set of sites are always in the sample; these sites are known as legacy sites.  Legacy sites are commonly selected as part of a previous probability sample, and this is the scenario for which we focus.  Suppose a new `sf` object, `legacy_sites`, contains five randomly selected legacy sites.
```{r}
# randomly choosing 5 observations to treat as legacy sites
legacy_sites <- st_sample(Illinois_River, size = 5) # 5 random sites
legacy_sites <- subset(legacy_sites, !st_is_empty(legacy_sites)) # removing non-legacy geometries
legacy_sites <- st_cast(legacy_sites, "POINT") # moving from multipoint geometry to point
legacy_sites <- st_sf(geometry = legacy_sites) # storing as an sf object
# printing the legacy sites
legacy_sites
```

To accomodate legacy sites in an unstratified, equal probability sample, run
```{r}
eqprob_legacy <- grts(Illinois_River, seltype = "equal", n_base = 60,
                         legacy_sites = legacy_sites)
```

* `seltype` is `"equal"` for an equal probability design
* `n_base` is 60, the desired sample size
* `legacy_sites` is the `sf` object containing the legacy sites

The relevant non-null design output is
```{r, echo = FALSE}
for (s in names(eqprob_legacy$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_legacy$dsgn[[s]])) == 0) eqprob_legacy$dsgn[[s]] <- NULL
}
eqprob_legacy$dsgn
```

To plot the population sites next to the legacy and sampled sites, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Sampled sites (sites_base) and legacy sites (legacy) obtained for an equal probability GRTS design."}
plot(eqprob_legacy, Illinois_River, showlegacy = TRUE, pch = 19, key.pos = 1)
```

With some data, you may encounter a warning message saying `Of the x grid cells from which sample points were selected, y (y/x%) of the cells contained more than one sample point. This warning message exists because sometimes the requirement that legacy sites must be included prevents the sample from being as spatially balanced as a sample from a standard application of the GRTS algorithm. 

## Minimum Distance Between Sites

The standard GRTS algorithm may select sites that are closer together than desired. You can adjust the algorithm to require a minimum distance between sites.  Supppose we requires sites be 1600 meters apart. To include this requirement in an unstratified, equal probability design, run
```{r}
# setting the minimum distance
mindis <- 1600
eqprob_mindis <- grts(Illinois_River, seltype = "equal", n_base = 60,
                         mindis = mindis)
```

* `seltype` is `"equal"` for an equal probability design
* `n_base` is 60, the desired sample size
* `mindis` is the minimum distance desired

The relevant non-`NULL` design output is
```{r, echo = FALSE}
eqprob_mindis$dsgn$legacy_option <- NULL
for (s in names(eqprob_mindis$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_mindis$dsgn[[s]])) == 0) eqprob_mindis$dsgn[[s]] <- NULL
}
eqprob_mindis$dsgn
```


To plot the population sites next to the sampled sites, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Sampled sites (sites_base) for an equal probability GRTS design having a minimum distance between sites of 1600 meters."}
plot(eqprob_mindis, Illinois_River, pch = 19, key.pos = 1)
```
It is possible that the GRTS algorithm will not be able to ensure the minimum distance is met for all sampled sites.  In this context, the algorithm will attempt to ensure the minimum distance is met for as many sites as possible and return a warning stating that the requirement was not met after several iterations of the algorithm.

## Replacement Sites

It is often the case that a selected site is unable to be sampled. This can happen for a variety of reasons: landowner denial, lack of funding, or lack of resources, to name a few. If you expect there are sites you may not be able to sample, you can include potential replacement sites in your GRTS design. `spsurvey` gives you two methods to select these replacement sites: reverse heirarchical ordering and nearby neighbors.

First, we discuss reverse heirarchical ordering replacement. In this context, the number desired sample size, $n$, and the number of replacement sites, $n_r$, must be specified before the design is selected. A standard GRTS algorithm is used to select a sample of size $n + n_r$. Then, these sites are placed into "reverse heirarchical order" and the $n$ base sites and $n_r$ replacement sites are classified. The $n$ base sites should be sampled first. If any of these $n$ sites are unable to be sampled, replacement sites should be used. For more details, see Stevens and Olsen (2004). To select a GRTS sample of size 60 with 10 reverse heirarchical ordering replacement sites, run
```{r}
eqprob_rho <- grts(Illinois_River, seltype = "equal", n_base = 60,
                         n_over = 10)
```

* `seltype` is `"equal"` for an equal probability design
* `n_base` is 60, the desired sample size
* `n_over` is the number of replacement sites using reverse heirarchical ordering

`eqprob_rho$sites_base` is the `sf` object containing the $n$ base sites, and `eqprob_reverse$sites_over` is the `sf` object containing the $n_r$ replacement sites. The relevant non-`NULL` design output is
```{r, echo = FALSE}
eqprob_rho$dsgn$legacy_option <- NULL
for (s in names(eqprob_rho$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_rho$dsgn[[s]])) == 0) eqprob_rho$dsgn[[s]] <- NULL
}
eqprob_rho$dsgn
```

To plot the population sites next to the sampled sites and the replacement sites, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Sampled sites (sites_base) and replacement sites (sites_over) for an equal probability GRTS design using reverse heirarchical ordering for the replacement sites."}
plot(eqprob_rho, Illinois_River, pch = 19, key.pos = 1)
```

The second method for selecting replacement sites is the nearby neighbors method. A base GRTS sample of size $n$ is first selected. If a site is then unable to be sampled, then nearby neighbor method provides $n_r$ (up to 10) replacement sites that are closest to the original, unsampled site. This implies that for each site in the GRTS sample, $n_r$ replacement sites are available. To select a GRTS sample of size 60 with a 10 nearby neighbors for each site, run
```{r}
eqprob_nn <- grts(Illinois_River, seltype = "equal", n_base = 60, n_near = 1)
```

* `seltype` is `"equal"` for an equal probability design
* `n_base` is 60, the desired sample size
* `n_over` is the number of replacement sites using nearby neighbors

`eqprob_rho$sites_base` is the `sf` object containing the $n$ base sites, and `eqprob_reverse$sites_over` is the `sf` object containing the $n_r$ replacement sites. This nearest neighbor method is more computationally intensive than the reverse heirarchical ordering method, and the above code can take a almost a minute to run on your machine. The relevant non-`NULL` design output is
```{r, echo = FALSE}
eqprob_nn$dsgn$legacy_option <- NULL
for (s in names(eqprob_nn$dsgn)) {
    if (length(Filter(Negate(is.null), eqprob_nn$dsgn[[s]])) == 0) eqprob_nn$dsgn[[s]] <- NULL
}
eqprob_nn$dsgn
```

To view the population sites, base sample sites, and each sampled site's nearest replacement site, run
```{r, fig.show = "hold", fig.align = "center", fig.cap = "Sampled sites (sites_base) and replacement sites (sites_near) for an equal probability GRTS design using nearby neighbors for the replacement sites."}
plot(eqprob_nn, Illinois_River, pch = 19, key.pos = 1)
```
You will notice that these replacement sites are very close to the originally sampled sites.


# Calculating the Spatial Balance of a Sample

To measure the spatial balance of a design, Stevens and Olsen (2004) proposed using a set of Voroni polygons. The Voroni polygon for a sampled site contains all linear segments closer to this sampled site than any other sampled site. The set of Voroni polygons for all sampled sites will contain the entire population. If a sample is spatially balanced, we would expect that the proportion of the total length in each Voroni polygon to be approximately $1/n$.

Let $p_i$ be the proportion of length in Voroni polygon $i$. The Pielou's evenness index is often used as a measure of spatial balance and can be defined as $1 + \frac{1}{n}\sum_{i = 1}^n p_i log(p_i)$. One attractive feature of Pielou's evenness index is that it is unitless; a value of 0 indicates perfect spatial balance, and a value of 1 indicates no spatial balance. To compute the Pielou's evenness index of the equal probabily GRTS and IRS designs, run
```{r}
# compute for grts
pielou_grts <- spbalance(eqprob, Illinois_River)
pielou_grts
# compute for irs
pielou_irs <- spbalance(eqprob_irs, Illinois_River)
pielou_irs
```
Unsurprisingly, the GRTS design has better spatial balance than the IRS design. By default, `spbalance()` computes only the Pielou's evenness index, but several other spatial balance metrics are available -- run `?spbalance` for more information.

# References

Stevens Jr, D. L. and Olsen, A. R. (2004). Spatially balanced sampling of natural resources.
*Journal of the American Statistical Association*, 99(465):262-278.