---
title: "Selecting and Analyzing Finite Population Designs using the spsurvey R package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{New_Finite_Points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
set.seed(1)
```

# Finite Designs

In this vignette, you will learn how to:

* Select a variety of probability samples using the GRTS algorithm.
* Analyze those probability samples.

First, you will load the `spsurvey` package:
```{r setup}
# loading spsurvey
library(spsurvey)
```

You will explore the NLA Lakes 2007 dataset throughout this vignette. More information can be found using `?NLA_2007`.
```{r load NLA}
# loading the data
data(NLA_2007)

# printing the first few rows
head(NLA_2007)
```

Beforing moving forward, we must store this data as an `sf` object. For more information about converting your data to an `sf` object, please visit the INSERT NAME vignette using INSERT VIGNETTE ACCESS CODE.

```{r}
# storing the data frame as an sf object
NLA_2007 <- st_as_sf(NLA_2007, coords = c("xcoord", "ycoord"), crs = 5069)
```

You will only a few of the column variables throughout this vignette. To subset the data, run:
```{r}
# subsetting the data to include relevant columns
NLA_2007 <- NLA_2007[, c("siteID", "wgt", "Lake_Origin", "Chla_cond", "Chla")]
```


```{r, include = FALSE}
# this will get taken out once I realize how to use the proper functions
pathgrts <- "C:/Users/mdumelle/Documents/grts_dev/grts_pts_test/test_07032020/"
source(paste0(pathgrts, "grtspts.R"))
source(paste0(pathgrts, "grtspts_returnused.R"))
source(paste0(pathgrts, "grtspts_ip.R"))
source(paste0(pathgrts, "grtspts_stratum.R"))
source(paste0(pathgrts, "grtspts_ipleg.R"))
source(paste0(pathgrts, "grtspts_select.R"))
source(paste0(pathgrts, "grtspts_mindis.R"))
source(paste0(pathgrts, "make_grid.R"))
source(paste0(pathgrts, "numLevels.R"))
source(paste0(pathgrts, "cellWeight.R"))
source(paste0(pathgrts, "dsgn_check.R"))
source(paste0(pathgrts, "stopprnt.R"))
source(paste0(pathgrts, "warnprnt.R"))
source(paste0(pathgrts, "constructAddr.R"))
source(paste0(pathgrts, "ranho.R"))
source(paste0(pathgrts, "pickGridCells.R"))
source(paste0(pathgrts, "pickFiniteSamplePoints.R"))
source(paste0(pathgrts, "replace_sites.R"))
```

## Using the GRTS Algorithm


The GRTS algorithm is a flexible tool you can use for a variety of contexts. You will first learn a variety of selection methods for single-strata designs:

* Equal probability of selection.
* Unequal probability of selection.
* Probability of selection proportional to an auxiliary variable.

You will then use the same selection methods for stratified designs.  Then, you use some of the additional features available in the GRTS algorithm to learn how to:

* Incorporate legacy sites (existing sites).
* Ensure a minimum distance between any two sites.
* Select replacement sites in the event that you are unable to visit originally sampled sites.

### Single-Strata Designs

#### Equal Probability of Selection

To select an equal proability GRTS sample, run:
```{r}
# equal probability design
eqprob <- grtspts(sframe = NLA_2007, seltype = "equal", nsamp = 30)
```
The function arguments are:

* `sframe` the `sf` object
* `seltype` the type of selection. `seltype = "equal"` is the default, so this argument did not need to be explicitly specified.
* `nsamp` the desired sample size

The output is a list containing two objects: `sites:base` and `dsgn`. `sites:base` is an `sf` object containing information about the sample.
```{r}
# storing the sites:base information
eqprob_sites <- eqprob$sites.base
# printing the first few rows
eqprob_sites
```

* The rownames of `eqprob_sites` match the rows locations from the original `sf` object. 
* `siteID` is the ID used to identify the ordering of sites in the GRTS sample.  This is a different variable than the `siteID` from the original `sf` object. These indices can be accessed with and used to recover the original site ID's using 
`NLA_2007[as.numeric(rownames(eqprob_sites)), "siteID"]`
* `replsite` is NA for all rows because we did not implement the replacement site option.  
* `siteuse` is `"Base"` for all rows because these samples are all part of the base sample (i.e. there are no replacement sites)
* `stratum` equals `None` for all rows because this is not a stratified design
* `wgt` is the vector of probability weights indicating the number of population units the site represents (equal to $N/n$, where $N$ is the number of population units and $n$ is the sample size). This is a different variable than the `wgt` from the original `sf` object.
* `ip` is the vector of inclusion probabilities indicating the probability of each unit being selected in the design (equal to $n/N$)
* `caty` equals `None` for all rows because this design does not incorporate unequal selection probabilities
* `aux` is NA for all rows because we did not implement the auxiliary site option for proportional sampling. 
* `legacy` is NA for all rows because we did not implement the legacy site option.  
* The rest of the variables are from the original `sf` object.

`dsgn` is a list containing information about the design specificiations. You should use this to verify the algorithm used reflects the intended design.

```{r}
# storing the design information
eqprob_dsgn <- eqprob$dsgn

# printing the output
eqprob_dsgn
```

* `stratum_var` The stratum variable in `eqprob_sites` indicating the name of the stratum variable.
* `caty_var`: The unequal probability variable in `eqprob_sites`.
* `aux_var` The auxiliary variabile `eqprob_sites` for proportional sampling.
* `legacy_var` The legacy variable in `eqprob_sites`.
* `stratum` The names of the strata
* `seltype` The selection type for each strata
* `nsamp` The sample size for each strata
* `caty.n` The expected sample size for each unequal probability category within each strata
* `over.n` The number of replacement sites using method1 for each strata
* `over.near`The number of replacement sites using method2 for each strata
* `mindis` The minimum distance between sites for each strata

#### Unequal Probability of Selection

Suppose we want to sample natural lakes twice as often as man-made lakes. To select an unequal probability GRTS sample, run:
```{r}
# expected sample size for each group
caty.n <- c(20, 10)
# names of the categories
names(caty.n) <- c("Natural", "Man-Made")
# selecting the unequal probability sample
uneqprob <- grtspts(sframe = NLA_2007, seltype = "unequal", 
                          nsamp = 30, caty_var = "Lake_Origin", 
                          caty.n = caty.n)
```

The new function arguments are:

* `caty_var`: A character vector specifying the name of the variable containing the unequal probability categories.
* `caty.n`: A character vector specifying the number of expected samples for each unequal probability category. This vector must be named, and the names must match the names the `"Lake_Origin"` variable of `NLA_2007`.

`uneqprob$sites.base` and `uneqprob$dsgn` contain the sampled site and design information, respectively.

#### Probability of Selection Proportional to an Auxiliary Variable

It is useful to sample proportional to a continuous auxiliary variable when there is a positive correlation between the response variable and the auxiliary variable because this can yield more precise estimates. To select a sample proportional to a continuous auxiliary variable, run:

```{r, include = FALSE}
# selection probability proportional to the Chla variable 
propprob <- grtspts(sframe = NLA_2007, seltype = "proportional", 
                          nsamp = 30, aux_var = "Chla") 

```

The new function arguments are:

* `aux_var`: A categorical variable specifying the continuous variable in `NLA_2007` to use for proportional selection.

`propprob$sites.base` and `propprob$dsgn` contain the sampled site and design information, respectively.

### Stratified Designs

#### Equal Probability of Selection

To select a stratified GRTS sample with equal probabilities of selection within strata, run:

```{r}
# names of the strata
strata <- c("Natural", "Man-Made")
# sample sizes of the strata
strata.n <- c(15, 15)
# selecting the stratified, equal probability sample
strat_eqprob <- grtspts(sframe = NLA_2007, stratum_var = "Lake_Origin",
                        stratum = strata, seltype = "equal", 
                        nsamp = strata.n)
```

The new function arguments are:

* `stratum_var`: The name of the strata variable in `NLA_2007`.
* `stratum`: The names of the strata in `"Lake_Origin"`
* `nsamp`: A vector whose elements specify the sample size in each strata as defined by the `stratum` argument

`strat_eqprob$sites.base` and `strat_eqprob$dsgn` contain the sampled site and design information, respectively.

#### Unequal Probability of Selection

To select a stratified GRTS sample with unequal (and possibly varying) unequal probabilities within strata, run:
```{r}
# names of the strata
strata <- c("Natural", "Man-Made")
# sample sizes of the strata
strata.n <- c(15, 15)
# expected number of samples in the natural strata for the Chla_cond variable
natural_caty.n <- c(9, 6)
# names of the Chla_cond variable
names(natural_caty.n) <- c("Good", "Poor")
# expected number of samples in the man-made strata for the Chla_cond variable
manmade_caty.n <- c(7, 8)
# names of the Chla_cond variable
names(manmade_caty.n) <- c("Good", "Poor")
# list of the unequal probability expected samples for each strata
caty.n_list <- list(natural_caty.n, manmade_caty.n)
# names of the strata
names(caty.n_list) <- c("Natural", "Man-Made")
# selecting the stratified, unequal probability sample
strat_uneqprob <- grtspts(sframe = NLA_2007, stratum_var = "Lake_Origin",
                        stratum = strata, seltype = "unequal", 
                        nsamp = strata.n, caty_var = "Chla_cond", 
                        caty.n = caty.n_list)
```

The new function arguments are:

* `caty_var`: The variable in the dataset for the unequal probability categories
* `caty.n`: A list whose names match the names of the `stratum` variable.  The elements of the list are vectors whose entries represent the expected sample sizes for the unequal probability classes within strata.  The names of these vectors must equal the names of the unqeual probability categories.

`strat_uneqprob$sites.base` and `strat_uneqprob$dsgn` contain the sampled site and design information, respectively.

#### Probability of Selection Proportional to an Auxiliary Variable

To select a stratified GRTS sample where each strata is sampled proportionally to a continuous auxiliary variable, run:

```{r}
# selecting the stratified, proportional probability sample
strat_propprob <-  grtspts(sframe = NLA_2007, stratum_var = "Lake_Origin",
                        stratum = strata, seltype = "proportional", 
                        nsamp = strata.n, aux_var = "Chla")
```

The new function arguments are:

* `aux_var`: The auxiliary variable in the dataset

`strat_propprob$sites.base` and `strat_propprob$dsgn` contain the sampled site and design information, respectively. It is not currently possible to choose different auxiliary variables for different strata.