---
title: "Spatially Balanced Sampling"
author: "Michael Dumelle, Tony Olsen, Tom Kincaid, and Marc Weber"
output: 
  html_document:
    theme: flatly
    number_sections: true
    highlighted: default 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Spatially Balanced Sampling}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
```

# Introduction

The Generalized Random Tessellation Stratified (GRTS) algorithm (Stevens and Olsen, 2004, Olsen et.al., 2012) is a spatially balanced sampling algorithm available in spsurvey. The GRTS algorithm is used to sample from finite populations (point resources -- e.g. lakes treated as a whole) and infinite populations (linear resources -- e.g. rivers and streams; areal resources -- e.g. wetlands and coastlines) and is implemented in spsurvey using the `grts()` function. The output from `grts()` contains the design sites and additional information about the design. More specifically, it is a list with five elements:

1. `sites_legacy`: legacy sites included in the base (main) sample (see [Legacy sampling])
2. `sites_base`: sites included in the base (main) sample that are not legacy sites
3. `sites_over`: replacement sites obtained using reverse hierarchical ordering (see [Reverse hierarchical ordering])
4. `sites_near`: replacement sites obtained using nearest neighbor (see [Nearest neighbor])
5. `design`: a list with specific details about the sampling design
    
The `NE_Lakes` sampling frame in spsurvey is used in this vignette to introduce spatially balanced sampling via the GRTS algorithm. `NE_Lakes` is a finite sampling frame and contains lakes (treated as a whole) from the Northeastern United States. There are four variables in `NE_Lakes` used in this vignette:

1. `AREA`: lake area in hectares
2. `AREA_CAT`: lake area categories (small and large)
3. `ELEV_CAT`: lake elevation categories (low and high)
4. `LEGACY`: a legacy site indicator

Before proceeding, we load spsurvey and set a reproducible seed:
```{r}
library(spsurvey)
set.seed(5)
```

# Unstratified sampling

## Equal inclusion probabilities

To select an unstratified GRTS sample where each site in the sampling frame has an equal probability of being included in the sample (inclusion probability), run
```{r}
eqprob <- grts(NE_Lakes, n_base = 50)
```

The first argument to `grts()` is the sampling frame. The second argument is `n_base`, which specifies the number of sites in the base (main) sample. The `sites_base` object in `eqprob` is an `sf` object and contains the original columns of `NE_Lakes` as well as a few additional columns such as a site identifier, latitude and longitude coordinates, inclusion probabilities, and design weights (to be used in analyses after collecting data). To visualize the design sites overlain onto the sampling frame, run
```{r}
sp_plot(eqprob, NE_Lakes, key.width = lcm(3))
```

The `key.width` argument extends the plot's margin to fit the legend text nicely within the plot. 

## Unequal inclusion probabilities

To select an unstratified GRTS sample where each site in the sampling frame has unequal inclusion probabilities according to some categorical variable, run
```{r}
caty_n <- c(small = 40, large = 10)
uneqprob <- grts(
  NE_Lakes,
  n_base = 50,
  caty_var = "AREA_CAT",
  caty_n = caty_n
)
```

`caty_var` is the unequal inclusion probability variable (`"AREA_CAT"` in `NE_Lakes`), and `caty_n` is a vector whose names are the levels in `caty_var` and values are the expected sample sizes. In this sample, inclusion probabilities are adjusted so that on average, there are 40 small lakes and 10 large lakes selected. 

## Proportional inclusion probabilities

To select an unstratified GRTS sample where each site in the sampling frame has inclusion probability proportional to a positive, continuous variable, run
```{r}
propprob <- grts(
  NE_Lakes,
  n_base = 50,
  aux_var = "AREA"
)
```

`aux_var` is the proportional probability variable (`"AREA"` in `NE_Lakes`). Proportional (to size) inclusion probabilities are useful because they can increase the precision of estimates when the response variable is positively correlated with the proportional probability variable.

# Stratified sampling

Stratified sampling designs partition the sampling frame into distinct groups called strata and select sites within each stratum independently of other strata. Stratified sampling is useful from a practical perspective because it allows for stratum-specific sample sizes and implementation practices (e.g. each region may have different sampling protocols). Stratified sampling is useful from a statistical perspective because they generally yield more precise estimates than estimates from unstratified sampling.

## Equal inclusion probabilities

To select a sample stratified by elevation category with equal inclusion probabilities within each stratum, run
```{r}
strata_n <- c(low = 25, high = 15)
strat_eqprob <- grts(NE_Lakes, n_base = strata_n, stratum_var = "ELEV_CAT")
```
`strata_n` is a named vector whose names represent the strata and whose values represent strata-specific sample sizes, and `stratum_var` is the stratification variable in `NE_Lakes`.

To visualize the design sites overlain onto the sampling frame (separately for each stratum), run
```{r}
sp_plot(
  strat_eqprob,
  formula = siteuse ~ ELEV_CAT,
  NE_Lakes,
  key.width = lcm(3)
)
```

## Unequal inclusion probabilities

To select a sample stratified by elevation category with unequal inclusion probabilities for each area category, run
```{r}
caty_n <- list(
  low = c(small = 20, large = 10),
  high = c(small = 15, large = 5)
)
strat_uneqprob <- grts(
  NE_Lakes,
  n_base = strata_n,
  stratum_var = "ELEV_CAT",
  caty_var = "AREA_CAT",
  caty_n = caty_n
)
```

`caty_n` is now a list: the first element contains the expected sample sizes for each area category in the low elevation stratum, and the second element contains the expected sample sizes for each area category in the high elevation stratum.

## Proportional inclusion probabilities

To select a sample stratified by elevation category with probabilities proportional lake area, run
```{r}
strat_propprob <- grts(
  NE_Lakes,
  n_base = strata_n,
  stratum_var = "ELEV_CAT",
  aux_var = "AREA"
)
```

# Additonal sampling options

## Legacy sampling

Legacy (historical) sites are sites that have been selected from a previous sampling design and should be always be included in the current sample. spsurvey accommodates legacy sites for finite populations slightly differently than for infinite populations. For finite populations, a variable in the sampling frame must be included that indicates whether or not a site is a legacy site: 
```{r}
legacy <- grts(NE_Lakes, n_base = 50, legacy_var = "LEGACY")
```

To visualize the legacy and base sites together, run
```{r}
sp_plot(legacy, key.width = lcm(3))
```

These points can be overlain onto the sampling frame by including `NE_Lakes` in the `sp_plot()` command.

For infinite populations, a separate `sf` object that contains the legacy sites must be provided to `grts()` using the `legacy_sites` argument. An example is provided in [Infinite Population Sampling].

## Minimum distance sampling

Though the GRTS algorithm selects spatially balanced samples, the algorithm can select sites that are closer together than a user may desire. To enforce a minimum distance between design sites, run
```{r}
mindis <- grts(NE_Lakes, n_base = 50, mindis = 1600)
```

Here we have specified that sites be no closer together than 1600 meters (meters are the units of the `sf` object). In some situations, the `grts()` function  will fail to enforce the minimum distance requirement for all sites. When this occurs, the function will enforce the requirement for as many sites as possible and then return a warning.

When stratifying, `mindis` will apply to both strata if it is a single value. It is possible to set stratum-specific minimum distance requirements by running
```{r}
mindis_list <- list(low = 1400, high = 1000)
strat_mindis <- grts(
  NE_Lakes,
  strata_n,
  stratum_var = "ELEV_CAT",
  mindis = mindis_list
)
```

## Replacement sampling

Sometimes data cannot be collected at a site. This may occur for a variety of reasons, some of which include landowner denial or inaccessible sites. The `grts()` function has options for selecting replacement sites: sites available to replace sites in the base (main) sample for which data cannot be collected. 

### Reverse hierarchical ordering

The first replacement site option is reverse hierarchical ordering (Stevens and Olsen, 2004). To select a base sample with reverse hierarchically ordered replacement sites, run
```{r}
rho_replace <- grts(NE_Lakes, n_base = 50, n_over = 25)
```
`n_base` indicates the desired sample size, and `n_over` indicates the number of replacement sites. Sites are first selected using the GRTS algorithm for a sample size of `n_base + n_over`. They are then determined as base sites or replacement sites in a way that preserves as much spatial balance as possible. The spatial balance of the base sites degrades as `n_over` increases, however, so it is important to choose a realistic value for `n_over`.

To visualize the base sites and reverse hierarchically ordered replacement sites, run
```{r}
sp_plot(rho_replace, key.width = lcm(3))
```

When stratifying, `n_over` will apply to both strata if it is a single value. It is possible to set stratum-specific reverse hierarchical ordering requirements by running
```{r}
over_list <- list(low = 2, high = 5)
strat_rho_replace <- grts(
  NE_Lakes,
  strata_n,
  stratum_var = "ELEV_CAT",
  n_over = over_list
)
```

### Nearest neighbor

The second replacement site option is nearest neighbor. To select a base sample with nearest neighbor replacement sites, run
```{r}
nn_replace <- grts(NE_Lakes, n_base = 50, n_near = 1)
```
`n_base` indicates the desired sample size, and `n_near` indicates the number of replacement sites for each base site. For `n_near = 1`, each site in the base sample has a replacement site associated with it -- this replacement site is the closest site (measured by Euclidean distance) to the base site (within a stratum).

When stratifying, `n_near` will apply to both strata if it is a single value. It is possible to set stratum-specific nearest neighbor requirements by running
```{r}
near_list <- list(low = 1, high = 2)
strat_nn_replace <- grts(
  NE_Lakes,
  strata_n,
  stratum_var = "ELEV_CAT",
  n_near = near_list
)
```

## Independent Random Sampling

If you do not want to select spatially balanced samples, use spsurvey's `irs()` function (shorthand for Independent Random Sampling). The function arguments for the `irs()` function are the same as the function arguments for the `grts()` function. This means that the flexible sampling design options available for the GRTS algorithm are also available for the IRS algorithm. To select an unstratified IRS sample where each site in the sampling frame has an equal inclusion probability, run
```{r}
eqprob_irs <- irs(NE_Lakes, n_base = 50)
```

To visualize the IRS design sites, run
```{r}
sp_plot(eqprob_irs, NE_Lakes, key.width = lcm(3))
```

Notice how these IRS design sites appear less spread out than the design sites from the unstratified GRTS sample with equal inclusion probabilities.

To select an IRS sample stratified by elevation category with equal inclusion probabilities within each stratum, run
```{r}
strata_n <- c(low = 25, high = 15)
strat_eqprob_irs <- irs(NE_Lakes, n_base = strata_n, stratum_var = "ELEV_CAT")
```

# Measuring spatial balance

So far we have talked generally about spatial balance but not how to measure it. Stevens and Olsen (2004) proposed measuring spatial balance using Voroni polygons. A design site's Voroni contains all sites in the sampling frame closer to this design site than any other design site. These Voroni polygons are combined with each site's inclusion probability and loss metric to measure spatial balance. 

Several spatial balance metrics are available in spsurvey. The default is Pielou's evenness index. Pielou's evenness index ranges from 0 (perfect spatial balance) to 1 (a complete lack of spatial balance). Larger values of the Pielou's evenness index indicate worse spatial balance. To measure the spatial balance of the unstratified GRTS and IRS samples with equal inclusion probabilities, run
```{r}
sp_balance(eqprob$sites_base, NE_Lakes) # grts
sp_balance(eqprob_irs$sites_base, NE_Lakes) # irs
```
The GRTS design sites are more spatially balanced than the IRS design sites.

In a stratified sample, spatial balance is calculated separately for each stratum:
```{r}
sp_balance(strat_eqprob$sites_base, NE_Lakes, stratum_var = "ELEV_CAT") # grts
sp_balance(strat_eqprob_irs$sites_base, NE_Lakes, stratum_var = "ELEV_CAT") # irs
```
The GRTS design sites are more spatially balanced than the IRS design sites in both strata.

# Infinite population sampling

Though this vignette used a finite population to illustrate the sampling options available in spsurvey, the same commands are used for sampling from infinite populations. The `Illinois_River` `sf` object in spsurvey is a linear resource and the `Lake_Ontario` `sf` object in spsurvey is an areal resource -- both are examples of infinite populations. To select a GRTS sample with equal inclusion probabilities from `Illinois_River`, run
```{r}
eqprob <- grts(Illinois_River, n_base = 50)
```

To visualize the design sites overlain onto the sampling frame, run
```{r}
sp_plot(eqprob, Illinois_River, key.width = lcm(3))
```

To accommodate legacy sites when sampling from an infinite population, an `sf` object containing the legacy sites as `POINT` (or `MULTIPOINT`) geometries must be provided to the `legacy_sites` argument:
```{r}
legacy <- grts(Illinois_River, n_base = 50, legacy_sites = Illinois_River_Legacy)
```

To select a GRTS sample with equal inclusion probabilities from `Lake_Ontario`, run
```{r}
eqprob <- grts(Lake_Ontario, n_base = 50)
```

To visualize the design sites (with closed circles) overlain onto the sampling frame, run
```{r}
sp_plot(eqprob, Lake_Ontario, pch = 19, key.width = lcm(3))
```

# Binding design sites

To bind `sites_legacy`, `sites_base`, `sites_over`, `sites_near` (four separate `sf` objects) into a single `sf` object, use `sp_rbind()`. For example, to combine the base and reverse hierarchically ordered replacement sites from `rho_replace` into a single `sf` object, run 
```{r}
combined <- sp_rbind(rho_replace)
```

Then it is straightforward to write out a single `sf` object using a function like `sf::write_sf()`. For example, to save the combined `sf` object as a shapefile named `"file_name.shp"` at the location on your machine called `"file_path"`, run
```{r, eval = FALSE}
write_sf(combined, "file_path/file_name.shp")
```

# References

Olsen, A. R., T. M. Kincaid and Q. Payton (2012). Spatially balanced survey designs for natural resources. Design and Analysis of Long-Term Ecological Monitoring Studies. R. A. Gitzen, J. J. Millspaugh, A. B. Cooper and D. S. Licht. Cambridge, UK, Cambridge University Press: 126-150.

Stevens Jr, D. L. and Olsen, A. R. (2004). Spatially balanced sampling of natural resources. *Journal of the American Statistical Association*, 99(465):262-278.
