---
title: "Spatially Balanced Sampling"
author: "Michael Dumelle, Tony Olsen, Tom Kincaid, and Marc Weber"
output: 
  html_document:
    theme: flatly
    number_sections: true
    highlighted: default 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Spatially Balanced Sampling}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
```

The Generalized Random Tessellation Stratified (GRTS) algorithm is a spatially balanced sampling algorithm available in `spsurvey` (Stevens and Olsen, 2004, Olsen et.al., 2012). The GRTS algorithm is used to sample from finite populations (point resources -- e.g. lakes treated as a whole) and infinite populations (linear resources -- e.g. rivers and streams; areal resources -- e.g. wetlands) and is implemented in `spsurvey` using the `grts()` function. The output from the `grts()` is a list with five elements:

1. `sites_legacy`: legacy sites included in the sample (see [Legacy sampling])
2. `sites_base`: sites included in the base (main) sample
3. `sites_over`: replacement sites obtained using reverse hierarchical ordering (see [Replacement sampling])
4. `sites_near`: replacement sites obtained using nearest neighbor (see [Replacement sampling])
5. `design`: a list with specific details regarding the sampling procedure that can be used to verify the sampling was performed as you intended
    
The `NE_Lakes` sample frame in `spsurvey` is used in this vignette to introduce spatially balanced sampling. The `NE_Lakes` sample frame is a finite and contains lakes (treated as a whole) from the Northeastern United States. There are four variables in `NE_Lakes` used in this vignette:

1. `AREA`: lake area in hectares
2. `AREA_CAT`: lake area categories (small and large)
3. `ELEV_CAT`: lake elevation categories (low and high)
4. `LEGACY`: a legacy site indicator

Before proceeding, we load `spsurvey` and set a reproducible seed:
```{r}
library(spsurvey)
set.seed(5)
```

# Unstratified sampling

## Equal inclusion probabilities

To select an unstratified GRTS sample where each site in the sample frame has an equal probability of being included in the sample (inclusion probability), run
```{r}
eqprob <- grts(NE_Lakes, n_base = 50)
```

The first argument to `grts()` is the sample frame and `n_base` specifies the number samples. The `sites_base` object in `eqprob` is an `sf` object and contains the original columns of `NE_Lakes` as well as a few additional columns such as a site identifier, latitude and longitude coordinates, inclusion probabilities, and sampling weights (to be used in analyses after collecting data). To visualize the sample overlain onto the sample frame, run
```{r}
plot(eqprob, NE_Lakes, pch = 19, key.width = lcm(3))
```

## Unequal inclusion probabilities

To select an unstratified GRTS sample where each site in the sample frame has unequal inclusion probabilities relative to a categorical variable, run
```{r}
caty_n <- c(small = 40, large = 10)
uneqprob <- grts(
  NE_Lakes,
  n_base = 50,
  seltype = "unequal",
  caty_var = "AREA_CAT",
  caty_n = caty_n
)
```

`seltype` is `"unequal"` to indicate unequal inclusion probabilities, `caty_var` is the unequal inclusion probability variable (`"AREA_CAT"` in `NE_Lakes`), and `caty_n` is a vector whose names are the levels in `caty_var` and values are the expected sample sizes. In this sample, inclusion probabilities are adjusted so that there will be approximately 40 small lakes and 10 large lakes selected. 

## Proportional inclusion probabilities

To select an unstratified GRTS sample where each site in the sample frame has inclusion probability proportional to a positive, continuous variable, run
```{r}
propprob <- grts(
  NE_Lakes,
  n_base = 50,
  seltype = "proportional",
  aux_var = "AREA"
)
```

`seltype` is `"proportional"` to indicate proportional inclusion probabilities, and `aux_var` is the proportional probability variable (`"AREA"` in `NE_Lakes`). Proportional (to size) inclusion probabilities are useful because they can increase the precision of estimates when the response variable is positively correlated with the auxiliary variable.

# Stratified sampling

In stratified sampling, the sample frame is partitioned into distinct groups called strata. Then samples are selected independently within each stratum. Stratified samples are useful from a practical perspective because they allow for stratum-specific sample sizes and implementation practices (e.g. each region may have different sampling protocols). Stratified samples are useful from a statistical perspective because they increase precision of estimates.

## Equal inclusion probabilities

To select a sample stratified by elevation category with equal inclusion probabilities within strata, run
```{r}
strata_n <- c(low = 25, high = 25)
strat_eqprob <- grts(NE_Lakes, n_base = strata_n, stratum_var = "ELEV_CAT")
```
`strata_n` is a named vector whose names represent the strata and whose values represent strata-specific sample sizes, and `stratum_var` is the stratification variable in `NE_Lakes`.

To visualize the sample overlain onto the sample frame (separately for each stratum), run
```{r}
plot(
  strat_eqprob,
  formula = siteuse ~ ELEV_CAT,
  NE_Lakes,
  pch = 19,
  key.width = lcm(3)
)
```

## Unequal inclusion probabilities

To select a sample stratified by elevation category with unequal inclusion probabilities for each area category, run
```{r}
caty_n <- list(
  low = c(small = 20, large = 10),
  high = c(small = 15, large = 5)
)
strat_uneqprob <- grts(
  NE_Lakes,
  n_base = strata_n,
  seltype = "unequal",
  stratum_var = "ELEV_CAT",
  caty_var = "AREA_CAT",
  caty_n = caty_n
)
```

`caty_n` is now a list: the first element contains the expected sample sizes for each area category in the low elevation stratum, and the second element contains the expected sample sizes for each area category in the high elevation stratum.

## Proportional inclusion probabilities

To select sample stratified by elevation category with probabilities proportional lake area, run
```{r}
strat_propprob <- grts(
  NE_Lakes,
  n_base = strata_n,
  stratum_var = "ELEV_CAT",
  seltype = "proportional",
  aux_var = "AREA"
)
```

# Additonal sampling options

## Legacy sampling

Legacy sites are sites that have been selected in a previous probability sample and are to be automatically included in a current probability sample. `spsurvey` accommodates legacy sites for finite populations slightly differently than for infinite populations. For finite populations, a variable in the sample frame must be included that indicates whether or not a site is a legacy site: 
```{r}
legacy <- grts(NE_Lakes, n_base = 45, legacy_var = "LEGACY")
```

To plot the legacy and base sites, run
```{r}
plot(legacy, pch = 19, key.width = lcm(3))
```
These points can be overlain onto the sample frame by including `NE_Lakes` in the `plot()` command.

For infinite populations, a separate `sf` object that contains the legacy sites (as `POINT` or `MULTIPOINT` geometries) must be provided to `grts()` using the `legacy_sites` argument.

## Minimum distance sampling

Though the GRTS algorithm selects spatially balanced samples, the algorithm can select sites that are closer together than a user may desire. To enforce a minimum distance between sampled sites, run
```{r}
mindis <- grts(NE_Lakes, n_base = 50, mindis = 1600)
```

Here we have specified that sites be no closer together than 1600 meters. In some situations, the `grts()` function may not be able to enforce the minimum distance requirement for all sites. When this occurs, the function will enforce the requirement for as many sites as possible and then return a warning.

## Replacement sampling

Sometimes sites are selected to be sampled but this sampling cannot occur. Sampling may not occur for a variety of reasons, some of which include landowner denial or inaccessible sites. The `grts()` function has options for selecting replacement sites, which are sites available to replace sites originally selected but cannot be sampled. The first replacement site option is the reverse hierarchical ordering option (Stevens and Olsen, 2004). To select a base sample with reverse hierarchically ordered replacement sites, run
```{r}
rho_replace <- grts(NE_Lakes, n_base = 50, n_over = 25)
```
`n_base` indicates the desired sample size, and `n_over` indicates the number of replacement sites. Sites are first selected using the GRTS algorithm for a sample size of `n_base + n_over`. Sites are then determined as base sites or replacement sites in a way that preserves as much spatial balance as possible. The spatial balance of the base sites degrades as `n_over` increases, however, so it is important to choose a realistic value for `n_over`.

To visualize the base sites and reverse hierarchically ordered replacement sites, run
```{r}
plot(rho_replace, pch = 19, key.width = lcm(3))
```

The second replacement site option is the nearest neighbor option. To select a base sample with replacement sites selected by nearest neighbor, run
```{r}
nn_replace <- grts(NE_Lakes, n_base = 50, n_near = 1)
```
`n_base` indicates the desired sample size, and `n_near` indicates the number of replacement sites for each base site. For `n_near = 1`, each site in the base sample has a replacement site associated with it -- this replacement site is the closest site (measured by Euclidean distance) to the base site.

## Independent Random Sampling

The `spsurvey` package also provides the `irs()` function to select non-spatially balanced samples using the Independent Random Sampling (IRS) algorithm. The syntax for the `irs()` function is equivalent to the syntax for the `grts()` function, so all of the previous options for the `grts()` function can be seamlessly applied to the `irs()` function. To select an unstratified IRS sample where each site in the sample frame has an equal inclusion probability, run
```{r}
eqprob_irs <- irs(NE_Lakes, n_base = 50)
```

To visualize the IRS sample, run
```{r}
plot(eqprob_irs, NE_Lakes, pch = 19, key.width = lcm(3))
```

Notice how this IRS sample appears less spread out than the unstratified GRTS sample with equal inclusion probabilities.

To select an IRS sample stratified by elevation category with equal inclusion probabilities within strata, run
```{r}
strata_n <- c(low = 25, high = 25)
strat_eqprob_irs <- irs(NE_Lakes, n_base = strata_n, stratum_var = "ELEV_CAT")
```


# Measuring spatial balance

So far this vignette has broadly mentioned spatial balance but has not yet discussed how to empirically measure it. Stevens and Olsen (2004) proposed measuring spatial balance using Voroni polygons. A Voroni polygon for a sampled site contains all sites in the sample frame closer to this sampled site than any other sampled site. The Voroni polygons are combined with a loss function to measure spatial balance. Several spatial balance metrics are available in `spsurvey`, but the default is Pielou's evenness index. Pielou's evenness index ranges from 0 (perfect spatial balance) to 1 (a complete lack of spatial balance). Larger values of the Pielou's evenness index indicate worse spatial balance. To measure the spatial balance of the unstratified GRTS and IRS samples with equal inclusion probabilities, run
```{r}
spbalance(eqprob$sites_base, NE_Lakes) # grts
spbalance(eqprob_irs$sites_base, NE_Lakes) # irs
```
The GRTS sample is more spatially balanced than the IRS sample.

Spatial balance is calculated separately for each stratum:
```{r}
spbalance(strat_eqprob$sites_base, NE_Lakes, stratum_var = "ELEV_CAT") # grts
spbalance(strat_eqprob_irs$sites_base, NE_Lakes, stratum_var = "ELEV_CAT") # irs
```
The GRTS sample is more spatially balanced than the IRS sample for both strata.

# Infinite population sampling

Though this vignette used a finite population to illustrate the sampling options available in `spsurvey`, the same commands are used for sampling from infinite populations. The `Illinois_River` `sf` object in `spsurvey` is a linear resource, and the `Great_Lakes` `sf` object in `spsurvey` is an areal resource -- both are examples of infinite populations. To select a GRTS sample with equal inclusion probabilities from `Illinois_River`, run
```{r}
eqprob <- grts(Illinois_River, n_base = 50)
```

To visualize the sample overlain onto the sample frame, run
```{r}
plot(eqprob, Illinois_River, pch = 19, key.width = lcm(3))
```

# References

Olsen, A. R., T. M. Kincaid and Q. Payton (2012). Spatially balanced survey designs for natural resources. Design and Analysis of Long-Term Ecological Monitoring Studies. R. A. Gitzen, J. J. Millspaugh, A. B. Cooper and D. S. Licht. Cambridge, UK, Cambridge University Press: 126-150.

Stevens Jr, D. L. and Olsen, A. R. (2004). Spatially balanced sampling of natural resources. *Journal of the American Statistical Association*, 99(465):262-278.
