---
title: "spsurvey - Start Here"
author: "Michael Dumelle, Tony Olsen, Tom Kincaid, and Marc Weber"
output: 
  html_document:
    theme: flatly
    number_sections: true
    highlighted: default 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
library(tidyverse)
library(devtools)
load_all()
```

# Installing and Loading `spsurvey`

If this is your first time using the `spsurvey` package, you will need to run:
```{r, eval = FALSE}
install.packages("spsurvey")
```
to install the package. You only need to run this code once per version of R.

After the `spsurvey` package is installed, you must load it into R each time you start a new R session. To load the `spsurvey` package into R, run
```{r setup, eval = FALSE}
library(spsurvey)
```


If you want to replicate the exact results in any of these vignettes at a later time, you will need to set a "seed" in your code just after the `library(spsurvey)` statement. To set the seed that matches the one we will use, run
```{r}
set.seed(5)
```

# `sf` Objects

To obtain random samples using the algorithms in `spsurvey`, your data must be an `sf` object. An `sf` object, short for a "simple features" object, is an object with a unique structure used to conveniently store spatial data in R. `sf` objects are constructed using the `sf` package in R (Pebesma, 2018). The `sf` package is loaded and installed alongside the `spsurvey` package, so you do not need to run `install.packages("sf")` or `library(spsurvey)` to access the `sf` package if `spsurvey` is already installed and loaded. 

There are three ways we recommend to easily construct `sf` objects in R. The first is to read a shapefile directly into R using `st_read()` or `read_sf()`. The second is to use the `st_sf()` function or the `st_as_sf()` function to combine an appropriate R object (most commonly a data frame) and an appropriate geometry object into an `sf` object. The third only applies if your data is an `sp` object --  Use the `sframe()` function in `spsurvey` to convert your `sp` object to an `sf` object. For more information regarding `sf` objects, see Pebesma (2018) and Lovelace et al. (2019). 

# Coordinate Reference Systems

Coordinate reference systems (CRS) identify a unique location on Earth's surface. CRSs are either geographic or projected. A geographic CRS uses longitude (east-west direction) and latitude (north-south direction) coordinates to represent location with respect to a specific ellipsoid or spheroid surface. Geographic CRSs are measured in degrees, *not meters*. This has important consequences; for example, a one degree difference in latitude is different at different longitudes. Projected CRSs are measured in standard cartesian coordinates with respect to a flat surface. They have x and y locations, an origin, and a unit of measurement (like meters). 

In R, CRS information has traditionally been stored as EPSG codes or proj4string values. This meant that in order to transform your coordinates from one CRS to another, you needed two EPSG codes or proj4strings: one for each CRS. With recent updates in R spatial that [follow GDAL and PROJ](https://www.r-spatial.org/r/2020/03/17/wkt.html), CRSs in `sf` are now stored in R as lists with two components, `input` and `wkt`, which contain information regarding the EPSG code and proj4string. For more information on CRSs and EPSG codes, see https://r-spatial.github.io/sf/index.html, Pebesma (2018), and Lovelace et al. (2019). To search for various coordinate reference systems and EPSG codes, visit https://spatialreference.org or https://epsg.io/ .

It is important that you use appropriate CRS in your `sf` object. If your CRS is mis-specified, you may get misleading results. `spsurvey` will use the coordinates in the geometry list-column stored in your `sf` object, so it is your responsibility to make sure they have the correct CRS.

# References

Lovelace, R., Nowosad, J., & Muenchow, J. (2019). Geocomputation with R. CRC Press.

Pebesma, E., (2018). Simple Features for R: Standardized Support for Spatial Vector Data. *The R Journal*, 10 (1):439-446. https://doi.org/10.32614/RJ-2018-009