---
title: "Summarizing and Visualizing Sample Frames, Samples, and Design Frames"
author: "Michael Dumelle, Tony Olsen, Tom Kincaid, and Marc Weber"
output: 
  html_document:
    theme: flatly
    number_sections: true
    highlighted: default 
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Summarizing and Visualizing Sample Frames, Samples, and Design Frames}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
```

The `summary()` and `plot()` functions in spsurvey are used to summarize and visualize sample frames, samples, and design frames. Both functions use a formula argument specifying the variables to be summarized or visualized. These functions behave differently for one-sided and two-sided formulas. To learn more about formulas in R, run `?formula`. Only the core functionality of `summary()` and `plot()` will be covered in this vignette, so to learn more about these functions, run `?summary` and `?plot`.

The `plot()` function in spsurvey is built on the `plot()` function in the `sf` package. spsurvey's `plot()` function accommodates all the arguments in sf's `plot()` function and adds a few additional features.  To learn more about the `plot()` function in sf, run `?plot.sf()`. 

Before proceeding with the rest of the vignette, load spsurvey by running
```{r}
library(spsurvey)
```

# Sample Frames

The sample frame contains all units that may be selected in the sample. Summarizing and visualizing the sample frame is often helpful to better understand your data and inform sampling options like stratification and unequal inclusion probabilities. The `NE_Lakes` data in spsurvey is a sample frame that contains lakes from the Northeastern United States. There are three variables in `NE_Lakes` you will summarize and visualize:

1. `AREA_CAT`: lake area categories (small and large)
2. `ELEV`: lake elevation
3. `ELEV_CAT`: lake elevation categories (low and high)

The `summary()` and `plot()` functions in spsurvey require your sample frame is an `sframe` object. To turn `NE_Lakes` into an `sframe` object, use the `sframe()` function:
```{r}
NE_Lakes <- sframe(NE_Lakes)
```

## One-sided formulas

One-sided formulas are used to summarize and visualize a variable's distribution. The variables of interest should be placed on the right-hand side of the formula. To summarize the distribution of `ELEV` (a numeric variable), run
```{r}
summary(NE_Lakes, formula = ~ ELEV)
```
The output contains two columns: `total` and `ELEV`. The `total` column returns the total number of lakes -- it functions as an "intercept" to the formula (and can by removed by supplying `- 1` to the formula). The `ELEV` column returns a numerical summary of lake elevation. To visualize `ELEV`, run
```{r}
plot(NE_Lakes, formula = ~ ELEV)
```

To summarize the distribution of `ELEV_CAT` (a categorical variable), run
```{r}
summary(NE_Lakes, formula = ~ ELEV_CAT)
```
The `ELEV_CAT` column returns the number of lakes in each elevation category. To visualize `ELEV_CAT`, run
```{r}
plot(NE_Lakes, formula = ~ ELEV_CAT, key.width = lcm(3))
```

The `key.width` argument extends the plot's margin to fit the legend text nicely within the plot. The plot's default title is the `formula` argument, though this is changed using the `main` argument to `plot()`.

The formula used by `summary()` and `plot()` is quite flexible. Additional variables are included using `+`:
```{r}
summary(NE_Lakes, formula = ~ ELEV_CAT + AREA_CAT)
```
The `plot()` function returns two plots, one for `ELEV_CAT` and another for `AREA_CAT`:
```{r}
plot(NE_Lakes, formula = ~ ELEV_CAT + AREA_CAT, key.width = lcm(3))
```

Interactions are included using the interaction operator, `:`. The interaction operator returns the interaction between variables and is most useful when used with categorical variables. To summarize the interaction between `ELEV_CAT` and `AREA_CAT`, run
```{r}
summary(NE_Lakes, formula = ~ ELEV_CAT:AREA_CAT)
```
Levels of each variable are separated by `:`. For example, there are 2451 lakes that are in the low elevation category and the small area category. To visualize this interaction, run
```{r}
plot(NE_Lakes, formula = ~ ELEV_CAT:AREA_CAT, key.width = lcm(3))
```

The formula accommodates the `*` operator, which combines the `+` and `:` operators. For example, `ELEV_CAT * AREA_CAT` is shorthand for `ELEV_CAT + AREA_CAT + ELEV_CAT + AREA_CAT`. The formula also accommodates the `.` operator, which is shorthand for all variables separated by `+`.

## Two-sided formulas

Two-sided formulas are used to summarize the distribution of a left-hand side variable for each level of each right-hand side variable. To summarize the distribution of `ELEV_CAT` for each level of `AREA_CAT`, run
```{r}
summary(NE_Lakes, formula = ELEV_CAT ~ AREA_CAT)
```
To visualize the distribution of `ELEV_CAT` for each level of `AREA_CAT`, run
```{r}
plot(NE_Lakes, formula = ELEV_CAT ~ AREA_CAT, key.width = lcm(3))
```

To summarize the distribution of `ELEV` for each level of `AREA_CAT`, run
```{r}
summary(NE_Lakes, formula = ELEV ~ AREA_CAT)
```
To visualize the distribution of `ELEV` for each level of `AREA_CAT`, run
```{r}
plot(NE_Lakes, formula = ELEV ~ AREA_CAT)
```

To only summarize or visualize a particular level of a right-hand side variable, use the `onlyshow` argument:
```{r}
summary(NE_Lakes, formula = ELEV ~ AREA_CAT, onlyshow = "small")
plot(NE_Lakes, formula = ELEV ~ AREA_CAT, onlyshow = "small")
```


## Adjusting graphical parameters

There are three arguments in `plot()` used to adjust graphical parameters:

1. `var_args` adjusts graphical parameters for all levels of a variable 
2. `varlevel_args` adjusts graphical parameters uniquely for each level of a variable
3. `...` adjusts graphical parameters for all levels of all variables
    
The `var_args` and `varlevel_args` arguments take lists whose names match variable names in the formula. For `varlevel_args`, each list element must have an entry called `levels` that matches the variable's levels. The following example combines all three graphical parameter adjustment arguments:
```{r}
list1 <- list(main = "Elevation Categories", pal = rainbow)
list2 <- list(main = "Area Categories")
list3 <- list(levels = c("small", "large"), pch = c(4, 19))
plot(
  NE_Lakes,
  formula = ~ ELEV_CAT + AREA_CAT,
  var_args = list(ELEV_CAT = list1, AREA_CAT = list2),
  varlevel_args = list(AREA_CAT = list3),
  cex = 0.75,
  key.width = lcm(3)
)
```

`var_args` uses `list1` and `list2` to give the `ELEV_CAT` visualization a new title and color palette and `AREA_CAT` a new title; `varlevel_args` uses `list2` to give the `AREA_CAT` visualization different shapes for the small and large levels; and `...` uses `cex = 0.75` to reduce the size of all points and `key.width` to adjust legend spacing. 

If a two-sided formula is used, it is possible to adjust graphical parameters for all levels of a left-hand side variable and all levels of a right-hand side variable when a sublist matching the structure of `varlevel_args` is used as an argument to `var_args`:
```{r}
sublist <- list(AREA_CAT = list3)
plot(
  NE_Lakes,
  formula = AREA_CAT ~ ELEV_CAT,
  var_args = list(ELEV_CAT = sublist),
  key.width = lcm(3)
)
```

# Samples

Samples can also be summarized and visualized using `summary()` and `plot()` in spsurvey. In this section, you will use the `grts()` function to select spatially balanced probability samples. The `grts()` function does incorporate randomness, so to match your results with this output exactly you will need to set a reproducible seed by running
```{r}
set.seed(5)
```
For more information on the `grts()` function, see the "Selecting Spatially Balanced Probaiblity Samples" vignette.

To select an equal probability GRTS sample of size 50 with 10 reverse hierarchically ordered replacement sites, run
```{r}
eqprob <- grts(NE_Lakes, n_base = 50, n_over = 10)
```

The `summary()` function applies a formula to the `grts()` output. The default formula is `~ siteuse`, which returns the number of samples in each `sites` object from the `grts()` output. By default, the formula is applied to all non-`NULL` `sites` objects in the `grts()` output. In this example, those are `sites_base` (for the base sites) and `sites_over` (for the reverse hierarchically ordered replacement sites).
```{r}
summary(eqprob)
```

The `summary()` function is more useful when you include `siteuse` as a left-hand side variable:
```{r}
summary(eqprob, formula = siteuse ~ AREA_CAT)
```

The `plot()` function makes separate visualizations for separate `sites` objects. The default formula is `~ siteuse`, which returns the sampled sites separated by their type:
```{r}
plot(eqprob, key.width = lcm(3))
```

The sample frame may be included as a second argument in the `plot()` function 
```{r}
# pch = 19 plots closed circles
plot(eqprob, NE_Lakes, pch = 19, key.width = lcm(3))
```

The `pch` argument changes the symbol used to represent each lake. The `pch` value of 19 uses closed circles instead of open circles (`pch = 1`; the default).

It is also useful to visualize the sampled sites separately for each level of a categorical variable:
```{r}
plot(eqprob, formula = siteuse ~ AREA_CAT, key.width = lcm(3))
```

# Design Frames

Design frames are data frames or `sf` objects that contain observed data. These are the objects used in spsurvey's analysis functions. spsurvey's `summary()` and `plot()` functions are available for design frames. Design frames are summarized and visualized almost exactly like sample frames, the main difference being that design frames are a bit more flexible than sample frames because design frames are not required to be `sf` objects.

The `NLA_PNW` data in spsurvey is a design frame that contains lakes from California, Oregon, and Washington. There are two variables in `NLA_PNW` you will summarize and visualize:

1. `NITR_COND` : Nitrogen content categories (good, fair, and poor)
2. `BMMI` : Benthic macroinvertebrate multi-metric index

```{r}
NLA_PNW <- dframe(NLA_PNW)
summary(NLA_PNW, formula = ~ NITR_COND + BMMI)
plot(NLA_PNW, formula = ~ NITR_COND + BMMI, key.width = lcm(3))
```

Visualizing design frames that are not `sf` objects requires specification of coordinates and a `crs` to `plot()` using the `xcoord`, `ycoord`, and `crs` arguments.
