---
title: "Analyzing Data"
author: "Michael Dumelle and Tom Kincaid"
output: 
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE
)
```

If you have yet not read the "Start Here" vignette, please do so by running
```{r, eval = FALSE}
vignette("start-here", "spsurvey")
```

# Introduction

spsurvey's analysis functions are used to analyze data in a variety of contexts. We focus mainly on using the `NLA_PNW` analysis data to introduce some of these analysis functions. The `NLA_PNW` data contains response variables measured at several lakes in the Pacific Northwest Region (PNW) of the United States. There are several variables in `NLA_PNW` you will use throughout this vignette:

* `SITE_ID`: a unique site identifier
* `WEIGHT`: the design weights
* `URBAN`: urban land categories (`Urban` and `Non-Urban`)
* `STATE`: state name (`California`, `Oregon`, and `Washington`)
* `BMMI`: benthic macroinvertebrate multi-metric index
* `BMMI_COND`: benthic macroinvertebrate multi-metric index condition categories (`Poor` and `Good`)
* `PHOS_COND`: phosphorous condition categories (`Poor` and `Good`)
* `NITR_COND`: nitrogen condition categories (`Poor` `Fair`, and `Good`)

Before proceeding, we load spsurvey by running
```{r}
library(spsurvey)
```

# Categorical Variable Analysis

Categorical variables are analyzed in spsurvey using the `cat_analysis()` function. The `cat_analysis()` function estimates the proportion of observations and the total units (i.e. extent) that belong to each level of the categorical variable (total units refer to the total number (point resources), total line length (linear network), or total area (areal network)). Several useful pieces of information are returned by `cat_analysis()`, including estimates, standard errors, margins of error, and confidence intervals. The analysis results contain columns with a `.P` and `.U` suffixes. The `.P` suffix corresponds to estimates of proportions for each category, while the `.U` suffix corresponds to estimates of total units (i.e. extent) for each category. 

## Unstratified Analysis

To estimate the proportion of total lakes in each nitrogen condition category and the total number of lakes in each nitrogen condition category, run
```{r}
cat_ests <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT"
)
cat_ests
```
The estimate of the proportion of lakes in `Good` condition is 51.35% with a 95% confidence interval of (36.8%, 65.9%), while the estimate of the total number of lakes in `Good` condition is 5484 lakes with a 95% confidence interval of (3086, 7882). In each case, the estimated standard error and margin of error is given. The confidence level can be changed using the `conf` argument. If more than one categorical variable is of interest,  then `vars` can be a vector of variables and separate analyses are performed for each variable.

Sometimes the goal is to estimate proportions and totals separately for different subsets of the population -- these subsets are called subpopulations. To estimate the proportion of total lakes and in each nitrogen condition category the total number of lakes in each nitrogen condition category separately for `California`, `Oregon`, and `Washington` lakes, run
```{r}
cat_ests_sp <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT",
  subpop = "STATE"
)
cat_ests_sp
```
If more than one type of subpopulation is of interest,  then `subpop` can be a vector of subpopulation variables and separate analyses are performed for each subpopulation. If both `vars` and `subpops` are vectors, separate analyses are performed for each variable and subpopulation combination.

Analysis results for all sites (ignoring subpopulations) can be presented alongside the subpopulation analysis results using the `All_Sites` argument:
```{r}
cat_ests_sp <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT",
  subpop = "STATE",
  All_Sites = TRUE
)
cat_ests_sp
```


## Stratified Analysis

To estimate the proportion of total lakes in each nitrogen condition category and the total number of lakes in each nitrogen condition category stratified by `URBAN` category (whether the lake is classified as `Urban` or `Non-Urban`), run
```{r}
strat_cat_ests <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT",
  stratumID = "URBAN"
)
strat_cat_ests
```

To then compute these estimates separately for `California`, `Oregon`, and `Washington`, run
```{r}
strat_cat_ests_sp <- cat_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "NITR_COND",
  weight = "WEIGHT",
  stratumID = "URBAN",
  subpop = "STATE"
)
strat_cat_ests_sp
```

# Continuous Variable Analysis

Continuous variables are analyzed in spsurvey using the `cont_analysis()` function. The `cont_analysis()` function estimates cumulative distribution functions (CDFs), percentiles, and means of continuous variables. By default, all these quantities are estimated (though this can be changed using the `statsitics` argument to `cont_analysis()`). For the quantities requiring estimation, several useful pieces of information are returned by `cont_analysis()`, including estimates, standard errors, margins of error, and confidence intervals. The `.P` suffix corresponds to estimates of proportions for each variable, while the `.U` suffix corresponds to estimates of total units (i.e. extent) for each variable.

## Unstratified Analysis

To estimate the cumulative distribution function (CDF), certain percentiles, and means of `BMMI`, run
```{r}
cont_ests <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT"
)
```

To view the analysis results for the mean estimates, run
```{r}
cont_ests$Mean
```

Similarly, the CDF and select percentile estimates can be viewed (the output is omitted here) by running
```{r, eval = FALSE}
cont_ests$CDF
cont_ests$Pct
```

To visualize the CDF estimates, run
```{r}
plot(cont_ests$CDF)
```

The solid line indicates the CDF estimates, and the dashed lines indicate lower and upper 95% confidence interval bounds for the CDF estimates. `cdf_plot()` can equivalently be used in place of `plot()` (`cdf_plot()` is currently maintained for backwards compatibility with previous spsurvey versions).

To estimate the CDF, certain percentiles, and means of `BMMI` separately for each state, run
```{r}
cont_ests_sp <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT",
  subpop = "STATE"
)
```

To view the analysis results for the mean estimates, run
```{r}
cont_ests_sp$Mean
```


## Stratified Analysis

To estimate the CDF, certain percentiles, and means of `BMMI` for a design stratified by `URBAN` category, run
```{r}
strat_cont_ests <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT",
  stratumID = "URBAN"
)
```

To view the analysis results for the mean estimates, run
```{r}
strat_cont_ests$Mean
```

To then compute these estimates separately for each state, run
```{r}
strat_cont_ests_sp <- cont_analysis(
  NLA_PNW,
  siteID = "SITE_ID",
  vars = "BMMI",
  weight = "WEIGHT",
  stratumID = "URBAN",
  subpop = "STATE",
)
```

To view the analysis results for the mean estimates, run
```{r}
strat_cont_ests_sp$Mean
```

# Model-Assisted GREG Estimation

When auxiliary information is available for the population, a generalized
regression (GREG) estimator can improve precision of population estimates by incorporating this auxiliarty information into estimation. Supply a one-sided
`formula` describing the auxiliaries and a named `popsize` vector containing
their known population totals. The response is still selected by `vars`.
Only a supplied formula selects GREG; omitting it retains the ordinary
analysis. For more details on GREG estimation, see the Technical Details vignette.

We continue with `NLA_PNW`. For this example, suppose an external inventory
reports 10,700 lakes in the PNW population, including 7,400 nonurban lakes and 3,300 urban lakes.
(these totals are chosen for illustration and do not represent known NLA totals). These totals are close to the design-weighted estimates from `NLA_PNW`
(about 10,680 lakes overall with 7,374 nonurban lakes and 3,306 urban lakes). In an application, use
actual frame or external totals; rounding estimates from the same sample
does not supply additional population information.

```{r greg-totals}
greg_data <- NLA_PNW
greg_data$URBAN <- factor(greg_data$URBAN, levels = c("Non-Urban", "Urban"))
greg_data$STATE <- factor(greg_data$STATE,
  levels = c("California", "Oregon", "Washington"))
greg_formula <- ~ URBAN - 1
colnames(model.matrix(greg_formula, sf::st_drop_geometry(greg_data)))
greg_popsize <- c("URBANNon-Urban" = 7400, "URBANUrban" = 3300)
```

The column totals are the population sizes; inspecting `model.matrix()`
reveals the exact names required by `popsize` (by default, **R** uses treatment contrasts). Categorical auxiliaries work
through their indicator columns, just as continuous auxiliaries work through
their numerical columns. Keep factor levels and contrasts consistent when
constructing totals and analyzing the sample.

## Population Estimates and Variance

Now we construct the GREG estimates:
```{r greg-population}
greg_ests <- cont_analysis(
  greg_data,
  vars = "BMMI",
  siteID = "SITE_ID",
  weight = "WEIGHT",
  formula = greg_formula,
  popsize = greg_popsize,
  statistics = c("Mean", "Total")
)
greg_ests$Mean
```

Similarly, `cat_analysis()` applies
the calibration to category indicators:

```{r greg-categories}
greg_categories <- cat_analysis(
  greg_data,
  vars = "NITR_COND",
  siteID = "SITE_ID",
  weight = "WEIGHT",
  formula = greg_formula,
  popsize = greg_popsize
)
greg_categories
```

## Subpopulations with Unknown or Known Totals

Recall we can specify `subpops` to request estimates by subpopulation. Without `subpopsize`, the
population GREG calibration also applies to each state, though this is generally not as precise as supplying `subpopsize` (`subpopsize` values may not be available). 

```{r greg-unknown-domains}
greg_states <- cont_analysis(
  greg_data,
  vars = "BMMI",
  subpops = "STATE",
  siteID = "SITE_ID",
  weight = "WEIGHT",
  formula = greg_formula,
  popsize = greg_popsize,
  statistics = "Mean"
)
greg_states$Mean
```

If the auxiliary totals are known within each subpopulation level, place them in
`subpopsize` as a nested list: the outer list names the subpopulation variable and the inner
list names its levels. These are totals of the same model-matrix columns.
For illustration, suppose the inventory provides the following state totals:

| State | Nonurban lakes | Urban lakes |
|:--|--:|--:|
| California | 1,500 | 1,000 |
| Oregon | 3,600 | 1,200 |
| Washington | 2,300 | 1,100 |

```{r greg-known-domains}
greg_state_totals <- list(
  California = c("URBANNon-Urban" = 1500, "URBANUrban" = 1000),
  Oregon = c("URBANNon-Urban" = 3600, "URBANUrban" = 1200),
  Washington = c("URBANNon-Urban" = 2300, "URBANUrban" = 1100)
)

greg_states_known <- cont_analysis(
  greg_data,
  vars = "BMMI",
  subpops = "STATE",
  siteID = "SITE_ID",
  weight = "WEIGHT",
  formula = greg_formula,
  popsize = greg_popsize,
  subpopsize = list(STATE = greg_state_totals),
  statistics = "Mean"
)
greg_states_known$Mean
```

Each state uses its own `subpopsize` to improve precision constraints. A state without
an entry in `subpopsize` uses `popsize` if
available.

## Stratification in a Global Formula

Sampling strata and GREG terms have different roles. `stratumID`
identifies independently sampled strata for variance estimation. The
formula determines which totals and GREG components
are shared. Continuing the earlier illustration of a design stratified by
`URBAN`, we can use urban-stratum-specific state effects:

```{r greg-stratified}
greg_counts <- data.frame(
  STATE = factor(rep(levels(greg_data$STATE), each = 2),
    levels = levels(greg_data$STATE)),
  URBAN = factor(rep(levels(greg_data$URBAN), times = 3),
    levels = levels(greg_data$URBAN)),
  lakes = c(1500, 1000, 3600, 1200, 2300, 1100)
)
greg_strat_formula <- ~ URBAN + URBAN:STATE - 1
greg_strat_popsize <- colSums(
  model.matrix(greg_strat_formula, greg_counts) * greg_counts$lakes
)
greg_strat_popsize

greg_strat_ests <- cont_analysis(
  greg_data,
  vars = "BMMI",
  siteID = "SITE_ID",
  weight = "WEIGHT",
  stratumID = "URBAN",
  formula = greg_strat_formula,
  popsize = greg_strat_popsize,
  statistics = "Mean"
)
greg_strat_ests$Mean
```

# Additional Analysis Approaches

Alongside the `cat_analysis()` and `cont_analysis()` functions, spsurvey offers functions for estimating attributable risk, relative risk, risk difference, change, and trend. Attributable risk analysis, relative risk analysis, and risk difference analysis quantify the attributable risk, relative risk, and risk difference, respectively, of environmental resources being in poor condition after exposure to a stressor. Attributable risk analysis is performed using the `attrisk_analysis()` function, relative risk analysis is performed using the `relrisk_analysis()` function, and risk difference analysis is performed using the `diffrisk_analysis()` function. Change and trend analysis capture the behavior of environmental resources between two samples, while trend analysis generalizes this approach to include more than two samples. Often, change and trend analyses are performed to study an environmental resource through time.  Change analysis is performed using the `change_analysis()` function, and trend analysis is performed using the `trend_analysis()` function. The `attrisk_analysis()`, `relrisk_analysis()`, `diffrisk_analysis()`, `change_analysis()` and `trend_analysis()` functions all support shared features like GREG estimation and share very similar syntax with the `cat_analysis()` and `cont_analysis()` functions.

## Attributable Risk, Relative Risk, and Risk Difference Analysis

The attributable risk is defined as 
$$1 -  \frac{P(Response = Poor | Stressor = Good)}{P(Response = Poor)},$$
where $P(\cdot)$ is a probability and $P(\cdot | \cdot)$ is a conditional probability. The attributable risk measures the proportion of the response variable in poor condition that could be eliminated if the stressor was always in good condition. To estimate the attributable risk of benthic macroinvertebrates with a phosphorous condition stressor, run
```{r}
attrisk_ests <- attrisk_analysis(
  NLA_PNW, 
  siteID = "SITE_ID",
  vars_response = "BMMI_COND",
  vars_stressor = "PHOS_COND",
  weight = "WEIGHT"
)
attrisk_ests
```

The relative risk is defined as 
$$\frac{P(Response = Poor | Stressor = Poor)}{P(Response = Poor | Stressor = Good)},$$ which measures the risk of the response variable being in poor condition relative to the stressor's condition. To estimate the relative risk of benthic macroinvertebrates being in poor condition with a phosphorous condition category stressor, run
```{r}
relrisk_ests <- relrisk_analysis(
  NLA_PNW, 
  siteID = "SITE_ID",
  vars_response = "BMMI_COND",
  vars_stressor = "PHOS_COND",
  weight = "WEIGHT"
)
relrisk_ests
```

The risk difference is defined as 
$$P(Response = Poor | Stressor = Poor) - P(Response = Poor | Stressor = Good),$$ which measures the risk of the response variable being in poor condition differenced by the stressor's condition. To estimate the risk difference of benthic macroinvertebrates being in poor condition with a phosphorous condition category stressor, run
```{r}
diffrisk_ests <- diffrisk_analysis(
  NLA_PNW, 
  siteID = "SITE_ID",
  vars_response = "BMMI_COND",
  vars_stressor = "PHOS_COND",
  weight = "WEIGHT"
)
diffrisk_ests
```

By default, the levels of the variables in `vars_response` and `vars_stressor` are assumed to equal `"Poor"` (event occurs) or `"Good"` (event does not occur). If those default levels do not match the levels of the variables in `vars_response` and `vars_stressor`, the levels of `vars_response` and `vars_stressor` must be explicitly stated using the `response_levels` and `stressor_levels` arguments, respectively. Similar to `cat_analysis()` and `cont_analysis()` from the previous sections, subpopulations and stratification are incorporated using `subpops` and `stratumID`, respectively. For more on attributable and relative risk in an environmental resource context, see Van Sickle and Paulsen (2008).

## Change and Trend Analysis

To demonstrate change analysis, we use the `NRSA_EPA7` data. There are several variables in `NRSA_EPA7` you will use next:

* `SITE_ID`: a unique site identifier
* `WEIGHT`: the survey design weights
* `NITR_COND`: nitrogen condition category (`Good`, `Fair`, and `Poor`)
* `BMMI`: benthic macroinvertebrate multi-metric index
* `YEAR`: probability sample (survey) year

To estimate the change between samples (time points) for `BMMI` (a continuous variable) and `NITR_COND` (a categorical variable), run
```{r}
change_ests <- change_analysis(
  NRSA_EPA7,
  siteID = "SITE_ID",
  vars_cont = "BMMI",
  vars_cat = "NITR_COND",
  surveyID = "YEAR",
  weight = "WEIGHT"
)
```
The `surveyID` argument is the variable in the data distinguishing the different samples (`YEAR` in the previous example).

To view the analysis results for `NITR_COND` (the categorical variable), run
```{r}
change_ests$catsum
```
Estimates are provided for the difference between the two samples and for each of the two individual samples (the `_1` and `_2` suffixes).

To view the analysis results for `BMMI` (the continuous variable), run
```{r}
change_ests$contsum_mean
```
Though we don't show an example here, the median can be estimated using the `test` argument in `change_anlaysis()`.

Trend analysis generalizes change analysis to more than two samples (usually time points). Though we omit an example here, the arguments to `trend_analysis()` are very similar to the arguments for `change_analysis()`. One difference is that `trend_analysis()` contains arguments that specify which statistical model to apply to the estimates from each sample.

# Variance Estimation

The default variance estimator in spsurvey is the local neighborhood variance estimator (Stevens and Olsen, 2003). The local neighborhood variance estimator incorporates the spatial locations of design sites into the variance estimation process. Because the local neighborhood variance estimator incorporates spatial locations, the resulting variance estimate tends to be smaller than variance estimates from approaches ignoring spatial locations. The local neighborhood variance estimator requires x-coordinates and y-coordinates of the design sites. When the analysis data is an `sf` object, the coordinates from the data's geometry are used. When the analysis data is just a data frame, x-coordinates and y-coordinates (from an appropriate coordinate reference system) must be provided using the `xcoord` and `ycoord` arguments, respectively, of the analysis function being used.

Several additional variance estimation options are available in spsurvey's analysis functions through the `vartype` and `jointprob` arguments.

# References

Sickle, J. V., & Paulsen, S. G. (2008). Assessing the attributable risks, relative risks, and regional extents of aquatic stressors. *Journal of the North American Benthological Society*, 27(4), 920-931.

Stevens Jr, D. L., & Olsen, A. R. (2003). Variance estimation for spatially balanced samples of environmental resources. *Environmetrics*, 14(6), 593-610.




