---
title: "Technical Details"
author: "Michael Dumelle, Tom Kincaid, Anthony R. Olsen, and Marc Weber"
bibliography: '`r system.file("references.bib", package="spsurvey")`'
toc: true
toc-title: Contents
toc-depth: 3
number-sections: true
documentclass: article
geometry:
  - margin=1in
output:
  html_document:
    number_sections: true
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    toc_depth: 3
    includes:
      in_header: mathjax-equation-numbers.html
---

```{r setup, include = FALSE}
# # jss style
# knitr::opts_chunk$set(prompt=TRUE, echo = TRUE, highlight = FALSE, continue = " + ", comment = "")
# options(replace.assign=TRUE, width=90, prompt="R> ")

# rmd style
knitr::opts_chunk$set(collapse = FALSE, comment = "#>", warning = FALSE, message = FALSE)

# loading
library(spsurvey)
```

# Introduction {#sec:introduction}

This vignette covers technical details regarding the functions that perform
computations in `spsurvey`. `spsurvey` builds its design-based inference from
`survey` [@lumley2004analysis; @lumley2024survey] and provides spatially
referenced tools for sampling and inference. `sf` objects [@pebesma2018simple]
are required for the spatially balanced sampling functions (i.e., `grts()`)
and are supported (but not required) for inference (i.e., the `*_analysis()`
functions). We first provide a notation guide and then describe relevant
details for core functions and then for auxiliary functions.

If you use `spsurvey` in a formal publication or report, please cite it.
Citing `spsurvey` lets its maintainers devote more resources to it in the
future [@dumelle2023spsurvey]. To view the `spsurvey` citation, run

```{r}
citation(package = "spsurvey")
```

# Notation Guide {#sec:notation}

```{=tex}
\begin{equation*}
  \begin{split}
   N, \, n & = \text{Sampling frame (population) size; sample size} \\
   s_i & = \text{The } i\text{th site (sampling unit)} \\
   \mathbf{I}(A) & = \text{Indicator function; one if } A \text{ holds and zero otherwise} \\
   I_i & = \mathbf{I}(s_i \text{ is selected into the sample}) \\
   \pi_i, \, \pi_{ij} & = \text{Inclusion probability for } s_i \text{; joint inclusion probability for } s_i, s_j \\
   \Delta_{ij} = \pi_{ij} - \pi_i \pi_j & = \text{Covariance of the sample membership indicators } I_i \text{ and } I_j \\
   w_i = \pi_i^{-1} & = \text{Design weight for } s_i \\
   y_i & = \text{Response value at } s_i \\
   \tau, \, \hat\tau & = \text{Population total; its estimator} \\
   \bar y, \, \hat{\bar y} & = \text{Population mean; its estimator} \\
   D, \, \delta_i = \mathbf{I}(s_i \in D) & = \text{Subpopulation (domain); its membership indicator} \\
   \mathcal{N}(s_i) & = \text{Local neighborhood of } s_i \\
   w_{ij} & = \text{Local neighborhood weight for the pair } (i, j) \\
   \hat V_{\mathrm{lnb}}(\cdot) & = \text{The local neighborhood variance estimator} \\
   a_i & = \text{Hierarchical (quadrant-recursive) address of } s_i \\
   h = 1, \ldots, H & = \text{Stratum index; number of strata} \\
   w_{1i}, \, w_{2j} & = \text{Stage one weight for cluster } i \text{; stage two weight for unit } j \\
   \hat\tau_{2i} & = \text{Estimated stage two total within cluster } i \\
   v_i & = \text{Voronoi inclusion probability sum for } s_i
  \end{split}
\end{equation*}
```

A hat denotes an estimator (e.g., $\hat\tau$) or a quantity evaluated at
estimated inputs. The index $i$ runs over sites throughout. Notation specific
to a single function or family of functions is introduced where it is first
used.

Unless stated otherwise, everything in this article assumes sampling without
replacement, in which a site can appear in the sample at most once. This is
the setting `grts()` and `irs()` select samples in, and it is what makes the
joint inclusion probabilities $\pi_{ij}$ (and hence $\Delta_{ij}$) the central
quantities of design-based variance estimation. When a variance estimator here
behaves as though sampling were with replacement, we say so explicitly.

# Core Functions

The core functions are the ones a user must invoke to go from a sampling frame
to a design-based inference: Select a sample, adjust its weights, and estimate
parameters and their uncertainty. This part is self-contained, so reading it
start to finish gives a complete account of sampling, weighting, and analysis
in `spsurvey` without reference to anything in the auxiliary functions part.

## Design-Based Inference {#sec:core-foundations}

`spsurvey` is a package for spatial design-based inference. In design-based
inference, the response values $y_1, \ldots, y_N$ for the $N$ units in the
sampling frame are treated as fixed, and randomness comes from which $n$ of
them happen to be selected into the sample. Nothing is assumed about how the
responses were generated. All probability statements refer to the sampling
design, that is, to the distribution of the sample membership indicators
$I_1, \ldots, I_N$.

For classical references on design-based inference, see @cochran1977sampling,
@sarndal1992model, and @lohr2022sampling.

### Populations, Sampling Frames, and Coverage {#sec:core-population}

Inference begins with a target population (i.e., population of interest),
the collection of units about which
we want to make statements (e.g., all lakes in the contiguous United States, or CONUS).
Henceforth we call this simply the population. We represent the population by
a sampling frame, the list of units actually eligible for selection. In
environmental surveys the sampling frame is typically a GIS (Geographic
Information System) layer, such as a hydrography layer derived from NHDPlus
[@moore2019nhdplus].

In a perfect world the population and the sampling frame match exactly. In
practice there are two kinds of coverage error. Overcoverage occurs when some
units in the sampling frame are not in the population. Undercoverage occurs
when some units in the population are not in the sampling frame. Of these,
undercoverage is far more damaging. With overcoverage, all members of the
population have a chance of being selected, but some ineligible units
are selected and should be identified during data collection (i.e., evaluation). The population
is then a subset of the population and its size can be estimated via the
weights from eligible and ineligible members as determined during evaluation. With
undercoverage there is unfortuntaely no such remedy; some members of the population had no
chance at all of being selected, so the inference applies to a narrower
population than the one intended.

### Inclusion Probabilities and Design Weights {#sec:core-incl-prob}

Each unit $s_i$ in the sampling frame has a known inclusion probability
$\pi_i = \Pr(I_i = 1) \in (0, 1]$. Its design weight is $w_i = \pi_i^{-1}$.
Informally, $s_i$ represents $w_i$ units in the sampling frame, including
itself. Two units $s_i$ and $s_j$ have a joint inclusion probability
$\pi_{ij} = \Pr(I_i = 1, I_j = 1)$, with $\pi_{ii} = \pi_i$. The covariance of
the two membership indicators is
$$
\Delta_{ij} = \mathrm{Cov}(I_i, I_j) = \pi_{ij} - \pi_i\pi_j ,
$$
with $\Delta_{ii} = \pi_i(1 - \pi_i)$. Under sampling without replacement the
$\Delta_{ij}$ are generally nonzero for $i \ne j$, because selecting one unit
changes the chance of selecting another. This dependence is precisely what
design-based variance estimation must account for.

`grts()` and `irs()` construct the $\pi_i$ and the corresponding $w_i$ at
selection time. The `*_analysis()` functions use the $w_i$ through the
`weight` argument. The sampling frame does not need to be used again except 
for potential weight adjustments (covered later).

### The Horvitz-Thompson Estimator {#sec:core-ht-est}

For a population total $\tau = \sum_{i=1}^N y_i$, the Horvitz-Thompson (HT)
estimator [@horvitz1952generalization] is
\begin{equation}\label{eq:ht-total}
\hat\tau = \sum_{i=1}^{N} I_i \frac{y_i}{\pi_i} = \sum_{i=1}^n w_i y_i ,
\end{equation}
where the second sum runs over the $n$ sampled units. Because
$E(I_i) = \pi_i$, each term has expectation $y_i$, so $\hat\tau$ is unbiased
for $\tau$ over repeated sampling under the design [@lohr2022sampling].
Intuitively, each sampled unit is inflated by the reciprocal of its own chance
of being selected.

When the population size $N$ is known, a population mean can be estimated as
$\hat{\bar y} = \hat\tau / N$. More often $N$ is not known exactly, and the
mean is estimated as a ratio of two HT totals,
$$
\hat{\bar y} = \frac{\hat\tau}{\hat N}, \qquad
\hat N = \sum_{i=1}^n w_i ,
$$
where $\hat N$ is the HT estimator of the population size. Even when $N$ is known, using $\hat N$
rather than $N$ in the denominator removes the uncertainty in the estimated
population size from the estimate of average behavior, which usually reduces
variance substantially [@hajek1971comment]. If all design weights are equal,
then $\hat N = N$ exactly and the two forms are equivalent.

### Totals, Means, Proportions, and Distribution Functions {#sec:core-estimands}

Every population estimate produced by the `*_analysis()` functions is either an
HT total or a ratio of two HT totals. It is convenient to write both at once.
For a transformation $g(\cdot)$ of the response, define
$$
\hat\tau(g) = \sum_{i=1}^n w_i \, g(y_i), \qquad
\hat R(g) = \frac{\hat\tau(g)}{\hat N} .
$$
Then the estimands used throughout the package are

- **Totals**: $\hat\tau(g)$ with $g(y) = y$ for a continuous response, or
  $g(y) = \mathbf{I}(y = c)$ for the size of category $c$ of a categorical
  response.
- **Means**: $\hat R(g)$ with $g(y) = y$.
- **Proportions**: $\hat R(g)$ with $g(y) = \mathbf{I}(y = c)$.
- **Distribution functions**: $\hat R(g)$ with $g(y) = \mathbf{I}(y \le t)$,
  giving the estimated cumulative distribution function
  $\hat F(t) = \widehat{\Pr}(y \le t)$ evaluated at a grid of $t$ values.

Because every estimand has this form, a single variance estimator applied to
an appropriately transformed and weighted variable suffices for all of them.
The next section describes the four such estimators `spsurvey` offers.

## Variance Estimation {#sec:core-variance}

`spsurvey` offers four variance estimators, selected with the `vartype`
argument: `"local"`, `"SRS"`, `"HT"`, and `"YG"`. They differ in what they
assume about the joint inclusion probabilities and in how much of the design
they attempt to reproduce.

### The Horvitz-Thompson and Yates-Grundy Estimators {#sec:core-ht-yg-var}

Because $\hat\tau$ in \eqref{eq:ht-total} is a linear combination of the
membership indicators, its variance follows directly from the covariances
$\Delta_{ij}$:
\begin{equation}\label{eq:ht-var}
V(\hat\tau) = \sum_{i=1}^N \sum_{j=1}^N \Delta_{ij} \frac{y_i}{\pi_i} \frac{y_j}{\pi_j} .
\end{equation}
This is the classical form given by @sarndal1992model. It says that the
variance of an HT total is a weighted sum of products of weighted responses,
with weights given by the design's own indicator covariances. Only the
sampling design enters randomly; the responses are fixed.

Equation \eqref{eq:ht-var} is a population quantity, since it sums over all
$N$ units. The Horvitz-Thompson variance estimator turns it into a sample
quantity by dividing each term by the probability $\pi_{ij}$ that the pair
$(i, j)$ appears in the sample:
\begin{equation}\label{eq:ht-var-est}
\hat V_{\mathrm{HT}}(\hat\tau) = \sum_{i=1}^n \sum_{j=1}^n \frac{\Delta_{ij}}{\pi_{ij}} \frac{y_i}{\pi_i} \frac{y_j}{\pi_j} .
\end{equation}
This is the same inflation device that produced $\hat\tau$ itself, applied to
pairs of units rather than to single units.

When the design has a fixed sample size, an algebraically equivalent
population expression is available:
$$
V(\hat\tau) = -\frac{1}{2} \sum_{i=1}^N \sum_{j \ne i}^N \Delta_{ij} \left(\frac{y_i}{\pi_i} - \frac{y_j}{\pi_j}\right)^2 ,
$$
and the corresponding Yates-Grundy (also Sen-Yates-Grundy) estimator
[@sen1953estimator; @yates1953selection] is
\begin{equation}\label{eq:yg-var}
\hat V_{\mathrm{YG}}(\hat\tau) = -\frac{1}{2} \sum_{i=1}^n \sum_{j \ne i}^n \frac{\Delta_{ij}}{\pi_{ij}} \left(\frac{y_i}{\pi_i} - \frac{y_j}{\pi_j}\right)^2 .
\end{equation}
The two estimators target the same variance but behave differently in finite
samples. Because \eqref{eq:yg-var} is built from contrasts of weighted
responses rather than from their products, it is invariant to adding a
constant to every $y_i$ and is generally more stable, particularly when the
responses are far from zero. It is also less prone to negative variance
estimates for many designs.

Both estimators require $\Delta_{ij}$, and hence $\pi_{ij}$, for every sampled
pair. `spsurvey` does not compute joint inclusion probabilities for GRTS samples
because there is generally no closed form for them. When
`vartype` is `"HT"` or `"YG"`, the $\pi_{ij}$ are therefore approximated.

### Joint Inclusion Probability Approximations {#sec:core-jointprob}

The `jointprob` argument selects the approximation, which is applied through
`survey`. Each approximation writes $\pi_{ij}$ as a correction to the
independence value $\pi_i\pi_j$ that would hold if the membership indicators
were uncorrelated.

- **`"overton"`** (the default): Overton's approximation
  [@overton1987sampling; @stehman1987estimating] uses
  \begin{equation}\label{eq:overton}
  \pi_{ij} \approx \pi_i \pi_j \, \frac{n - 1}{\, n - (\pi_i + \pi_j)/2 \,}, \qquad i \ne j .
  \end{equation}
- **`"hr"`**: The truncated Hartley-Rao approximation [@hartley1962sampling]
  is closely related but uses the sum of the two inclusion probabilities
  rather than their average, together with a term summarizing how uneven the
  inclusion probabilities are across the sampling frame:
  \begin{equation}\label{eq:hartley-rao}
  \pi_{ij} \approx \pi_i \pi_j \, \frac{n - 1}{\, n - (\pi_i + \pi_j) + n^{-1}\sum_{k=1}^N \pi_k^2 \,}, \qquad i \ne j .
  \end{equation}
- **`"brewer"`**: Brewer's method [@brewer2002combined; @berger2004simple] for
  sampling with probability proportional to size without replacement.

In each case the factor $(n-1)/n$ that dominates the correction reflects the
basic feature of sampling without replacement: Conditioning on one unit being
selected leaves one fewer slot for the rest, so $\pi_{ij}$ falls slightly
below $\pi_i\pi_j$. These are approximations and are not exact for GRTS
samples, whose spatial ordering induces a joint inclusion structure that none
of the three formulas attempts to capture.

### The Simple Random Sampling Estimator {#sec:core-srs-var}

Setting `vartype = "SRS"` names an estimator, not a design. It is the
estimator that would be exactly correct under simple random sampling, and
`spsurvey` will use it whatever the actual design was.

Under simple random sampling without replacement of $n$ units from $N$, every
sample of size $n$ is equally likely, so $\pi_i = n/N$, all design weights
equal $w_i = N/n$, and $\pi_{ij} = n(n-1)/\{N(N-1)\}$ for $i \ne j$.
Substituting these into \eqref{eq:ht-var-est} and simplifying gives the
familiar textbook form
\begin{equation}\label{eq:srs-var}
\hat V_{\mathrm{SRS}}(\hat\tau) = N^2 \left(1 - \frac{n}{N}\right) \frac{s^2}{n}, \qquad s^2 = \frac{1}{n - 1}\sum_{i=1}^n (y_i - \bar y)^2 ,
\end{equation}
where $\bar y$ is the sample mean. All the design information has collapsed
into two numbers, the sample size and the sampling fraction $n/N$.

When the design weights are not all equal, the design is not a simple random
sample and \eqref{eq:srs-var} does not apply. What `spsurvey` reports in that
case is the with-replacement (WR) approximation of @hansen1943theory, which treats
the sample as though the $n$ units had been drawn independently:
\begin{equation}\label{eq:wr-var}
\hat V_{\mathrm{WR}}(\hat\tau) = \frac{n}{n - 1} \sum_{i=1}^n \left(w_i y_i - \frac{\hat\tau}{n}\right)^2 .
\end{equation}
Equation \eqref{eq:wr-var} is exactly what \eqref{eq:ht-var-est} becomes when
every off-diagonal $\Delta_{ij}$ is set to zero. It discards all information
about how the design correlates the selection of different units, and keeps
only the spread of the weighted responses. Equation \eqref{eq:srs-var} is the
special case of \eqref{eq:wr-var} when the weights are equal and a
finite population correction is supplied, which is why the same `vartype`
covers both.

Because \eqref{eq:wr-var} ignores the negative correlation that sampling
without replacement induces among the membership indicators, it is usually
conservative, meaning it tends to overstate the variance of a well-spread
sample. That conservatism is why it is a safe fallback, and also why it
leaves efficiency on the table for a spatially balanced design.

### The Local Neighborhood Variance Estimator {#sec:core-local-var}

The local neighborhood variance estimator [@stevens2003variance], requested
with `vartype = "local"`, is the default for every analysis function and is
one of `spsurvey`'s signature contributions. For a weighted response or residual
vector $z$ with entries $z_i$, it is
\begin{equation}\label{eq:local-var}
\hat V_{\mathrm{lnb}}(z) = \sum_{i=1}^n \sum_{j \in \mathcal{N}(s_i)} w_{ij} \left(z_j - \bar z_{\mathcal{N}(s_i)}\right)^2 , \qquad
\bar z_{\mathcal{N}(s_i)} = \sum_{j \in \mathcal{N}(s_i)} w_{ij} z_j .
\end{equation}
Each site $s_i$ contributes the weighted spread of its own spatial
neighborhood about that neighborhood's own weighted mean. Applied to
$z_i = w_i y_i$, equation \eqref{eq:local-var} estimates $V(\hat\tau)$;
applied to $z_i = w_i (y_i - \hat{\bar y}) / \hat N$, it estimates
$V(\hat{\bar y})$ after the ratio linearization described in
[Subpopulations](#sec:core-subpops).

#### Why Neighborhoods Estimate the Variance {#sec:core-local-why}

The estimator conditions on the property that GRTS delivers, namely spatial
balance. A spatially balanced sample spreads its sites evenly across the
sampling frame, so it behaves like a design that has partitioned the region
into $n$ compact cells of roughly equal inclusion probability and taken one
site from each. That is a highly efficient design, but it creates a familiar
difficulty: With a single site per cell there is no within-cell replication,
so no cell can supply a variance estimate on its own.

The classical remedy in such designs is to collapse neighboring cells and
treat the sites in the merged group as replicates of one another. The local
neighborhood estimator does exactly this, but adaptively and in two
dimensions. It merges each site with its nearest sampled neighbors, treats
that small group as locally homogeneous, and reads the variance contribution
of the site from how much its neighbors disagree with one another. Where the
response varies smoothly in space, neighbors agree closely, the local
contrasts in \eqref{eq:local-var} are small, and the reported variance is
correspondingly smaller than \eqref{eq:wr-var} would give. Where the response
has no spatial structure, neighbors disagree as much as any two sites would,
and the two estimators give similar answers. This is the sense in which the
local neighborhood estimator converts spatial balance into precision, and it
is why it is preferred for GRTS samples. Because the neighborhoods are constructed
from the realized sample rather than derived from the design, the estimator
never requires $\pi_{ij}$.

#### Local Neighborhoods and Their Weights {#sec:core-local-weights}

`localmean_weight()` builds the neighborhoods $\mathcal{N}(s_i)$ and the
weights $w_{ij}$ in three steps.

- **Nearest neighbors**: For each site $s_i$, the neighborhood begins as $s_i$
  itself together with its `nbh - 1` nearest sampled sites by Euclidean
  distance in the sample's coordinates. The default `nbh = 4` gives each site
  three neighbors.
- **Symmetrization**: The neighborhoods are then made mutual. If $s_j$ belongs
  to $\mathcal{N}(s_i)$, then $s_i$ is added to $\mathcal{N}(s_j)$. Nearest
  neighbor relationships are not symmetric in general, and without this step a
  site in a sparse part of the sampling frame could appear in many
  neighborhoods while an isolated site appeared in almost none.
- **Weighting**: Within a neighborhood, an initial weight is assigned by a
  taper on distance rank, so the nearest neighbor receives the largest weight
  and the weight decreases linearly with rank, scaled by $\pi_j^{-1}$ so that
  a neighbor representing more of the sampling frame counts for more. The
  weights in each neighborhood are then normalized to sum to one, making
  $\bar z_{\mathcal{N}(s_i)}$ a genuine weighted average rather than a
  weighted sum.

Normalizing each neighborhood makes the rows of the weight matrix
$\{w_{ij}\}$ sum to one. It does not make the columns sum to one, and the
column sums are what tie the local pieces back to the estimate they are
supposed to describe. To see why, sum the neighborhood means over all sites:
$$
\sum_{i=1}^n \bar z_{\mathcal{N}(s_i)} = \sum_{i=1}^n \sum_{j \in \mathcal{N}(s_i)} w_{ij} z_j = \sum_{j=1}^n z_j \left(\sum_{i=1}^n w_{ij}\right) .
$$
If every column sums to one, the parenthesized factor disappears and the
neighborhood means add up to $\sum_j z_j$, which for $z_j = w_j y_j$ is the HT
total $\hat\tau$ itself. In other words, the neighborhoods decompose the
overall estimate into $n$ local pieces that add back up to it exactly, with no
part of any site's contribution counted twice or dropped. Given that, the
contrasts in \eqref{eq:local-var} are contrasts within a genuine
decomposition, and $\hat V_{\mathrm{lnb}}$ is the sum of a variance
computed separately in each local neighborhood.

To obtain this property, `localmean_weight()` adjusts the row-normalized
weights as little as possible, in a constrained least squares sense, subject
to every column also summing to one. The adjustment is computed with the
Moore-Penrose generalized inverse [@penrose1955generalized]. The resulting
matrix is doubly stochastic, meaning all of its rows and all of its columns
sum to one.

#### The Covariance Generalization {#sec:core-local-cov}

Several `spsurvey` estimands are functions of more than one total, and
combining them correctly requires covariances as well as variances.
`localmean_cov()` supplies these by applying the same neighborhood contrasts
to a pair of columns rather than to one. For columns $z^{(k)}$ and $z^{(l)}$,
\begin{equation}\label{eq:local-cov}
\widehat{\mathrm{Cov}}_{\mathrm{lnb}}\!\left(z^{(k)}, z^{(l)}\right) = \sum_{i=1}^n \sum_{j \in \mathcal{N}(s_i)} w_{ij} \left(z^{(k)}_j - \bar z^{(k)}_{\mathcal{N}(s_i)}\right)\left(z^{(l)}_j - \bar z^{(l)}_{\mathcal{N}(s_i)}\right) .
\end{equation}
Setting $l = k$ recovers \eqref{eq:local-var}, so \eqref{eq:local-cov}
produces a full covariance matrix whose diagonal holds the local neighborhood
variances. This is what lets `spsurvey` combine several totals without
assuming they are independent, as required by the contingency table of
[cont_cdftest()](#sec:core-cont-cdftest), the cell totals of the
[risk functions](#sec:core-risk-analysis), and the paired surveys of
[change_analysis()](#sec:core-change-analysis).

#### Computational Cost {#sec:core-local-cost}

Building the neighborhoods requires a dense $n \times n$ matrix of pairwise
distances, and the doubly stochastic correction requires a generalized inverse
of a matrix of that size. Memory therefore grows with $n^2$ and computing time
with $n^3$. For the sample sizes typical of environmental monitoring this is
negligible, but it becomes noticeable in the low thousands of sites, which
matters most for the `subset_local = FALSE` option described in
[Subpopulations](#sec:core-subpops), where the neighborhood structure is built
from the full sample rather than from the subpopulation alone.

### Choosing Among Variance Estimators {#sec:core-vartype}

The four `vartype` options divide into two groups by what they need from the
design.

- **`"HT"` and `"YG"`**: These reproduce the design's joint inclusion
  structure, using \eqref{eq:ht-var-est} and \eqref{eq:yg-var} with $\pi_{ij}$
  supplied by the chosen `jointprob` approximation. They are the right choice
  when the design genuinely has known or well-approximated joint inclusion
  probabilities. For GRTS samples the approximations do not describe the
  spatial ordering, so these options neither exploit spatial balance nor
  reflect it.
- **`"SRS"` and `"local"`**: These use only the design weights. `"SRS"`
  discards the joint structure entirely, per \eqref{eq:wr-var}, and is
  typically conservative for a spatially balanced sample. `"local"` recovers
  the missing information from the realized spatial arrangement of the sample
  rather than from the design specification, per \eqref{eq:local-var}.

The point estimates are identical under all four options, since they all
follow from \eqref{eq:ht-total}. Only the reported standard errors differ.
For GRTS samples, `"local"` is recommended and is the default.

## Design Features {#sec:core-design-features}

The remaining arguments to the `*_analysis()` functions describe features of
the design that change the estimator, the variance estimator, or both.

### Subpopulations {#sec:core-subpops}

A subpopulation, or domain, is a subset $D$ of the sampling frame that the
analysis is restricted to, such as a single ecoregion or a single lake size
class. Every `*_analysis()` function loops over one or more `subpops`
variables and reports a full set of estimates for each of their levels. When
`subpops` is not supplied, a single level covering every site is created
automatically, and when `subpops` is supplied together with
`All_Sites = TRUE`, that all sites level is reported alongside the
user-supplied levels.

Domain estimation deserves care, because the domain is defined in the sampling
frame but observed only through the sample.

#### The Domain Total and Its Variance {#sec:core-subpops-var}

Let $\delta_i = \mathbf{I}(s_i \in D)$ record domain membership, which is a
fixed property of the sampling frame. The domain total is
$\tau_D = \sum_{i=1}^N \delta_i y_i$, and its HT estimator is obtained by
applying \eqref{eq:ht-total} to the transformed response $\delta_i y_i$:
$$
\hat\tau_D = \sum_{i=1}^n w_i \delta_i y_i .
$$
Its variance follows from \eqref{eq:ht-var} applied to the same transformed
response,
\begin{equation}\label{eq:domain-var}
V(\hat\tau_D) = \sum_{i=1}^N \sum_{j=1}^N \Delta_{ij} \frac{\delta_i y_i}{\pi_i} \frac{\delta_j y_j}{\pi_j} ,
\end{equation}
with the corresponding estimator obtained by dividing each term by $\pi_{ij}$
and summing oversampled pairs, exactly as in \eqref{eq:ht-var-est}.

The important feature of \eqref{eq:domain-var} is the range of the sums. They
run over the whole sampling frame, not over the domain. Terms involving a
non-domain unit contribute zero, but the units contributing zero are
determined by which units were sampled, and that is random. The realized
number of sampled domain members is itself a random variable, and its
variability is part of $V(\hat\tau_D)$. This is the standard design-based
treatment of domains [@cochran1977sampling; @sarndal1992model].

The alternative is to keep only the sampled domain members and treat them as
if they were the entire sample. That conditions on the realized domain sample
size, removes one source of variability, and generally understates the
variance. The gap widens as the domain occupies a smaller share of the
sampling frame, because the realized domain sample size is then relatively
more variable.

For a domain mean, the same logic applies after linearizing the ratio
$\hat{\bar y}_D = \hat\tau_D / \hat N_D$, where
$\hat N_D = \sum_i w_i \delta_i$. To first (Taylor series) order the variance of
$\hat{\bar y}_D$ is the variance of an HT total of the linearized variable
$$
z_i = \frac{\delta_i \left(y_i - \hat{\bar y}_D\right)}{\hat N_D} ,
$$
so the domain indicator again multiplies the quantity whose variance is being
estimated, and again over the full sample rather than the domain alone.

#### What This Means for the Local Neighborhood Weights {#sec:core-subpops-local}

Substituting $z_i$ into \eqref{eq:local-var} makes the consequence concrete.
The neighborhood structure is built from a set of sampled sites, and whether
that set is the domain or the whole sample changes the weight matrix
$\{w_{ij}\}$ and therefore the estimate.

If the neighborhoods are built from the sampled domain members only, the
weight matrix has one row and column per domain member. Every neighborhood
consists entirely of domain members, and the contrasts measure only how
variable $y$ is among the domain's own spatial neighbors.

If instead the neighborhoods are built from every sampled site, the weight
matrix has one row and column per sampled site, and non-domain sites enter
with $z_i = 0$. Such a site is not dropped; rather, it joins its neighbors'
neighborhoods, takes a share of the neighborhood weight, and contributes a
zero to the neighborhood mean, pulling that mean toward zero. A neighborhood
that happens to contain few domain members therefore produces larger
contrasts than one that is entirely inside the domain. That is exactly the
behavior \eqref{eq:domain-var} calls for, since it is how the randomness of
the realized domain sample size enters the variance.

The two constructions also give different neighborhoods geometrically. Domain
members that are far apart in the sampling frame can be adjacent once
non-domain neighbors are removed, so restricting to the domain first can pair
sites that much further away than its non-domain neighbors.

#### The subset_local Argument {#sec:core-subpops-subsetlocal}

The `subset_local` argument selects between the two constructions.

- **`subset_local = FALSE`**: The neighborhood structure is built from every
  sampled site, with domain membership carried by $\delta_i$. This is the
  standard design-based treatment of domains described above, and it is the
  theoretically correct one. It matches how `survey` treats domains for the other `vartype`
  options.
- **`subset_local = TRUE`**: The neighborhood structure is built from the
  sampled domain members only. This conditions on domain membership and
  generally understates the variance, especially for population totals, when the domain does not fill the
  sampling frame. If domains are spatially compact, this approach gives should give
  very similar variance estimates compared to `subset_local = FALSE`, as the 
  neighborhoods will be similar.

`subset_local = TRUE` remains the default for backward compatibility, and it
is considerably cheaper, since the neighborhood machinery then scales with the
domain size rather than with the full sample size (see
[Computational Cost](#sec:core-local-cost)). Users who want design-based
domain variances should set `subset_local = FALSE`. The option is currently not
available for two-stage designs, where the affected functions raise an error
rather than silently returning a conditioned estimate.

### Stratification {#sec:core-stratification}

This is the only section that uses stratification notation. A stratified
design partitions the sampling frame into $h = 1, \ldots, H$ strata and
samples each one independently. Supplying `stratumID` to an `*_analysis()`
function tells it to estimate within each stratum and then combine.

How the per-stratum variances combine depends on the estimand, and it must be
derived from the estimand's own structure rather than assumed.

For a total, the strata totals simply add, $\hat\tau = \sum_h \hat\tau_h$, so
independence across strata gives
\begin{equation}\label{eq:strat-total}
\hat V(\hat\tau) = \sum_{h=1}^H \hat V(\hat\tau_h) ,
\end{equation}
an unweighted sum. For a mean or proportion, the strata estimates combine as a
weighted average, $\hat{\bar y} = \sum_h (\hat N_h/\hat N) \hat{\bar y}_h$
with $\hat N_h = \sum_{i \in h} w_i$, so the same independence gives
\begin{equation}\label{eq:strat-mean}
\hat V(\hat{\bar y}) = \sum_{h=1}^H \left(\frac{\hat N_h}{\hat N}\right)^2 \hat V(\hat{\bar y}_h) ,
\end{equation}
a sum weighted by squared population shares.

Both rules follow from the estimand, not from the variance estimator, so they
hold whichever `vartype` produced each $\hat V_h$.

Stratification also changes what the local neighborhoods are. They are built
and truncated separately within each stratum, so a site's neighbors always
come from its own stratum and never from across a stratum boundary. As the
number of strata grows relative to the sample size, more of each site's true
nearest spatial neighbors fall outside its stratum and become unavailable,
which erodes the efficiency advantage of `vartype = "local"` over
`vartype = "SRS"`.

At selection time, `grts()` and `irs()` stratify through `stratum_var` and
treat each stratum as a completely independent design. The base sample size
`n_base` is then a named vector giving a size per stratum, and `caty_n`,
`mindis`, `n_over`, and `n_near` may each be supplied as per-stratum lists
named to match `n_base`.

### Two-Stage (Cluster) Sampling {#sec:core-twostage}

In a two-stage design the sampling frame is grouped into clusters. A sample of
clusters is drawn at the first stage, and units are then sampled within each
selected cluster at the second stage. Environmental examples include selecting
watersheds and then stream reaches within them, or selecting lakes and then
locations within them.

#### The Standard Two-Stage Estimator {#sec:core-twostage-standard}

Let cluster $i$ have first-stage inclusion probability $\pi_{1i}$ and weight
$w_{1i} = \pi_{1i}^{-1}$, and let unit $j$ within a selected cluster have
second-stage weight $w_{2j}$, the reciprocal of its conditional inclusion
probability given that its cluster was selected. The estimated total within
cluster $i$ is $\hat\tau_{2i} = \sum_{j \in i} w_{2j} y_j$, and the overall HT
estimator is
\begin{equation}\label{eq:twostage-total}
\hat\tau = \sum_{i=1}^{n_1} w_{1i} \hat\tau_{2i} ,
\end{equation}
where $n_1$ is the number of sampled clusters. Conditioning on which clusters
were selected splits the variance into a between-cluster and a within-cluster
part,
$$
V(\hat\tau) = \underbrace{\sum_{i=1}^{N_1} \sum_{j=1}^{N_1} \Delta_{1ij} \frac{\tau_{2i}}{\pi_{1i}} \frac{\tau_{2j}}{\pi_{1j}}}_{\text{between clusters}} \; + \; \underbrace{\sum_{i=1}^{N_1} \frac{V_{2i}}{\pi_{1i}}}_{\text{within clusters}} ,
$$
where $\Delta_{1ij}$ is the first-stage indicator covariance, $\tau_{2i}$ is
the true cluster total, $V_{2i}$ is the second-stage variance within cluster
$i$, and $N_1$ is the number of clusters in the sampling frame. The between
term treats each cluster's total as a single observation drawn under the
first-stage design; the within term accumulates the second-stage variances of
the selected clusters. This is the decomposition `survey` implements for
ordinary two-stage designs, and its sample estimator is
$$
\hat V(\hat\tau) = \hat V_{\text{between}}\!\left(\left\{w_{1i}\hat\tau_{2i}\right\}\right) + \sum_{i=1}^{n_1} w_{1i} \hat V_{2i} .
$$
The first-stage weight appears to the first power in the within term, not
squared. Squaring it would double count the first-stage inflation, since
$\hat V_{2i}$ already estimates a within-cluster variance for one realized
cluster rather than for the sampling frame.

#### The Local Neighborhood Extension {#sec:core-twostage-local}

`spsurvey` requests a two-stage design with `clusterID` for the first-stage
units, together with stage-two `siteID`, `weight`, and coordinates, and
stage-one `weight1` and coordinates. Optional `sweight` and `sweight1` supply
size weights, and `Ncluster` and `stage1size` supply the finite population
correction.

Setting `vartype = "local"` replaces each variance in the decomposition above
with \eqref{eq:local-var}:
\begin{equation}\label{eq:twostage-var}
\hat V(\hat\tau) = \hat V_{\mathrm{lnb}}\!\left(\left\{w_{1i}\hat\tau_{2i}\right\}_{i=1}^{n_1}\right) + \sum_{i=1}^{n_1} w_{1i} \hat V_{\mathrm{lnb}}\!\left(\left\{w_{2j} y_j\right\}_{j \in i}\right) .
\end{equation}
The two terms use different coordinates. The between term is evaluated at the
stage-one coordinates with stage-one inclusion probabilities, treating each
cluster's weighted total as a single observation located at the cluster's
position. The within term is evaluated separately inside each cluster at the
stage-two coordinates. A cluster containing a single stage-two unit
contributes nothing to the within term, since no within-cluster variability
can be estimated from one unit.

### Size Weights {#sec:core-sizeweight}

Sometimes a sampled site stands for an amount of resource rather than for a
count of units. Setting `sizeweight = TRUE` and supplying `sweight` (and
`sweight1` for the first stage of a two-stage design) multiplies the design
weight by a size variable, so that
$$
w_i^{\mathrm{sw}} = w_i \times \texttt{sweight}_i .
$$

An example makes the construction concrete. Suppose lakes are sampled from a
sampling frame of lake polygon centers (i.e., a point), a particular lake is selected with
$\pi_i = 0.02$, so $w_i = 50$, and that lake has area 30 hectares. With no
size weight, $\hat\tau$ built from $w_i$ estimates a count: This lake stands
for 50 lakes. With `sweight` set to the lake area, the site's weight becomes
$w_i^{\mathrm{sw}} = 50 \times 30 = 1500$ hectares, and $\hat\tau$ built from
$w_i^{\mathrm{sw}}$ estimates total lake area rather than total lake count.
The same device converts a stream sample from a count of reaches into a total
length by setting `sweight` to reach length.

The size weight enters as a multiplicative adjustment to the design weight
before any variance computation begins. It is used wherever $w_i$ appears in
the estimators above, including inside \eqref{eq:local-var}, so no separate
size-weight term appears in any variance formula.

### The Finite Population Correction {#sec:core-fpc}

The finite population correction is often presented as a separate factor
applied to an otherwise infinite-population variance. It is more usefully
understood as a summary of information that the general variance formula
already contains.

Equation \eqref{eq:ht-var} makes no distinction between finite and infinite
populations. It accounts for the finiteness of the sampling frame through the
off-diagonal $\Delta_{ij}$, which are negative for sampling without
replacement because selecting one unit reduces the chance of selecting
another. If joint inclusion probabilities were known, \eqref{eq:ht-var-est}
would need no correction of any kind, since all of the relevant information is
already in $\Delta_{ij}/\pi_{ij}$.

The correction appears only when those off-diagonal terms are discarded.
Comparing \eqref{eq:srs-var} with \eqref{eq:wr-var} shows what happens: Under
simple random sampling the entire effect of the off-diagonal $\Delta_{ij}$
collapses into the single scalar $1 - n/N$, so the with-replacement estimator
can be repaired by multiplying it by that factor. This is the familiar finite
population correction, and it can only be factored out because
simple random sampling ensures the correction to be the same
for every pair of units.

The local neighborhood estimator is a variance estimator in the same sense
that \eqref{eq:ht-var-est} is, not a corrected version of \eqref{eq:wr-var}.
It recovers pair information from the realized spatial arrangement rather
than from $\pi_{ij}$, and that information does not collapse into a single
scalar, so there is no finite population correction to factor out of it. It is
also derived for an effectively continuous population, following the extension
of the Horvitz-Thompson framework to continuous universes by
@cordy1993extension.

In practice, `fpc` controls whether population size information is passed
through to `survey`. Supplying it affects the `"SRS"`, `"HT"`, and `"YG"`
variances, and has no effect on $\hat V_{\mathrm{lnb}}$. For a single-stage
design `fpc` gives the resource size; for a two-stage design `Ncluster` and
`stage1size` give the number of clusters in the sampling frame and the number
of stage-two units within each sampled cluster.

### Known Population Sizes {#sec:core-popsize}

Supplying `popsize` incorporates known sampling frame totals into the weights.
The adjustment is chosen by the class of the object supplied.

- **A `data.frame`, `table`, or `xtabs`**: Post-stratification. Weights are
  adjusted within each post-stratum so that the post-stratum's weight total
  matches its known count in the sampling frame.
- **A `list`**: Calibration, with the first element supplying the overall
  population total and subsequent elements supplying margin totals for each
  calibration variable. This is a generalized regression (GREG) style weight
  calibration.

Both adjustments act on the weights, and the adjusted weights are then used
everywhere an unadjusted weight would be, including in the point estimates and
in \eqref{eq:local-var}.

The variance consequences differ by `vartype`. For `"SRS"`, `"HT"`, and
`"YG"`, the calibration is handled by `survey`, which accounts for the
adjustment in its variance calculation. For `"local"`, the local neighborhood
estimator is applied to the adjusted weight times the raw response. This
correctly reflects the effect of the calibration on the weights themselves,
but it is not the residual-based variance a full GREG treatment would use.
A full treatment would regress the response on the calibration variables and
pass the model residuals to the variance estimator, so that the variability
explained by the auxiliary variables is removed before the local contrasts are
formed. Extending the local neighborhood estimator in this way is planned
future work (see [Future Directions](#sec:future-directions)). Until then, a
user relying on `popsize` with a strongly predictive auxiliary variable should
not expect the reported `"local"` standard error to show the full efficiency
gain that calibration makes available.

## Spatially Balanced Sampling {#sec:core-sampling}

### Inclusion Probabilities for Sampling {#sec:core-sampling-ip}

Before any site is selected, `grts()` and `irs()` compute an inclusion
probability for every unit in the sampling frame. These probabilities are what
the selection algorithms in the following sections realize into an actual
sample, and what the reported design weights are built from.

#### Equal Probability {#sec:core-ip-equal}

With `seltype = "equal"`, every unit in the sampling frame receives
$\pi_i = n/N$, where $n$ is the requested sample size and $N$ is the number of
units in the sampling frame.

For a point sampling frame, $N$ is simply the number of points. For a linear
or areal sampling frame, there is no finite $N$ to count: A stream network is a
continuum of locations along its lines, and a set of wetland polygons is a
continuum of locations within its areas. The sampling frame is then infinite,
and the design is described by a density rather than by a per-unit
probability. A sample of $n$ sites from a network of total length $L$ has
inclusion density $n/L$, so that a segment of length $\ell$ has expected count
$n\ell/L$, and similarly with total area in place of total length. The design
weight of a selected site is the reciprocal of that density, and so carries
units of length or area rather than counts. How `spsurvey` realizes such a
design in practice is described in
[Linear and Areal Resources](#sec:core-grts-linear-areal).

#### Unequal Probability {#sec:core-ip-unequal}

With `seltype = "unequal"`, categories come from `caty_var` and expected
per-category sample sizes from `caty_n`. Within category $c$ containing $N_c$
units in the sampling frame, every unit receives $\pi_i = n_c/N_c$. Requesting
$n_c$ greater than $N_c$ implies an inclusion probability greater than one and returns an error.

The values in `caty_n` are expected sample sizes, not guaranteed ones.
Selection still proceeds across the whole sampling frame at once, so the
realized number of sites in each category is random with mean $n_c$. This is
the essential difference from stratification. A stratified design draws each
stratum separately and therefore delivers exactly the requested count in every
stratum, whereas an unequal probability design delivers the requested counts
only on average.

For GRTS samples, giving up exact control is often useful. Because the sites are drawn in one
pass, the local neighborhoods used by `vartype = "local"` are formed from
whichever sampled sites are genuinely closest, rather than being truncated at
category boundaries the way they are truncated at stratum boundaries (see
[Stratification](#sec:core-stratification)). An unequal probability design can
therefore deliver the same emphasis on rare categories while preserving more
of the variance reduction that spatial balance provides. Stratification
remains the right choice when the strata are small exact per-stratum sample sizes are required,
for instance when they are fixed by a reporting obligation.

#### Proportional to Size {#sec:core-ip-proportional}

With `seltype = "proportional"` and a size variable `aux_var`, inclusion
probabilities start out proportional to the auxiliary variable $x_i$ and
normalized to sum to $n$. Because a very large unit can receive a value above
one, which is not a probability, such units are pinned to one and the
remaining probability budget is redistributed. Pinning can push further units
above one, so the step repeats until no unit exceeds one. Units with a zero or
negative auxiliary value receive $\pi_i = 0$ and are excluded from selection,
with a warning.

As an example, suppose five lakes with areas $x = (1, 2, 3, 10, 40)$ hectares
are available and $n = 3$ sites are requested. The areas sum to $56$, so the
initial probabilities $\pi_i = 3x_i/56$ are

$$
(0.054, \; 0.107, \; 0.161, \; 0.536, \; 2.143) .
$$

The fifth lake exceeds one, so it is pinned at $\pi_5 = 1$ and removed from
further rescaling. That leaves a budget of $3 - 1 = 2$ to spread over the
first four lakes, whose areas sum to $16$, giving $\pi_i = 2x_i/16$:

$$
(0.125, \; 0.250, \; 0.375, \; 1.250) .
$$

The fourth lake now exceeds one, so it is pinned at $\pi_4 = 1$ as well. The
remaining budget of $3 - 2 = 1$ is spread over the first three lakes, whose
areas sum to $6$, giving $\pi_i = x_i/6$:

$$
(0.167, \; 0.333, \; 0.500) .
$$

No further unit exceeds one, so the loop stops. The final inclusion
probabilities are $\pi = (1/6, 1/3, 1/2, 1, 1)$, which sum to $3$ as required,
and the corresponding design weights are $w = (6, 3, 2, 1, 1)$. The two
largest lakes are selected in every possible sample
and hence, represent only themselves.

#### Legacy Sites {#sec:core-ip-legacy}

Legacy sites are previously selected sites that must appear in the current
sample. They are supplied through `legacy_var` or `legacy_sites`, with
optional `legacy_stratum_var`, `legacy_caty_var`, and `legacy_aux_var` to
place them within the stratification, category, or proportional-to-size scheme. Legacy sites
are distinct from handpicked sites, which a user may want surveyed but which
were never randomly selected and therefore cannot be used for design-based
inference. @foster2017spatially discuss legacy site designs in the former,
statistically valid sense.

Each legacy site receives $\pi_i = 1$, and the same pin-and-rescale iteration
described above then redistributes the remaining probability across the other
units so that the whole vector still sums to $n$. Two points about this are
easy to miss.

- **The sample size includes the legacy sites**: The value $n$ here is the
  total requested sample size, not the number of new sites to add alongside
  the legacy ones.
- **The rescaling only drives selection**: Design weights are computed from
  the inclusion probabilities each site would have had under the ordinary
  scheme, before legacy sites were inflated to certainty. Legacy sites
  therefore do not carry weight one in the analysis. They carry the weight
  implied by the design that would have been run without them being treated 
  as legacy sites.

### grts() {#sec:core-grts}

`grts()` implements the Generalized Random Tessellation Stratified design of
@stevens2004spatially. The idea is to impose an ordering on the sampling frame
that respects two-dimensional proximity, and then to draw a sample that is
spread evenly along that ordering. A sample that is evenly spread along a
proximity-respecting ordering is spread evenly in space.

#### Hierarchical Addressing {#sec:core-grts-address}

The ordering comes from a quadrant-recursive address. Divide the bounding
region of the sampling frame into four quadrants and label them $0$, $1$, $2$,
$3$. Divide each quadrant into four sub-quadrants and label those the same
way, and continue. After $k$ levels, every location has a $k$-digit base-4
address $a_i$ recording which quadrant it fell in at each level. Sorting by
address puts locations that share coarse quadrants next to one another, and
within those, locations that share finer quadrants next to one another.

Consider four sites on the unit square:

$$
s_1 = (0.10,\, 0.20), \quad s_2 = (0.40,\, 0.80), \quad s_3 = (0.60,\, 0.30), \quad s_4 = (0.90,\, 0.70) .
$$

At each level, record one bit per coordinate, equal to $0$ in the lower half
of the current interval and $1$ in the upper half, and combine the two bits
into a base-4 digit as $2 b_y + b_x$. The first two levels give

| Site | Level 1 $(b_x, b_y)$ | Level 1 digit | Level 2 $(b_x, b_y)$ | Level 2 digit | Address $a_i$ |
|---|---|---|---|---|---|
| $s_1$ | $(0, 0)$ | 0 | $(0, 0)$ | 0 | 00 |
| $s_2$ | $(0, 1)$ | 2 | $(1, 1)$ | 3 | 23 |
| $s_3$ | $(1, 0)$ | 1 | $(0, 1)$ | 2 | 12 |
| $s_4$ | $(1, 1)$ | 3 | $(1, 0)$ | 1 | 31 |

```{r address-diagram, echo = FALSE, fig.width = 7.2, fig.height = 3.9, fig.align = "center", fig.cap = "Quadrant-recursive addressing of four sites on the unit square. At level 1 the square is split into four quadrants, each labeled by combining the two coordinate bits. At level 2 every quadrant is split the same way, so each of the sixteen cells carries a two-digit base-4 address. Sorting by address groups sites first by quadrant and then by sub-quadrant."}
sites <- data.frame(
  x = c(0.10, 0.40, 0.60, 0.90),
  y = c(0.20, 0.80, 0.30, 0.70),
  lab = c("s1", "s2", "s3", "s4")
)
lab_pos <- ifelse(sites$x > 0.5, 2, 4)

new_panel <- function(main) {
  plot(NA,
    xlim = c(-0.08, 1.08), ylim = c(-0.08, 1.08),
    asp = 1, axes = FALSE, ann = FALSE
  )
  title(main = main, cex.main = 1.05, font.main = 1, line = 0.4)
}

draw_quadrants <- function() {
  rect(0, 0, 1, 1, border = "grey25", lwd = 2)
  segments(0.5, 0, 0.5, 1, col = "grey25", lwd = 2)
  segments(0, 0.5, 1, 0.5, col = "grey25", lwd = 2)
}

draw_sites <- function() {
  points(sites$x, sites$y,
    pch = 21, bg = "#c05746", col = "white", cex = 1.6, lwd = 1.2
  )
  text(sites$x, sites$y, labels = sites$lab, pos = lab_pos, offset = 0.45, cex = 0.9)
}

op <- par(mfrow = c(1, 2), mar = c(0.4, 0.4, 1.8, 0.4))

new_panel("Level 1")
draw_quadrants()
for (i in 0:1) {
  for (j in 0:1) {
    text(i * 0.5 + 0.07, j * 0.5 + 0.43,
      labels = 2 * j + i, cex = 2.1, col = "grey78"
    )
  }
}
draw_sites()

new_panel("Level 2")
for (k in c(0.25, 0.75)) {
  segments(k, 0, k, 1, col = "grey82", lwd = 0.7)
  segments(0, k, 1, k, col = "grey82", lwd = 0.7)
}
draw_quadrants()
for (i in 0:3) {
  for (j in 0:3) {
    text(i * 0.25 + 0.125, j * 0.25 + 0.125,
      labels = paste0(2 * (j %/% 2) + (i %/% 2), 2 * (j %% 2) + (i %% 2)),
      cex = 0.95, col = "grey70"
    )
  }
}
draw_sites()

par(op)
```

Reading the addresses off the right-hand panel confirms the table: $s_1$ sits
in cell 00, $s_3$ in cell 12, $s_2$ in cell 23, and $s_4$ in cell 31.

Sorting by address gives the order $s_1, s_3, s_2, s_4$. The first digit does
most of the work, placing the sites in quadrant order, and the second digit
resolves ties within a quadrant. With more levels, the address distinguishes
locations at finer and finer resolution, and the sort order becomes a genuine
one-dimensional summary of two-dimensional position.

Two implementation details are worth knowing.

- **The labels are randomized**: At every level, the four quadrant labels are
  independently randomly permuted. Without this, the ordering would always
  sweep the region in the same direction, so the sample would carry a fixed
  orientation with respect to the sampling frame. Randomizing removes that
  orientation while leaving intact the property that matters, namely that
  sites sharing a coarse quadrant remain adjacent in the ordering.
- **The address is computed from coordinate bits, not by recursion**: The
  original formulation, and earlier versions of `spsurvey`, built the address
  by literally subdividing the sampling frame level by level and re-indexing
  the units in each cell. That is expensive on large sampling frames. The
  current implementation rescales the coordinates to integers and interleaves
  their binary representations, since the $k$th bit of the rescaled $x$ and
  $y$ coordinates is exactly the level-$k$ quadrant information. Interleaving
  the bits of the two coordinates therefore produces the same address in one
  pass over the sampling frame, with no recursion and no intermediate grid.
  The example above shows why: The level-$k$ digit depends only on the $k$th
  binary digit of each coordinate. See @chauvet2021inference for more.

#### Selecting the Sample {#sec:core-grts-selection}

Once the sampling frame is in address order, a sample of size $n$ is drawn
that respects the prescribed inclusion probabilities and spreads itself along
the ordering. `spsurvey` currently does this with ordered pivotal sampling
[@deville1998unequal; @tille2025sampling]. At each step, two units that are
not yet decided have their inclusion probabilities adjusted against one
another, one moving toward zero and the other toward one, in a way that
preserves both units' marginal inclusion probabilities. 
Repeating until every unit is decided yields a sample that matches the
prescribed $\pi_i$ exactly and has fixed size. Applied to units already sorted
by hierarchical address, competition is always between spatial neighbors, so
two nearby units rarely both end up selected. This is the scheme
analyzed by @chauvet2021inference.

@stevens2004spatially instead map the addressed units onto a line of length
$n$ and take a systematic sample from that line with a random start and unit (i.e., length-one)
step. The two approaches share the same addressing, target
the same inclusion probabilities, and produce samples with very similar
spatial balance, but they are not the same algorithm and will not produce the
same sample from the same random seed. Ordered pivotal sampling was adopted
mainly for convenience, since a well-tested implementation was already
available and studied by @chauvet2021inference. A future
version of `spsurvey` may return to the systematic sampling formulation.

#### Reverse Hierarchical Ordering {#sec:core-grts-rho}

Sites come out of selection in address order, so consecutive sites tend to
share coarse quadrants. That is undesirable for reporting and for
replacement, because the first few sites in the list would all come from the
same corner of the sampling frame. Reverse hierarchical ordering fixes this by
re-sorting the selected sites so that any prefix of the list is itself
spatially balanced.

The device is to reverse the digits of each site's address. Consider $16$
selected sites, whose positions in address order are $0$ through $15$ and
whose two-digit base-4 addresses run $00, 01, 02, \ldots, 33$. Reversing the
digits and reading the result back as a base-4 number gives a new sort key:

| Position | Address | Reversed | Key |
|---|---|---|---|
| 0 | 00 | 00 | 0 |
| 1 | 01 | 10 | 4 |
| 2 | 02 | 20 | 8 |
| 3 | 03 | 30 | 12 |
| 4 | 10 | 01 | 1 |
| 5 | 11 | 11 | 5 |
| 6 | 12 | 21 | 9 |
| 7 | 13 | 31 | 13 |
| $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ |
| 15 | 33 | 33 | 15 |

Sorting by the key visits the original positions in the order

$$
0,\; 4,\; 8,\; 12,\; 1,\; 5,\; 9,\; 13,\; 2,\; 6,\; 10,\; 14,\; 3,\; 7,\; 11,\; 15 .
$$

The first hierarchical address digit is the coarse quadrant, so positions $0$ through $3$
lie in the first quadrant, positions $4$ through $7$ in the second, and so on.
Reading the reverse hierarchically ordered sequence, the quadrants cycle as first, second, third,
fourth, first, second, and so on. Any prefix of length four therefore visits
every quadrant exactly once, any prefix of length eight visits every quadrant
twice, and the same holds recursively at finer levels. Reversing the digits
promotes the finest quadrant distinction to the most significant position,
which is exactly what makes short prefixes spread out rather than cluster.

The number of address digits used is the smallest number sufficient to
distinguish the selected sites, so it grows with $\log_4$ of the number of
sites selected.

Because the base sample and any oversample are determined from one reverse
hierarchically ordered sequence rather than drawn separately, requesting a
large `n_over` relative to `n_base` degrades the base sample's own spatial
balance. The base sample is a prefix of a sequence balanced for the combined
size, not a sequence optimized for its own size.

#### Linear and Areal Resources {#sec:core-grts-linear-areal}

When the sampling frame has linear (`LINESTRING`) or areal (`POLYGON`)
geometry, there is no finite list of units to address (see
[Equal Probability](#sec:core-ip-equal)). `spsurvey` handles this by first
approximating the resource with a dense systematic point sample, regular
along a line and hexagonal over an area, at `pt_density` points per requested
sample site (default 10, i.e., |`pt_density`| = 10n). The addressing and selection above then run on those
points. Inclusion probabilities are approximate, assigned as if they were randomly drawn from
the infinite population rather than the dense grid. This two-step application of GRTS for
continuous resources follows @cordy1993extension.

The dense point approximation is used for computational efficiency. Earlier
versions of `spsurvey` mapped the continuous resource onto the
one-dimensional hierarchical ordering directly, which is faithful to the
original description but much slower and considerably more intricate. With a
sufficiently dense point set, the approximation is close enough that the
difference is negligible in practice, and the total point count is capped to
keep memory bounded.

#### Minimum Distance {#sec:core-grts-mindis}

Setting `mindis` requests a minimum geographic separation between selected
sites. An ordinary GRTS sample is drawn, pairwise distances are checked, and
if any pair violates the requirement the sample is redrawn with the sites that
already satisfied it automatically included, repeating up to `maxtry` times
(default 10) and warning if the requirement is still unmet.

This trades spatial balance with respect to the sampling frame for guaranteed
geographic spread. The sites automatically included on a later attempt were not
selected with probability one under the original design, so the realized
sample is not necessarily the GRTS sample that the reported inclusion probabilities
describe. Users should treat `mindis` as a real, if usually small,
perturbation of the design rather than as an inconsequential refinement of it.

#### Replacement Sites {#sec:core-grts-replacement}

Two independent mechanisms supply replacement sites when a selected site
cannot be visited: Oversample sites and nearest neighbor sites.

- **`n_over`** (Oversample sites): Reverse hierarchical ordering draws
  $n_{\mathrm{base}} + n_{\mathrm{over}}$ sites as one spatially balanced
  sequence and takes the trailing $n_{\mathrm{over}}$ as an oversample. The
  oversample is itself spatially balanced and is a probability sample in its
  own right, so it can be analyzed with the same machinery as the base sample.
- **`n_near`** (Nearest Neighbor sites): For each base site independently, the $n_{\mathrm{near}}$
  closest unsampled units in the sampling frame are identified by Euclidean
  distance and labeled `"Near-1st"`, `"Near-2nd"`, and so on, up to ten.

Oversample sites must be used in the order they are listed, which is the
order of their site identifiers. This is a direct consequence of reverse
hierarchical ordering: What that ordering guarantees is that prefixes are
spatially balanced, so taking the first three oversample sites gives a
spatially balanced set of three, while taking three chosen for convenience
gives no such guarantee. Skipping oversample sites, or picking from the
middle of the list, forfeits the property the ordering was constructed to
provide.

The two mechanisms are not interchangeable. Near sites are not a probability
sample of the sampling frame, since they are the geometrically closest
available units to one specific base site rather than randomly selected units.
They are used under the assumption that they inherit the weight of the base
site they replace.

### irs() {#sec:core-irs}

The IRS in `irs()` stands for independent random sampling, where independent
refers to spatial location rather than to the sample membership indicators.
An IRS sample is drawn without regard to where the units sit, so the
selections carry no spatial structure. The membership indicators are not
independent of one another in the probabilistic sense: The design still has
fixed sample size and unequal inclusion probabilities, so $\Delta_{ij}$ is
generally nonzero, exactly as for any other design considered here.

Everything else in the pipeline is shared with `grts()`. Inclusion
probabilities, legacy sites, minimum distance handling, and the split into
base, over, and near sites all behave as described above. The one difference
is that the hierarchical ordering step is replaced by a random shuffle of the
sampling frame, and the pivotal selection then runs on that shuffled order.
When the sampling frame is no larger than the requested sample size, every
unit is taken, which is a census rather than a sample (the same is true for GRTS).

Reverse hierarchical ordering is unnecessary for `irs()`. Any subset of an
already-random ordering is itself a valid random sample, so the oversample
is simply the next block of the shuffled draw.

## Design Weight Adjustments {#sec:core-weights}

`adjwgt()` and `adjwgtNR()` post-process the design weights produced by
sample selection, adjusting them to known totals in the sampling frame or for
nonresponse before they reach analysis. Both are class-based ratio adjustments
in the sense of @sarndal1992model and @lohr2022sampling, and both operate on
the weight vector directly.

### adjwgt() {#sec:core-adjwgt}

`adjwgt()` scales weights so that they sum to a known size in the sampling
frame within each of one or more categories given by `wgtcat`. Within category
$c$,
\begin{equation}\label{eq:adjwgt}
w_i^{\mathrm{adj}} = w_i \cdot \frac{F_c}{\sum_{j \in c} w_j} ,
\end{equation}
where $F_c$ is the known size of category $c$ in the sampling frame, supplied
through `framesize`, and the denominator sums the initial weights over the
sites in category $c$ flagged with `sites = TRUE`.

Sites flagged `sites = FALSE`, such as units later determined not to belong to
the target population, receive an adjusted weight of exactly zero rather than
being removed, so the returned vector has the same length as the input. This
is the statistical adjustment that incorporates overcoverage, discussed in
[Populations, Sampling Frames, and Coverage](#sec:core-population), as units
actually in the population have their weights adjusted at a rate that reflects 
the realized degree of overcoverage in the sample. In this context, the population
can be viewed as a subpopulation (domain) of the sampling frame.

### adjwgtNR() {#sec:core-adjwgtnr}

`adjwgtNR()` adjusts weights for unit nonresponse. It assumes that within
adjustment classes given by `MARClass`, whether a unit responded is unrelated
to its response value, which is the missing at random assumption [@dumelle2025missing]. The
`EvalStatus` variable places each unit into one of three groups: Units that
are in the target population and responded (`TRClass`), units that are in the
target population and did not respond (`TNRClass`), and units that are outside
the target population altogether.

Within adjustment class $c$, each responding unit's weight is scaled up by the
ratio of the total weight known to be eligible to the responding weight:
\begin{equation}\label{eq:adjwgtnr}
w_i^{\mathrm{adj}} = w_i \cdot \frac{\sum_{j \in c} \left(r_j + m_j\right) w_j}{\sum_{j \in c} r_j w_j} ,
\end{equation}
where $r_j$ equals one if unit $j$ responded and is in the target population
and zero otherwise, and $m_j$ equals one if unit $j$ is in the target
population but did not respond and zero otherwise. The numerator therefore
sums the weights of the responders and the eligible nonresponders together,
and the denominator sums the weights of the responders alone. Responders come
to represent the nonresponders assumed missing at random within their class.
Nonresponders, and any unit outside the target population, receive an adjusted
weight of zero.

If an adjustment class contains eligible nonresponders but no responders at
all, the ratio in \eqref{eq:adjwgtnr} is undefined, since its denominator is
zero. That class's adjusted weights are set to `NA` with a warning rather than
being silently dropped or zero-filled, because there is no basis in the data
for redistributing that class' weight and the user must make a subsequent decision
as to how to combine currently existing weight categories into a grouping for
which there is at least one responder.

Neither function propagates its adjustment into any variance estimator. Both
adjusted weight vectors are handed to analysis as fixed design weights,
just as an unadjusted weight would be. When the adjustment factors are
themselves estimated with meaningful uncertainty, treating them as fixed
understates the reported variance somewhat, which is the usual caveat for
ratio and nonresponse weighting adjustments.

## Analysis {#sec:core-analysis}

### What the Analysis Functions Estimate {#sec:core-analysis-overview}

The analysis functions take a data frame of sites, their responses, and their
design weights, and returns estimates with standard errors and corresponding confidence
intervals. They share the same structhre: Each one validates its inputs,
constructs a survey design object carrying the weights (for `vartype` of
`"local"` or `"SRS"`) or the inclusion probabilities and a joint probability
method (for `"HT"` or `"YG"`), applies stratification, clustering, and any
`popsize` adjustment, and then loops over subpopulations and response
variables. Only the last step differs fundamentally between them.

- **`cat_analysis()`**: Proportions and totals for each level of a categorical
  response.
- **`cont_analysis()`**: Means, totals, percentiles, and cumulative
  distribution functions for a continuous response.
- **`cont_cdftest()`**: A test of whether two or more subpopulations have the
  same cumulative distribution function.
- **`relrisk_analysis()`, `attrisk_analysis()`, `diffrisk_analysis()`**:
  Measures of association between a binary response and a binary stressor.
- **`change_analysis()`**: Change in a mean, total, or median between two
  surveys of the same design.
- **`trend_analysis()`**: A trend fitted across three or more surveys.

Every one of these estimates is either $\hat\tau(g)$ or $\hat R(g)$ from
[Totals, Means, Proportions, and Distribution Functions](#sec:core-estimands),
or a smooth function of several of them. The subsections below give the
explicit form in each case. Their standard errors come from the variance
estimator selected by `vartype`, applied to the appropriately transformed and
weighted variable, and combined across strata by
\eqref{eq:strat-total} or \eqref{eq:strat-mean} as the estimand requires.

### cat_analysis() {#sec:core-cat-analysis}

For a categorical response with levels $c = 1, \ldots, C$, `cat_analysis()`
estimates the size and the proportion of each level:
$$
\hat\tau_c = \sum_{i=1}^n w_i \mathbf{I}(y_i = c), \qquad
\hat p_c = \frac{\hat\tau_c}{\hat N} .
$$
The total row that appears alongside each response variable's categories is
$\hat N$ itself, computed by the same machinery with $g(y) \equiv 1$, so the
resource size estimate and the category size estimates are consistent by
construction.

Sizes are totals and so combine across strata by \eqref{eq:strat-total};
proportions are ratios and so combine by \eqref{eq:strat-mean}.

### cont_analysis() {#sec:core-cont-analysis}

`cont_analysis()` estimates one or more of four statistics for a continuous
response, selected with the `statistics` argument.

- **`"Total"`**: $\hat\tau = \sum_i w_i y_i$, the direct application of
  \eqref{eq:ht-total}.
- **`"Mean"`**: $\hat{\bar y} = \hat\tau / \hat N$, the ratio estimator.
- **`"CDF"`**: $\hat F(t) = \sum_i w_i \mathbf{I}(y_i \le t) / \hat N$ on the
  proportion scale and $\sum_i w_i \mathbf{I}(y_i \le t)$ on the total scale,
  evaluated at every distinct observed value of the response. The
  `var_nondetect` argument supports left-censored (nondetect) data by
  restricting which comparisons $\mathbf{I}(y_i \le t)$ can be resolved.
- **`"Pct"`**: Percentiles, at the probabilities given by `pctval`. The
  estimated $q$th percentile is
  $\hat y_q = \inf\{t : \hat F(t) \ge q\}$, obtained by inverting the
  estimated distribution function ($\inf\{\cdot\}$ represents a lower bound).

Percentiles are the one exception to the pattern of this article. Their
confidence intervals come from `survey`'s own quantile machinery rather than
from the variance estimator selected by `vartype`, so `vartype` does not
affect them. Ties are handled by rounding, for backward compatibility with
`spsurvey` versions before 5.0.0.

### cont_cdftest() {#sec:core-cont-cdftest}

`cont_cdftest()` tests whether two or more subpopulations share the same
cumulative distribution function. The response is discretized into $C$ classes
by cutting at a common set of values, giving a contingency table of
subpopulation by class. Its cells are the estimated proportions
$$
\hat p_{kc} = \frac{1}{\hat N} \sum_{i=1}^n w_i \, \mathbf{I}(s_i \in \text{subpopulation } k) \, \mathbf{I}(y_i \in \text{class } c) ,
$$
with row and column margins $\hat p_{k\cdot}$ and $\hat p_{\cdot c}$. Equal
distribution functions correspond to independence of row and column in this
table. The corresponding cell totals $\hat\tau_{kc} = \hat N \hat p_{kc}$ are
used on the total scale.

Under simple random sampling the classical statistic would be Pearson's
$$
X^2 = n \sum_{k}\sum_{c} \frac{\left(\hat p_{kc} - \hat p_{k\cdot}\hat p_{\cdot c}\right)^2}{\hat p_{k\cdot}\hat p_{\cdot c}} .
$$
Under a complex design this is not distributed as $\chi^2$, because the cells
are correlated in ways the design determines. `spsurvey` supplies the full
covariance matrix of the cells from \eqref{eq:local-cov} when
`vartype = "local"`, and from `survey` otherwise, and offers six ways of using
it, selected with the `statistic` argument.

- **`"adjWald"`** (the default): The design-based Wald statistic
  $\hat{\mathbf{d}}^{\top}\hat{\mathbf{V}}^{-1}\hat{\mathbf{d}}$, where
  $\hat{\mathbf{d}}$ collects the differences between observed cell totals and
  those expected under independence and $\hat{\mathbf{V}}$ is their estimated
  covariance matrix, then scaled down when the number of sites is small
  relative to the degrees of freedom. @thomas1987small found this adjustment
  beneficial in small samples.
- **`"Wald"`**: The same statistic without the small sample adjustment
  [@koch1975strategy].
- **`"F"`**: The Rao-Scott second-order correction [@rao1984chisquared], which
  divides $X^2$ by an estimated mean generalized design effect computed from
  $\hat{\mathbf{V}}$ and refers the result to an $F$ distribution with
  Satterthwaite degrees of freedom.
- **`"Chisq"`**: The same design effect correction referred to a $\chi^2$
  distribution.
- **`"lincom"`** and **`"saddlepoint"`**: The Rao-Scott statistic's exact
  asymptotic distribution is a linear combination of $\chi^2$ variables with
  weights given by the eigenvalues of a design effect matrix. These options
  evaluate that distribution by numerical integration and by a saddlepoint
  approximation, respectively, instead of using the Satterthwaite
  approximation.

The two Wald statistics use the total-scale covariance matrix, and the
Rao-Scott family uses the proportion-scale matrix. Both are built by applying
\eqref{eq:local-cov} across the cells of the table jointly, and both combine
across strata by the rule matching their scale.

### Risk Analysis {#sec:core-risk-analysis}

Three functions quantify the association between a binary response, such as
whether a site is in poor condition, and a binary stressor, such as whether a
disturbance is present. The response and stressor variables are given by
`vars_response` and `vars_stressor`, and which level of each counts as the
adverse one is set by `response_levels` and `stressor_levels`.

All three are built from the same $2 \times 2$ table of estimated totals.
Writing $r$ for the response level and $c$ for the stressor level, with level
$1$ adverse and level $2$ not, the cells are
$\hat\tau_{rc} = \sum_i w_i \mathbf{I}(y_i = r)\mathbf{I}(x_i = c)$, and the
table is

|  | Stressor poor | Stressor good | Total |
|---|---|---|---|
| **Response poor** | $\hat\tau_{11}$ | $\hat\tau_{12}$ | $\hat\tau_{1\cdot}$ |
| **Response good** | $\hat\tau_{21}$ | $\hat\tau_{22}$ | $\hat\tau_{2\cdot}$ |
| **Total** | $\hat\tau_{\cdot 1}$ | $\hat\tau_{\cdot 2}$ | $\hat N$ |

The two conditional risks that drive everything below are the estimated
probability of poor response given each stressor level,
$$
\hat p_1 = \frac{\hat\tau_{11}}{\hat\tau_{\cdot 1}}, \qquad
\hat p_2 = \frac{\hat\tau_{12}}{\hat\tau_{\cdot 2}} .
$$
The joint covariance matrix of the four cell totals comes from
\eqref{eq:local-cov} when `vartype = "local"`, giving a $4 \times 4$ matrix
that the delta method then propagates.

- **Relative risk**, from `relrisk_analysis()`: The ratio of the two
  conditional risks,
  $$
  \widehat{RR} = \frac{\hat p_1}{\hat p_2} = \frac{\hat\tau_{11}\hat\tau_{\cdot 2}}{\hat\tau_{\cdot 1}\hat\tau_{12}} .
  $$
  A value of $2$ means poor condition is estimated to be twice as likely where
  the stressor is poor.
- **Risk difference**, from `diffrisk_analysis()`: The difference of the same
  two conditional risks, $\widehat{DR} = \hat p_1 - \hat p_2$.
- **Attributable risk**, from `attrisk_analysis()`: Define
  $$
  \hat\theta = \frac{\hat N \hat\tau_{12}}{\hat\tau_{1\cdot}\hat\tau_{\cdot 2}} = \frac{\widehat{\Pr}(\text{response poor} \mid \text{stressor good})}{\widehat{\Pr}(\text{response poor})}, \qquad \widehat{AR} = 1 - \hat\theta .
  $$
  Attributable risk answers a different question from the other two. It
  estimates the fraction of poor condition in the population that would be
  eliminated if the stressor were in good condition everywhere, so it combines
  the strength of the association with how common the stressor is.
  @vansickle2008assessing is the primary reference for this formulation, while
  @vansickle2006using motivates its use in aquatic condition assessment.

Relative risk and attributable risk are both bounded ratios whose sampling
distributions are skewed, so their standard errors and confidence intervals
are computed on the log scale. The partial derivatives of
$\log \widehat{RR}$, or of $\log\hat\theta$, with respect to the four cell
totals are combined with the $4 \times 4$ covariance matrix by the delta
method, the interval is formed on the log scale, and the endpoints are then
back-transformed. This avoids a normal approximation applied directly to a
ratio, which can behave poorly near zero.

Risk difference needs no such treatment, and needs no covariance term at all.
Every site falls at exactly one stressor level, so $\hat p_1$ and $\hat p_2$
are computed from disjoint sets of sites and are independent. Hence
$$
\hat V\!\left(\widehat{DR}\right) = \hat V(\hat p_1) + \hat V(\hat p_2) .
$$
This is the one place where the disjointness of the two stressor groups buys a
simplification. Relative risk and attributable risk both mix cells from a
single shared table and therefore require the full covariance matrix.

All three functions accept `subset_local`, applied to the neighborhood
structure exactly as described in [Subpopulations](#sec:core-subpops).

### change_analysis() {#sec:core-change-analysis}

`change_analysis()` estimates change between two surveys of the same design,
for `test` equal to `"mean"`, `"total"`, or `"median"`. For means and totals,
the estimate is the difference of the two surveys' own estimates,
$\widehat{\Delta} = \hat\theta_2 - \hat\theta_1$, and its variance is
\begin{equation}\label{eq:change-var}
\hat V(\widehat\Delta) = \hat V(\hat\theta_1) + \hat V(\hat\theta_2) - 2\,\widehat{\mathrm{Cov}}(\hat\theta_1, \hat\theta_2) .
\end{equation}

The covariance term is what makes a revisit design worthwhile, and it can only
be estimated from sites visited in both surveys. Let $R$ denote those
repeat-visit sites. For each $i \in R$, form the pair of contributions that
the site makes to the two surveys' estimates. For totals these are
$$
z_{i1} = w_{i1} y_{i1}, \qquad z_{i2} = w_{i2} y_{i2} ,
$$
and for means they are the linearized versions centered on each survey's own
estimate,
$$
z_{i1} = \frac{w_{i1}\left(y_{i1} - \hat\theta_1\right)}{\hat N_1}, \qquad
z_{i2} = \frac{w_{i2}\left(y_{i2} - \hat\theta_2\right)}{\hat N_2} .
$$
The two columns $z_{\cdot 1}$ and $z_{\cdot 2}$ are then passed to
\eqref{eq:local-cov} as a pair, and the off-diagonal entry of the resulting
$2 \times 2$ matrix is the covariance estimate:
$$
\widehat{\mathrm{Cov}}(\hat\theta_1, \hat\theta_2) = \sum_{i \in R} \sum_{j \in \mathcal{N}(s_i)} w_{ij} \left(z_{j1} - \bar z_{1, \mathcal{N}(s_i)}\right)\left(z_{j2} - \bar z_{2, \mathcal{N}(s_i)}\right) ,
$$
with the neighborhoods built from the repeat-visit sites. A site whose value
is above its neighborhood mean in both surveys contributes positively, which
is the signature of a persistent site effect and is what makes the
change estimate more precise than the difference of two independent surveys.

The `revisitwgt` argument controls how the repeat-visit sites are weighted for
this term. With `revisitwgt = TRUE`, each site's actual design weight is used,
which is valid when that weight is genuinely the same in both surveys. With
`revisitwgt = FALSE`, the default, the repeat-visit sites are given equal
weights for the covariance calculation only, since two surveys' design weights
for the same physical site need not otherwise be comparable. If there are no
repeat-visit sites, the covariance term is treated as zero and a warning is
recorded, which returns \eqref{eq:change-var} to the variance of a difference
of independent estimates (i.e. zeroes-out the covariance term).

The `subset_local` argument affects each survey's own variance estimate. The
repeat-visit covariance term always uses domain members only.

For `test = "median"`, the first survey's estimated median is used to bin both
surveys' responses into two categories, and the change is then analyzed as a
shift in a categorical proportion using the machinery above.

### trend_analysis() {#sec:core-trend-analysis}

`trend_analysis()` fits a trend across three or more surveys, with independent
model choices for categorical responses (`model_cat`, one of `"SLR"`, `"WLR"`,
or `"GLMM"`) and continuous responses (`model_cont`, one of `"SLR"`, `"WLR"`,
or `"LMM"`). The first two operate on the per-year estimates; the last
operates on the site-level data.

- **`"SLR"`**: Simple linear regression of each year's point estimate on year,
  unweighted. Because it is unweighted, the per-year variance estimates are
  computed but never used by the trend model, so `vartype` and `subset_local`
  do not affect the fitted trend, only the per-year estimates reported
  alongside it.
- **`"WLR"`**: The same regression weighted by the inverse of each year's
  estimated variance, so that precisely estimated years count for more. Here
  `vartype` and `subset_local` do affect the trend.
- **`"LMM"` and `"GLMM"`**: Linear and generalized linear mixed models
  [@bates2015fitting] fitted to the site-level data across all years, with a
  default structure of a fixed linear trend in year plus a random
  site-specific intercept and slope and a random year effect, following
  the formulation of @piepho2002simple. Year is centered at the first
  surveyed year, so the fixed intercept is interpreted there rather than
  at an arbitrary calendar origin.
  The right-hand side may be replaced through `cont_rhs` and `cat_rhs`.
  `"GLMM"` requires each categorical response to have exactly two levels.

For `"LMM"`, an optional case-resampling bootstrap refits the model on sites
resampled with replacement and collects the trend coefficient, its standard
error, the variance components, and AIC from each replicate. This gives a
resampling-based alternative to the model's own asymptotic approach (asymptotic inference
can yield p-values that are too small when the sample size is small). The bootstrap is not
available for `"GLMM"`.

# Auxiliary Functions

The auxiliary functions are diagnostic, planning, and display machinery. They
characterize a sample that has already been drawn, plan a monitoring schedule
before any data exist, or support plotting. None are required for the sampling
to inference flow in the core functions.

Several exported functions are deliberately not covered here, because they
perform no statistical estimation. `sp_frame()` and `sp_unframe()` add and
remove the class that the summary and plotting methods recognize,
`sp_rbind()` combines the rows of a drawn sample with the data collected at
those sites, and `warnprnt()`, `errorprnt()`, and `stopprnt()` display the
messages accumulated during a call. `summary()` and `plot()` (equivalently
`sp_summary()` and `sp_plot()`) are exploratory tools: They describe the rows
they are given and do not use design weights, so they summarize the sampling
frame or the sample rather than the target population. Population estimates
and their standard errors come from the core analysis functions.

## sp_balance() {#sec:aux-spbalance}

`sp_balance()` measures how spatially balanced a drawn sample is with respect
to its sampling frame. The idea is that a spatially balanced sample of size
$n$ should divide the sampling frame into $n$ pieces of equal inclusion
probability (i.e., one piece for each sampled site).

The pieces are Voronoi (Dirichlet) polygons. A Voronoi tessellation is built
around the $n$ sample sites, so that each polygon contains the locations
closer to its own site than to any other. Each polygon is intersected with the
sampling frame, and each unit's inclusion probability is apportioned to
whichever polygons it overlaps: The whole probability for a point sampling
frame, and the overlapping share of length or area for a linear or areal one.
Summing the apportioned probability within polygon $i$ gives $v_i$. By
construction $\sum_i v_i = n$, and under perfect spatial balance $E(v_i) = 1$
for every $i$. Writing $p_i = v_i/n$, perfect balance corresponds to every
$p_i$ equalling $1/n$, and the metrics below all measure departure from that.

| Metric | Formula |
|---|---|
| `pielou` (default) | $1 + \sum_i p_i \log p_i \big/ \log(n)$ |
| `simpsons` | $\sum_i p_i^2 - 1/n$ |
| `mse` | $\frac{1}{n}\sum_i (p_i - 1/n)^2$ |
| `rmse` | $\sqrt{\mathrm{mse}}$ |
| `mae` | $\mathrm{mean}_i\big(n \lvert p_i - 1/n\rvert\big)$ |
| `medae` | $\mathrm{median}_i\big(n \lvert p_i - 1/n\rvert\big)$ |
| `chisq` | $n \sum_i (p_i - 1/n)^2$ |

Every metric is zero under perfect spatial balance and increases as balance
worsens. The choice among them is a choice about which departures to
penalize. `mse`, `rmse`, and `chisq` square the departures and so are driven
by the worst polygons; `mae` and `medae` do not, and `medae` in particular is
insensitive to a small number of badly balanced polygons. `mae` and `medae`
are relative errors, expressed as multiples of $1/n$, whereas `mse` and `rmse`
are absolute. `pielou` [@pielou1966measurement] and `simpsons` come from the
ecological diversity literature and treat $\{p_i\}$ as a distribution whose
evenness should be assessed.

The `ip` argument supplies the inclusion probabilities used in the
apportionment, and defaults to an equal probability design within strata when
not given.

On average, GRTS samples should score better on these metrics than IRS samples,
as the spatial balance of a GRTS sample comes
from the quadrant-recursive ordering of
[Hierarchical Addressing](#sec:core-grts-address) spreading selections
across nested quadrants, which is what the Voronoi
apportionment measures. The metrics and the design are assessing the same
geometry. However, it is possible that in a single, realized sample, the
spatial balance for an IRS sample is better than for a GRTS sample.

## Monitoring Designs {#sec:aux-monitoring}

Three functions plan a multi-year monitoring schedule before any data exist,
rather than analyzing data already collected.

### revisit_dsgn() and revisit_bibd() {#sec:aux-revisit}

`revisit_dsgn()` builds a panel-by-period matrix with panels as rows, time
periods as columns, and entries giving the number of units visited. It encodes
one of three revisit structures through a single schedule vector `pnl_dsgn`,
following @mcdonald2003review. Reading the odd positions of `pnl_dsgn` as runs
of consecutive sampled periods and the even positions as runs of unsampled
periods, the final value determines the structure.

- **A trailing zero**: An always-revisit panel, sampled every remaining period
  once the initial schedule is satisfied.
- **A trailing `NA`**: A rotating panel, in which each unit is visited once
  and never again.
- **A trailing value greater than zero**: A serially alternating panel, with a
  fixed on and off cycle repeated for the study's duration.

The three differ in how they trade information about status against
information about change. An always-revisit panel maximizes the covariance
term in \eqref{eq:change-var} and so estimates change most precisely, while a
rotating panel visits the most distinct sites and so estimates status most
precisely (i.e., there are more unique sites sampled across the survey).
Serially alternating panels sit between the two.

`revisit_bibd()` instead constructs a panel assignment from a balanced
incomplete block design [@cook1989computer], which is useful when more panels
are wanted than can each be visited every period. `revisit_rand()` produces a
randomized assignment. All three produce the panel matrix used by
`cov_panel_dsgn()` and `power_dsgn()`.

### cov_panel_dsgn() {#sec:aux-covpanel}

`cov_panel_dsgn()` builds the covariance matrix of a panel design's
period-by-period estimates under a variance component model, following
@urquhart1993comparing, @urquhart1999designs, and @urquhart2012role. Writing
$\Phi$ for the full covariance matrix, the model has four components,
\begin{equation}\label{eq:panel-cov}
\Phi = \Sigma_{\mathrm{unit}} + \Sigma_{\mathrm{period}} + \Sigma_{\mathrm{unit}\times\mathrm{period}} + \Sigma_{\mathrm{index}} ,
\end{equation}
representing persistent differences among units, common shifts affecting all
units in a period, unit-specific responses to a period, and residual
measurement or index variability.

The unit and period components decay geometrically in the period lag, as
$\texttt{unit\_var} \times \texttt{unit\_rho}^{|t - t'|}$ and
$\texttt{period\_var} \times \texttt{period\_rho}^{|t - t'|}$, and every
component accommodates the number of units in a panel and combined across
panels and periods with Kronecker products. The defaults are informative:
`unit_rho = 1` makes the unit component's correlation constant across all
lags, representing a unit effect that persists indefinitely, and
`period_rho = 0` makes periods independent beyond the unit and unit-by-period
terms. The `nrepeats` argument, giving revisits to the same unit within one
panel-period cell, adds a further index variance term accommodating the repeat
count.

The relative sizes of these components determine which panel design is most appropriate.
When the unit component dominates, revisiting the same units removes most of
the noise from a change estimate and always-revisit panels are most appropriate. When the
index component dominates, there is little to be gained from revisiting and
rotating panels are most appropriate.

### power_dsgn() {#sec:aux-power}

`power_dsgn()` computes the power to detect a specified linear trend, across
combinations of panel design, indicator, trend magnitude, and significance
level $\alpha$. `pd_summary()` tabulates the results.

Given the panel covariance matrix $\Phi$ from \eqref{eq:panel-cov} and a
design matrix $\mathbf{X}$ containing an intercept and time, the trend is
estimated by generalized least squares,
\begin{equation}\label{eq:power-dsgn}
\hat{\boldsymbol\beta} = \left(\mathbf{X}^\top \Phi^{-1} \mathbf{X}\right)^{-1}\mathbf{X}^\top \Phi^{-1}\mathbf{y}, \qquad
\mathrm{SE}\!\left(\hat\beta_{\mathrm{trend}}\right) = \sqrt{\left[\left(\mathbf{X}^\top \Phi^{-1}\mathbf{X}\right)^{-1}\right]_{22}} ,
\end{equation}
where the subscript picks out the diagonal entry belonging to the slope. Only
$\Phi$ and $\mathbf{X}$ are needed for the standard error, which is why power
can be computed before any data exist. The hypothesized trend magnitude
$\Delta$ is standardized as $z = \Delta / \mathrm{SE}(\hat\beta_{\mathrm{trend}})$
and substituted into the two-sided power formula
\begin{equation}\label{eq:power-formula}
\mathrm{power} = \Phi_{\mathrm{N}}\!\left(z_{\alpha/2} - z\right) + \left[1 - \Phi_{\mathrm{N}}\!\left(z_{1-\alpha/2} - z\right)\right] ,
\end{equation}
with $\Phi_{\mathrm{N}}(\cdot)$ the standard normal distribution function.

The `trend_type` argument sets what $\Delta$ means. With
`trend_type = "mean"`, it is a percent change in the indicator's mean per
period. With `trend_type = "percent"`, it is a percent change in the
proportion of the distribution above or below a fixed cutoff, which is
converted to an equivalent shift in the mean by a normal-distribution search
before \eqref{eq:power-formula} is applied.

## Density Estimation and Plotting Support {#sec:aux-density}

`ash1_wgt()` computes a weighted averaged shifted histogram [@scott1985averaged],
which is the density display used by `cont_cdfplot()` and `ppd_plot()`. It is
a display aid rather than an inferential estimator, and it is not connected to
any variance estimator in this article.

A histogram's appearance depends on where its bin boundaries happen to fall.
The averaged shifted histogram removes that arbitrariness by averaging many
histograms of the same bin width at staggered origins. It does so without
constructing them separately. The weighted data are first binned finely at
width $\delta$, determined by the requested range and number of bins, and each
fine bin's density height is then a weighted average of the raw counts in the
$m$ bins on either side:
$$
\hat f(x) \propto \sum_{k = -m}^{m} K\!\left(\frac{k}{m}\right) \nu_{b(x) + k}, \qquad
K(t) = \frac{15}{16}\left(1 - t^2\right)^2 ,
$$
where $\nu_b$ is the weighted count in fine bin $b$, $b(x)$ is the bin
containing $x$, and $K$ is the biweight (quartic) kernel. This is 
equivalent to averaging histograms of width $m\delta$ at
staggered origins. Near the edges of the binned range, where fewer than $m$
neighbors exist, the kernel weights are renormalized over the neighbors that
do exist so that the displayed density does not fall off spuriously. Setting
`support = "Ordinal"` replaces the fine binning with one integer-wide bin per
value.

Two further helpers support the CDF plotting machinery. Piecewise linear
interpolation is used to read a CDF's value at an arbitrary percentile, and to
align a right-side axis to a left-side axis at matching CDF values. Neither
performs any statistical estimation.

# Future Directions {#sec:future-directions}

The estimators in this article are design-based throughout but generally ignore 
information in the sampling frame. The natural next step is model-assisted estimation. A generalized regression
(GREG) estimator uses continuous auxiliary variables measured across the whole
sampling frame to build a working model for the response, and then estimates
the total as the model's prediction plus a design-weighted correction built
from the model residuals. Its variance is the design variance of those
residuals rather than of the raw weighted responses, so a good auxiliary
variable reduces the reported variance directly. Extending `spsurvey` in this
direction means passing GREG residuals to the local neighborhood variance
estimator in place of the weighted responses, which would let calibration and
spatial balance contribute to precision at the same time. See
@sarndal1992model for the underlying theory. However, further study and implementation plans are warranted.

Readers who want a fully model-based rather than design-based alternative may
find the `spmodel` package useful [@dumelle2023spmodel]. See @dumelle2022comparison (and relevant references therein)
for an evaluation of design-based versus model-based inference for spatial data.

# References {#sec:references .unnumbered}
